---
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r, warning=FALSE, message=FALSE, include=FALSE}
library(data.table)
library(ggplot2)
library(FactoMineR)
library(circular)
library(CircStats)
library(BIADconnect)
library(webshot)
conn <- init.conn()
knitr::opts_chunk$set(warning = FALSE, message = FALSE,
                      root.dir = "/Users/msb290/Library/CloudStorage/OneDrive-UniversityofCopenhagen/Graves")
```

# Burial orientation
## Data
```{r b-o data}
#### DATA ####
graves <- fread("./Data/Burial.csv", na.strings = "")
graves$DepositionType <- as.factor(graves$DepositionType)
graves$BodyPositioning <- as.factor(graves$BodyPositioning)
graves$BurialSide <- as.factor(graves$BurialSide)
graves[order(IndividualID)]
str(graves)

# Sites table (BIAD)
sites <- setDT(query.database("SELECT * FROM Sites", conn))

# Culture and Period
culture <- fread("./Data/burial rites_culture_period.csv", na.strings = "\\N")
culture <- culture[!duplicated(IndividualID)]

## Add info to graves table
graves <- merge(graves, sites[, .(SiteID, Country)], by = "SiteID", all.x = T)
rm(sites)
graves <- merge(graves, culture, by = "IndividualID", all.x = T)

## Remove outlier
graves <- graves[!is.na(YearBP)]
graves <- graves[YearBP < 15000]
summary(graves$YearBP)
hist(graves$YearBP, breaks = 100)

## Create data table
orientation <- graves[, .(Num, BurialOrientationMin, BurialOrientationMax, BurialSide_back, BurialSide_left, BurialSide_right)]
```

### Calculating the average

Now that we have imported the data, we need to calculate the mean angle for each individual. We have the min and max, but we will work with the mean. There are cases where the min angle is the only angle available. In that case, the min will be used as the mean.

We have two ways to get the mean angles:

1.  Use a custom function that we create

2.  Use the `mean.circular` function of the `circular` package.

We can compare the two to see how they perform.

```{r b-o mean function}
## We can't just average between the min and max angles because for some angles the average will be 0, so we will not get burials facing south.
## We create a function to transform the angles that works around this issue.
#----------------------------------------------------------------------
find.mid.deg <- function(min.deg, max.deg){
  
  difference <- max.deg - min.deg
  i.accute <- difference<180
  i.obtuse <- difference>180
  i.bad <- difference==180
  
  mid <- numeric()
  mid[i.accute] <- max.deg[i.accute] - difference[i.accute]/2
  mid[i.obtuse] <- max.deg[i.obtuse] + (360 - difference[i.obtuse])/2
  mid[i.bad] <- NA
  
  i <- mid>=360
  mid[i] <- mid[i]-360
  
  i <- mid<0
  mid[i] <- mid[i]+360           
  return(mid)}
#----------------------------------------------------------------------
min.deg <- c(5,15,170,270)
max.deg <- c(356,300,185,360)
find.mid.deg(min.deg, max.deg)
#----------------------------------------------------------------------
```

```{r b-o orientation data}
# Remove NAs in both min and max orientation columns
orientation <- orientation[!(is.na(BurialOrientationMin) & is.na(BurialOrientationMax))]

# Create a table for records that have a single angle value (in BurialOrientationMin)
single_angles <- orientation[is.na(BurialOrientationMax)]
single_angles[, BurialOrientationMean := BurialOrientationMin]

# Remove from overall table
orientation <- orientation[!(is.na(BurialOrientationMin) | is.na(BurialOrientationMax))]

# Average across min and max 
## Using Adrian's function
orientation[, BurialOrientationMean := find.mid.deg(BurialOrientationMin, BurialOrientationMax)]

## Using circular package
means <- round(sapply(1:nrow(orientation), function(x) mean.circular(circular(c(orientation$BurialOrientationMin[x], orientation$BurialOrientationMax[x]), type = "angles", units = "degrees", rotation = "clock"))),2)
means <- sapply(1:length(means), function(x) ifelse(means[x] < 0, 360 + means[x], means[x]))

## Check differences
orientation$BurialOrientationMin[698]; orientation$BurialOrientationMax[698]
orientation$BurialOrientationMean[698]; means[698]
```

In this last example we can see that the custom function we have does not recognize 270 as the minimum angle, but it takes 0 as the minimum making the mean **135**. However, because 0 is 360, if we consider 270 as the minimum, the mean should be **315**, as determined by the circular package. 

I believe this is the correct way to move forward, so I will use the mean.circular function to calculate the means.

```{r b-o final data}
orientation[, BurialOrientationMean := means]
# Bind tables together
orientation <- rbind(orientation, single_angles)
orientation <- orientation[order(Num)]

# Calculate cosine and sine from radians
orientation[, BurialOrientationMean_cos := cos(BurialOrientationMean*pi/180)]
orientation[, BurialOrientationMin_cos := cos(BurialOrientationMin*pi/180)]
orientation[, BurialOrientationMax_cos := cos(BurialOrientationMax*pi/180)]
orientation[, BurialOrientationMean_sin := sin(BurialOrientationMean*pi/180)]
orientation[, BurialOrientationMin_sin := sin(BurialOrientationMin*pi/180)]
orientation[, BurialOrientationMax_sin := sin(BurialOrientationMax*pi/180)]

# Add Country, Culture, Period and Ancestry info
orientation$Longitude <- graves[Num %in% orientation$Num]$Longitude
orientation$Latitude <- graves[Num %in% orientation$Num]$Latitude
orientation$Country <- graves[Num %in% orientation$Num]$Country
orientation$Period <- graves[Num %in% orientation$Num]$Period
orientation$Culture <- graves[Num %in% orientation$Num]$Culture
orientation$Sex <- graves[Num %in% orientation$Num]$Sex
orientation$Age <- graves[Num %in% orientation$Num]$gaussianModelMu
orientation$EHG <- graves[Num %in% orientation$Num]$ANCE2_v2
orientation$WHG <- graves[Num %in% orientation$Num]$ANCE4_v2
orientation$LVN <- graves[Num %in% orientation$Num]$ANCE6_v2
orientation$CHG <- graves[Num %in% orientation$Num]$ANCE8_v2
orientation$mobility <- graves[Num %in% orientation$Num]$mobility_MDS_250y_retrospective

# orientation <- fread("./Data/BurialOrientation.csv", na.strings = "")
```

Now we can start visualizing the circular orientation data using histograms, circular histograms and scatterplots that combine sine and cosine values.

### Histograms

```{r b-o hists}
#### HISTOGRAMS ####
ggplot(data = orientation) +
  geom_histogram(aes(x = BurialOrientationMean), bins = 30, fill = "white", col = "black") +
  xlab("Burial orientation (mean)") +
  theme_bw()

ggplot(data = orientation) +
  geom_histogram(aes(x = BurialOrientationMin), bins = 30, fill = "white", col = "black") +
  xlab("Burial orientation (min)") +
  theme_bw()

ggplot(data = orientation) +
  geom_histogram(aes(x = BurialOrientationMax), bins = 30, fill = "white", col = "black") +
  xlab("Burial orientation (max)") +
  theme_bw()

ggplot(data = orientation) +
  geom_histogram(aes(x = BurialOrientationMean_cos), bins = 30, fill = "white", col = "black") +
  xlab("Burial orientation cos(mean)") +
  theme_bw()

ggplot(data = orientation) +
  geom_histogram(aes(x = BurialOrientationMin_cos), bins = 30, fill = "white", col = "black") +
  xlab("Burial orientation cos(min)") +
  theme_bw()

ggplot(data = orientation) +
  geom_histogram(aes(x = BurialOrientationMax_cos), bins = 30, fill = "white", col = "black") +
  xlab("Burial orientation cos(max)") +
  theme_bw()

ggplot(data = orientation) +
  geom_histogram(aes(x = BurialOrientationMean_sin), bins = 30, fill = "white", col = "black") +
  xlab("Burial orientation sin(mean)") +
  theme_bw()

ggplot(data = orientation) +
  geom_histogram(aes(x = BurialOrientationMin_sin), bins = 30, fill = "white", col = "black") +
  xlab("Burial orientation sin(min)") +
  theme_bw()

ggplot(data = orientation) +
  geom_histogram(aes(x = BurialOrientationMax_sin), bins = 30, fill = "white", col = "black") +
  xlab("Burial orientation sin(max)") +
  theme_bw()
```

For the circular histograms, we have to manually create some adjustments in order for ggplot to plot the data correctly. Because there is no way to tell ggplot to create a histogram that can cross zero, we have to manually bin the data in order for data that crosses zero to be in the same bin. This won't shift the data on the axis, or remove data (as ggplot would).

```{r b-o binning function}
#### --------------------------------------------------------------------------------------------------
##### Function to count decimal places #####
decimalplaces <- function(x) {
  if (abs(x - round(x)) > .Machine$double.eps^0.5) {
    nchar(strsplit(sub('0+$', '', as.character(x)), ".", fixed = TRUE)[[1]][[2]])
  } else {
    return(0)
  }
}

##### Function for binning when plotting circular histogram #####
circular.bins <- function(v, binwidth = NULL){
  decimals <- max(sapply(v, decimalplaces))
  bounds.zero <- c(360 - binwidth/2, 0 + binwidth/2)
  # bounds.zero <- round(bounds.zero, decimals)
  center.zero <- seq(round(bounds.zero[1], decimals), 360, by = 1/10^decimals)
  center.zero <- center.zero[-length(center.zero)] # remove 360
  center.zero <- append(center.zero, seq(0, bounds.zero[2], by = 1/10^decimals))
  bins <- seq(bounds.zero[2], bounds.zero[1], binwidth)
  cat <- NA
  for(b in 1:(length(bins) - 1)){
    ix <- which(between(v, bins[b], bins[b+1]))
    cat[ix] <- b
  }
  ix2 <- which(v %in% center.zero)
  cat[ix2] <- 0
  return(cat)
}
#### --------------------------------------------------------------------------------------------------
```

Now that we have the function, we can create circular histograms with binwidth of 22.5, which is 360/16.

## Exploration and visualization
### All individuals
```{r b-o inds, fig.dim=c(10,10)}
##### All individuals ######
dim_angles <- orientation[, .N, by = c("BurialOrientationMean_sin", "BurialOrientationMean_cos")]

ggplot(data = dim_angles) +
  geom_point(aes(BurialOrientationMean_sin, BurialOrientationMean_cos, size = N), 
             shape = 21) +
  scale_size_continuous("Count", breaks = c(1, 10, 50, 100, 150, 200, 250),
                        range = c(3,10)) +
  xlab(expression(sin(theta))) +
  ylab(expression(cos(theta))) +
  coord_equal() +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16))

# Circular histogram
w <- 22.5
orientation[, Bins := circular.bins(BurialOrientationMean, binwidth = w)]
ggplot(data = orientation, aes(x = Bins)) +
  geom_bar(fill = "grey80", color = "black") +
  scale_x_continuous(breaks = seq(0, length(seq(0, 360, w))-1, 1),
                     labels = seq(0, 360, w)) +
  coord_polar(start = -.2) +
  theme_light() +
  theme(axis.title = element_blank(),
        axis.text = element_text(size = 12))
```

We should also check how burial orientation relates to latitude. This is because it is possible that people were buried facing east, where the sun rises. However the angle of where the sun rises changes with latitude, creating a variation in the east-west axis. We can check whether this is true by just plotting burial orientation against latitude.

```{r b-o latitude, fig.dim=c(10,10)}
ggplot(data = orientation) +
  geom_point(aes(y = Latitude, x = BurialOrientationMean)) +
  theme_light() +
  xlab("Angle")
```

### Burial side
```{r b-o burial side, fig.dim=c(10,10)}
##### Burial side ######
## Separate data based on burial side
left <- orientation[, .(Num, BurialOrientationMean, 
                        BurialOrientationMin, 
                        BurialOrientationMax, 
                        BurialOrientationMean_cos, 
                        BurialOrientationMin_cos, 
                        BurialOrientationMax_cos, 
                        BurialOrientationMean_sin, 
                        BurialOrientationMin_sin, 
                        BurialOrientationMax_sin,
                        Bins), by = "BurialSide_left"][BurialSide_left == 1]
left[, BurialSide := "Left"]
right <- orientation[, .(Num, BurialOrientationMean, 
                         BurialOrientationMin, 
                         BurialOrientationMax, 
                         BurialOrientationMean_cos, 
                         BurialOrientationMin_cos, 
                         BurialOrientationMax_cos, 
                         BurialOrientationMean_sin, 
                         BurialOrientationMin_sin, 
                         BurialOrientationMax_sin,
                         Bins), by = "BurialSide_right"][BurialSide_right == 1]
right[, BurialSide := "Right"]
back <- orientation[, .(Num, BurialOrientationMean, 
                        BurialOrientationMin, 
                        BurialOrientationMax, 
                        BurialOrientationMean_cos, 
                        BurialOrientationMin_cos, 
                        BurialOrientationMax_cos, 
                        BurialOrientationMean_sin, 
                        BurialOrientationMin_sin, 
                        BurialOrientationMax_sin,
                        Bins), by = "BurialSide_back"][BurialSide_back == 1]
back[, BurialSide := "Back"]

# Bind together in a ne<- table
sides <- rbind(left[, -1], right[, -1], back[, -1])
# Group by count within same burial side and same angles
# Count number of entries per burial side
dim_side <- sides[, .N, by = c("BurialSide", "BurialOrientationMean_sin", "BurialOrientationMean_cos")]

ggplot(data = dim_side) +
  geom_point(aes(BurialOrientationMean_sin, BurialOrientationMean_cos, fill = factor(BurialSide, levels = c("Left", "Right", "Back")), size = N),
             position = position_jitter(0.05, 0.05),
             shape = 21) +
  scale_fill_manual("Burial side", values = pals::brewer.paired(3)) +
  scale_size_continuous("Count", breaks = c(1, 10, 50, 100, 150),
                        range = c(3,10)) +
  guides(fill = guide_legend(override.aes = list(size=5))) +
  xlab(expression(sin(theta))) +
  ylab(expression(cos(theta))) +
  coord_equal() +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16))

# Circular histogram
cols <- pals::brewer.paired(3)
ggplot(data = sides, aes(x = Bins,
                         fill = factor(BurialSide, levels = c("Left", "Right", "Back"))),
       alpha = 0.5) +
  geom_bar(color = "black", position = "dodge") +
  scale_x_continuous(breaks = seq(0, length(seq(0, 360, w))-1, 1),
                     labels = seq(0, 360, w)) +
  scale_y_continuous() +
  scale_fill_manual("Burial side", values = pals::brewer.paired(3)) +
  coord_polar(start = -.2) +
  ylab("Count") +
  theme_light() +
  theme(axis.text = element_text(size = 12))

# Single plots
ggplot(data = sides, aes(x = Bins, 
                         fill = factor(BurialSide, levels = c("Left", "Right", "Back")))) +
  geom_bar(color = 'black') +
  scale_x_continuous(breaks = seq(0, length(seq(0, 360, w))-1, 2),
                     labels = c("N", "NE", "E", "SE", "S", "SW", "W", "NW", "N")) +
  scale_y_continuous() +
  scale_fill_manual("Burial side", values = pals::brewer.paired(3)) +
  facet_wrap(~factor(BurialSide, levels = c("Left", "Right", "Back")),
             scales = "free_y") +
  ylab("Number of individuals") +
  coord_polar(start = -.2) +
  theme_light() +
  theme(axis.text = element_text(size = 12),
        axis.title.x = element_blank(),
        legend.position = "none")
```

### Country
```{r b-o country, fig.dim=c(10,10)}
##### Country #####
dim_country <- orientation[, .N, by = c("Country", "BurialOrientationMean_sin", "BurialOrientationMean_cos")]
dim_country$Country <- factor(dim_country$Country, 
                              levels = c("Netherlands",
                                         "France",
                                         "Germany",
                                         "Denmark",
                                         "Sweden",
                                         "Latvia",
                                         "Poland",
                                         "Czechia / Czech Republic",
                                         "Slovakia",
                                         "Austria",
                                         "Hungary",
                                         "Croatia",
                                         "Romania",
                                         "Serbia",
                                         "Bulgaria",
                                         "Ukraine",
                                         "Russia",
                                         "Armenia",
                                         "Kazakhstan"))

# Check number of individuals by country
orientation[, .N, by = "Country"][order(N)]

# Plot
ggplot(data = dim_country[!(Country %in% c("Armenia", 
                                           "Kazakhstan"))]) +
  geom_point(aes(BurialOrientationMean_sin, 
                 BurialOrientationMean_cos, 
                 fill = Country, size = N),
             position = position_jitter(0.05, 0.05),
             shape = 21) +
  scale_fill_manual("Country", values = pals::parula(19)) +
  scale_size_continuous("Count", breaks = c(1, 10, 25, 50, 75, 100),
                        range = c(3,10)) +
  guides(fill = guide_legend(override.aes = list(size=5))) +
  xlab(expression(sin(theta))) +
  ylab(expression(cos(theta))) +
  coord_equal() +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16))

# Circular histogram
ggplot(data = orientation[!(Country %in% c("Armenia", "Kazakhstan"))],
       aes(x = Bins, fill = Country),
       color = 'black') +
  geom_bar(color = 'black', position = "dodge") +
  scale_x_continuous(breaks = seq(0, length(seq(0, 360, w))-1, 1),
                     labels = seq(0, 360, w)) +
  scale_y_continuous() +
  scale_fill_manual("Country", values = pals::tol.rainbow(19)) +
  coord_polar(start = -.2) +
  theme_light() +
  theme(axis.text.y = element_blank(),
        axis.title = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_text(size = 12),
        axis.title.x = element_blank())


# Single plots
orientation_country <- orientation[!(Country %in% c("Armenia", "Kazakhstan"))]
orientation_country$Country <- factor(orientation_country$Country,
                                      levels = c("Netherlands",
                                                        "France",
                                                        "Germany",
                                                        "Denmark",
                                                        "Norway",
                                                        "Sweden",
                                                        "Latvia",
                                                        "Poland",
                                                        "Czechia / Czech Republic",
                                                        "Slovakia",
                                                        "Austria",
                                                        "Hungary",
                                                        "Croatia",
                                                        "Romania",
                                                        "Serbia",
                                                        "Bulgaria",
                                                        "North Macedonia",
                                                        "Belarus",
                                                        "Ukraine",
                                                        "Russia",
                                                        "Georgia",
                                                        "Armenia",
                                                        "Kazakhstan"))

ggplot(data = orientation_country, 
       aes(x = Bins, fill = Country)) +
  geom_bar(color = 'black') +
  scale_x_continuous(breaks = seq(0, length(seq(0, 360, w))-1, 2),
                     labels = c("N", "NE", "E", "SE", "S", "SW", "W", "NW", "N")) +
                     # labels = seq(0, 360, w)) +
  scale_y_continuous() +
  # scale_fill_manual("Country", values = pals::tol.rainbow(19)) +
  scale_fill_manual("Country", 
                    values = paletteer::paletteer_c("grDevices::RdYlBu",23)
                    [c(1,2,3,4,6,7,8,9,10,11,12,13,14,15,16,18,19)]) +
  facet_wrap(~Country, scales = "free_y") +
  ylab("Number of individuals") +
  coord_polar(start = -.2) +
  theme_light() +
  theme(axis.text = element_text(size = 12),
        axis.title.x = element_blank(),
        legend.position = "none")
```

### Period
```{r b-o period, fig.dim=c(10,10)}
##### Period ######
dim_period <- orientation[, .N, by = c("Period", "BurialOrientationMean_sin", "BurialOrientationMean_cos")]
dim_period$Period <- factor(dim_period$Period, 
                            levels = c("Mesolithic",
                                       "Mesolithic/Neolithic",
                                       "Neolithic",
                                       "Eneolithic",
                                       "Eneolithic/Bronze Age",
                                       "Bronze Age",
                                       "Bronze Age/Iron Age",
                                       "Iron Age"))

ggplot(data = dim_period) +
  geom_point(aes(BurialOrientationMean_sin, 
                 BurialOrientationMean_cos, 
                 fill = Period, size = N),
             position = position_jitter(0.05, 0.05),
             shape = 21) +
  scale_fill_manual("Period", values = pals::parula(8)) +
  scale_size_continuous("Count", breaks = c(1, 10, 50, 100, 200),
                        range = c(3,10)) +
  guides(fill = guide_legend(override.aes = list(size=5))) +
  xlab(expression(sin(theta))) +
  ylab(expression(cos(theta))) +
  coord_equal() +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16))

# Check number of individuals by period
orientation[, .N, by = "Period"][order(N)]

# Circular histogram
ggplot(data = orientation[Period != "Iron Age"],
       aes(x = Bins, fill = factor(Period,
                                   levels = c("Mesolithic",
                                              "Mesolithic/Neolithic",
                                              "Neolithic",
                                              "Eneolithic",
                                              "Eneolithic/Bronze Age",
                                              "Bronze Age",
                                              "Bronze Age/Iron Age",
                                              "Iron Age"))),
       color = 'black') +
  geom_bar(color = 'black', position = "dodge") +
  scale_x_continuous(breaks = seq(0, length(seq(0, 360, w))-1, 1),
                     labels = seq(0, 360, w)) +
  scale_y_continuous() +
  scale_fill_manual("Period", values = pals::parula(8)) +
  coord_polar(start = -.2) +
  theme_light() +
  theme(axis.text.y = element_blank(),
        axis.title = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_text(size = 12))

# Single plots
ggplot(data = orientation[Period != "Iron Age"], 
       aes(x = Bins, fill = factor(Period,
                                   levels = c("Mesolithic",
                                              "Mesolithic/Neolithic",
                                              "Neolithic",
                                              "Eneolithic",
                                              "Eneolithic/Bronze Age",
                                              "Bronze Age",
                                              "Bronze Age/Iron Age",
                                              "Iron Age")))) +
  geom_bar(color = 'black') +
  scale_x_continuous(breaks = seq(0, length(seq(0, 360, w))-1, 2),
                     labels = c("N", "NE", "E", "SE", "S", "SW", "W", "NW", "N")) +
  scale_y_continuous() +
  # scale_fill_manual("Period", values = pals::parula(8)) +
  scale_fill_manual("Period", values = paletteer::paletteer_d("MetBrewer::Hiroshige",
                                                              10)[c(1,2,3,5,6,7,8)]) +
  facet_wrap(~factor(Period,
                     levels = c("Mesolithic",
                                "Mesolithic/Neolithic",
                                "Neolithic",
                                "Eneolithic",
                                "Eneolithic/Bronze Age",
                                "Bronze Age",
                                "Bronze Age/Iron Age",
                                "Iron Age")), scales = "free_y") +
  ylab("Number of individuals") +
  coord_polar(start = -.2) +
  theme_light() +
  theme(axis.text = element_text(size = 12),
        axis.title.x = element_blank(),
        legend.position = "none")

```

### Culture
```{r b-o culture, fig.dim=c(12,10)}
##### Culture #####
dim_culture <- orientation[!is.na(Culture)][, .N, by = c("Culture", "BurialOrientationMean_sin", "BurialOrientationMean_cos")]
# Plot only the 10 most frequent cultures
culture_keep <- orientation[!is.na(Culture)][, .N, by = "Culture"][order(N, decreasing = T)][, Culture][1:10]
orientation[Culture %in% culture_keep][, .N, by = "Culture"][order(N, decreasing = T)]

dim_culture_keep <- dim_culture[Culture %in% culture_keep]
dim_culture_keep$Culture <- factor(dim_culture_keep$Culture, 
                               labels = c("Linearbandkeramik",
                                          "Lažňany",
                                          "Sopot",
                                          "Bell Beaker",
                                          "Únětice",
                                          "Corded Ware",
                                          "Baden",
                                          "Dnieper-Donets",
                                          "Trichterbecher",
                                          "Baltic Meso-Neolithic"),
                               levels = unique(dim_culture_keep$Culture))

cols <- rev(pals::ocean.thermal(12))[2:11]

ggplot(data = dim_culture_keep) +
  geom_point(aes(BurialOrientationMean_sin, 
                 BurialOrientationMean_cos, 
                 fill = Culture, size = N),
             position = position_jitter(0.05, 0.05),
             shape = 21) +
  scale_fill_manual("Culture", values = cols) +
  scale_size_continuous("Count", breaks = c(1, 10, 25, 50, 75), 
                        range = c(3,10)) +
  guides(fill = guide_legend(override.aes = list(size=5))) +
  xlab(expression(sin(theta))) +
  ylab(expression(cos(theta))) +
  coord_equal() +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16))

# Circular histogram
orientation_culture <- orientation[Culture %in% culture_keep]
orientation_culture$Culture <- factor(orientation_culture$Culture,
                                      labels = c("Linearbandkeramik",
                                          "Lažňany",
                                          "Sopot",
                                          "Bell Beaker",
                                          "Únětice",
                                          "Corded Ware",
                                          "Baden",
                                          "Dnieper-Donets",
                                          "Trichterbecher",
                                          "Baltic Meso-Neolithic"),
                               levels = unique(orientation_culture$Culture))
ggplot(data = orientation_culture,
       aes(x = Bins, fill = Culture),
       color = 'black') +
  geom_bar(color = 'black', position = "dodge") +
  scale_x_continuous(breaks = seq(0, length(seq(0, 360, w))-1, 2),
                     labels = c("N", "NE", "E", "SE", "S", "SW", "W", "NW", "N")) +
  scale_y_continuous() +
  scale_fill_manual("Culture", values = rev(pals::ocean.thermal(12))[2:11]) +
  coord_polar(start = -.2) +
  theme_light() +
  theme(axis.text.y = element_blank(),
        axis.title = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_text(size = 12))

# Single plots
ggplot(data = orientation_culture,
       aes(x = Bins, fill = Culture),
       color = 'black') +
  geom_bar(color = 'black') +
  scale_x_continuous(breaks = seq(0, length(seq(0, 360, w))-1, 2),
                     labels = c("N", "NE", "E", "SE", "S", "SW", "W", "NW", "N")) +
  scale_y_continuous() +
  scale_fill_manual("Culture", values = rev(pals::ocean.thermal(12))[2:11]) +
  ylab("Number of individuals") +
  coord_polar(start = -.2) +
  facet_wrap(~Culture, scales = "free_y") +
  theme_light() +
  theme(axis.title.x = element_blank(),
        legend.position = "none")
```

### Sex

In this case, for the circular histograms, I had to transform the sex data into factor (even though this is continuous data).

```{r b-o sex, fig.dim=c(10,10)}
##### Sex #####
dim_sex <- orientation[, .N, by = c("Sex", "BurialOrientationMean_sin", "BurialOrientationMean_cos")]

ggplot(data = dim_sex) +
  geom_point(aes(BurialOrientationMean_sin, 
                 BurialOrientationMean_cos, 
                 fill = Sex, size = N),
             position = position_jitter(0.08, 0.08),
             shape = 21) +
  scale_fill_gradientn("Sex", colors = rev(pals::brewer.piyg(100))) +
  scale_size_continuous("Count", breaks = c(1, 30, 60, 90, 120), 
                        range = c(3,10)) +
  xlab(expression(sin(theta))) +
  ylab(expression(cos(theta))) +
  coord_equal() +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16))

# Group sex
orientation[, Sex := ifelse(Sex < 0.5, 0, ifelse(Sex > 0.5, 1, Sex))]

# Circular histogram
ggplot(data = orientation,
       aes(x = Bins, fill = as.factor(Sex)),
       color = 'black') +
  geom_bar(color = 'black', position = "dodge") +
  scale_x_continuous(breaks = seq(0, length(seq(0, 360, w))-1, 1),
                     labels = seq(0, 360, w)) +
  scale_y_continuous() +
  scale_fill_manual("Sex", values = rev(pals::brewer.piyg(3))) +
  coord_polar(start = -.2) +
  theme_light() +
  theme(axis.text.y = element_blank(),
        axis.title = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.x = element_text(size = 12))

# Single plots
ggplot(data = orientation,
       aes(x = Bins, fill = as.factor(Sex)),
       color = 'black') +
  geom_bar(color = 'black') +
  scale_x_continuous(breaks = seq(0, length(seq(0, 360, w))-1, 2),
                     labels = c("N", "NE", "E", "SE", "S", "SW", "W", "NW", "N")) +
  scale_y_continuous() +
  scale_fill_manual("Sex", values = paletteer::paletteer_c("grDevices::Tropic", 3)) +
  ylab("Number of individuals") +
  coord_polar(start = -.2) +
  facet_wrap(~as.factor(Sex), scales = "free_y", ) +
  theme_light() +
  theme(axis.title.x = element_blank(),
        legend.position = "none")
```

#### Culture by sex
```{r b-o culture sex, fig.dim=c(12,10)}
# Single plots
orientation_culture[, Sex := ifelse(Sex < 0.5, 0, ifelse(Sex > 0.5, 1, Sex))]

ggplot(data = orientation_culture,
       aes(x = Bins, fill = as.factor(Sex)),
       color = 'black') +
  geom_bar(color = 'black',
           position = "dodge") +
  scale_x_continuous(breaks = seq(0, length(seq(0, 360, w))-1, 2),
                     labels = c("N", "NE", "E", "SE", "S", "SW", "W", "NW", "N")) +
  scale_fill_manual("Sex", values = paletteer::paletteer_c("grDevices::Tropic", 3)) +
  coord_polar(start = -.2) +
  facet_wrap(~Culture, scales = "free_y") +
  ylab("Number of individuals") +
  theme_light() +
  theme(axis.title.x = element_blank(),
        legend.position = "right")
```

### Age
Here we are plotting the data by age. First, we consider the whole time period of the individuals, but then we restrict the data to contain only individuals between 8k to 2k years BP.

Again, for the circular histograms I had to transform the data into factor, by binning the ages into 500 years bins.

```{r b-o age, fig.dim=c(10,10)}
##### Age #####
dim_age <- orientation[, .N, by = c("Age", "BurialOrientationMean_sin", "BurialOrientationMean_cos")]

ggplot(data = dim_age) +
  geom_point(aes(BurialOrientationMean_sin, BurialOrientationMean_cos, fill = Age, size = N),
             position = position_jitter(0.1, 0.1),
             shape = 21) +
  scale_fill_gradientn("Age", colors = pals::coolwarm(100)) +
  scale_size_continuous("Count", breaks = c(1, 20, 50, 75), range = c(3,10)) +
  xlab(expression(sin(theta))) +
  ylab(expression(cos(theta))) +
  coord_equal() +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16))

# Age interval between 8k and 2k BP
ggplot(data = dim_age[between(Age, 2000, 8000)]) +
  geom_point(aes(BurialOrientationMean_sin, BurialOrientationMean_cos, fill = Age, size = N),
             position = position_jitter(0.1, 0.1),
             shape = 21) +
  scale_fill_gradientn("Age", colors = pals::coolwarm(100), limits = c(2000, 8000),
                       breaks = seq(2000, 8000, 2000)) +
  scale_size_continuous("Count", breaks = c(1, 25, 50, 75, 100), range = c(3,10)) +
  xlab(expression(sin(theta))) +
  ylab(expression(cos(theta))) +
  coord_equal() +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16))

# Histograms
orientation$YearBin <- plyr::round_any(orientation$Age, 500)
ggplot(data = orientation[between(Age, 3000, 8000)],
       aes(x = Bins, 
           fill = factor(YearBin, 
                         levels = as.character(rev(seq(3000, 8000, 500))))),
       color = 'black') +
  geom_bar(color = 'black') +
  scale_x_continuous(breaks = seq(0, length(seq(0, 360, w))-1, 1),
                     labels = seq(0, 360, w)) +
  scale_y_continuous() +
  scale_fill_manual("YearBin", values = rev(pals::brewer.piyg(11))) +
  coord_polar(start = -.2) +
  ylab("Count") +
  facet_wrap(~factor(YearBin, 
                     levels = as.character(rev(seq(3000, 8000, 500)))), scales = "free_y") +
  theme_light() +
  theme(axis.title.x = element_blank(),
        legend.position = "none")
```

### Ancestry
For the ancestry, we decided to take the average ancestry value for each individual that is in the same angle position.

#### Western hunter-gatherer ancestry
```{r b-o whg, fig.dim=c(10,10)}
dim_whg <- orientation[, mean(WHG, na.rm = T), by = c("BurialOrientationMean_sin", "BurialOrientationMean_cos")]
dim_whg <- dim_whg[!is.nan(V1),]
names(dim_whg)[3] <- "WHG"

ggplot(data = dim_whg) +
  geom_point(aes(BurialOrientationMean_sin, 
                 BurialOrientationMean_cos, 
                 fill = WHG),
             shape = 21,
             size = 5) +
  scale_fill_gradientn("WHG", 
                       colors = pals::parula(100), 
                       limits = c(0,1)) +
  xlab(expression(sin(theta))) +
  ylab(expression(cos(theta))) +
  coord_equal() +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16))

# Histograms
orientation$WHG_bin <- plyr::round_any(orientation$WHG, 0.25)
ggplot(data = orientation[!is.na(WHG)],
       aes(x = Bins, fill = as.factor(WHG_bin)),
       color = 'black') +
  geom_bar(color = 'black') +
  scale_x_continuous(breaks = seq(0, length(seq(0, 360, w))-1, 1),
                     labels = seq(0, 360, w)) +
  scale_y_continuous() +
  scale_fill_manual("WHG", values = pals::parula(5)) +
  coord_polar(start = -.2) +
  ylab("Count") +
  facet_wrap(~as.factor(WHG_bin), scales = "free_y") +
  theme_light() +
  theme(axis.title.x = element_blank(),
        legend.position = "none")
```

#### Eastern hunter-gatherer ancestry
```{r b-o ehg, fig.dim=c(10,10)}
dim_ehg <- orientation[, mean(EHG, na.rm = T), by = c("BurialOrientationMean_sin", "BurialOrientationMean_cos")]
dim_ehg <- dim_ehg[!is.nan(V1),]
names(dim_ehg)[3] <- "EHG"

ggplot(data = dim_ehg) +
  geom_point(aes(BurialOrientationMean_sin, BurialOrientationMean_cos, fill = EHG),
             shape = 21,
             size = 5) +
  scale_fill_gradientn("EHG", colors = pals::parula(100), limits = c(0,1)) +
  xlab(expression(sin(theta))) +
  ylab(expression(cos(theta))) +
  coord_equal() +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16))

# Histograms
orientation$EHG_bin <- plyr::round_any(orientation$EHG, 0.25)
ggplot(data = orientation[!is.na(EHG)],
       aes(x = Bins, fill = as.factor(EHG_bin)),
       color = 'black') +
  geom_bar(color = 'black') +
  scale_x_continuous(breaks = seq(0, length(seq(0, 360, w))-1, 1),
                     labels = seq(0, 360, w)) +
  scale_y_continuous() +
  scale_fill_manual("EHG", values = pals::parula(5)) +
  coord_polar(start = -.2) +
  ylab("Count") +
  facet_wrap(~as.factor(EHG_bin), scales = "free_y") +
  theme_light() +
  theme(axis.title.x = element_blank(),
        legend.position = "none")
```

#### Caucasus hunter-gatherer ancestry
```{r b-o chg, fig.dim=c(10,10)}
##### Ancestry: CHG ##### 
dim_chg <- orientation[, mean(CHG, na.rm = T), by = c("BurialOrientationMean_sin", "BurialOrientationMean_cos")]
dim_chg <- dim_chg[!is.nan(V1),]
names(dim_chg)[3] <- "CHG"

ggplot(data = dim_chg) +
  geom_point(aes(BurialOrientationMean_sin, 
                 BurialOrientationMean_cos, 
                 fill = CHG),
             shape = 21,
             size = 5) +
  scale_fill_gradientn("CHG", colors = pals::parula(100), limits = c(0,1)) +
  xlab(expression(sin(theta))) +
  ylab(expression(cos(theta))) +
  coord_equal() +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16))

# Histograms
orientation$CHG_bin <- plyr::round_any(orientation$CHG, 0.05)
ggplot(data = orientation[!is.na(CHG)],
       aes(x = Bins, fill = as.factor(CHG_bin)),
       color = 'black') +
  geom_bar(color = 'black') +
  scale_x_continuous(breaks = seq(0, length(seq(0, 360, w))-1, 1),
                     labels = seq(0, 360, w)) +
  scale_y_continuous() +
  scale_fill_manual("CHG", values = pals::parula(5)) +
  coord_polar(start = -.2) +
  ylab("Count") +
  facet_wrap(~as.factor(CHG_bin), scales = "free_y") +
  theme_light() +
  theme(axis.title.x = element_blank(),
        legend.position = "none")
```

#### Anatolian farmers ancestry
```{r b-o LVN, fig.dim=c(10,10)}
##### Ancestry: LVN ##### 
dim_lvn <- orientation[, mean(LVN, na.rm = T), by = c("BurialOrientationMean_sin", "BurialOrientationMean_cos")]
dim_lvn <- dim_lvn[!is.nan(V1),]
names(dim_lvn)[3] <- "LVN"

ggplot(data = dim_lvn) +
  geom_point(aes(BurialOrientationMean_sin, 
                 BurialOrientationMean_cos, 
                 fill = LVN),
             shape = 21,
             size = 5) +
  scale_fill_gradientn("LVN", colors = pals::parula(100), limits = c(0,1)) +
  xlab(expression(sin(theta))) +
  ylab(expression(cos(theta))) +
  coord_equal() +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16))

# Histograms
orientation$LVN_bin <- plyr::round_any(orientation$LVN, 0.25)
ggplot(data = orientation[!is.na(LVN)],
       aes(x = Bins, fill = as.factor(LVN_bin)),
       color = 'black') +
  geom_bar(color = 'black') +
  scale_x_continuous(breaks = seq(0, length(seq(0, 360, w))-1, 1),
                     labels = seq(0, 360, w)) +
  scale_fill_manual("CHG", values = pals::parula(5)) +
  scale_y_continuous() +
  coord_polar(start = -.2) +
  ylab("Count") +
  facet_wrap(~as.factor(LVN_bin), scales = "free_y") +
  theme_light() +
  theme(axis.title.x = element_blank(),
        legend.position = "none")
```

### Mobility
```{r b-o mobility, fig.dim=c(10,10)}
##### Mobility #####
dim_mob <- orientation[, mean(mobility, na.rm = T), by = c("BurialOrientationMean_sin", "BurialOrientationMean_cos")]
dim_mob <- dim_mob[!is.nan(V1),]
names(dim_mob)[3] <- "mobility"

ggplot(data = dim_mob) +
  geom_point(aes(BurialOrientationMean_sin, 
                 BurialOrientationMean_cos, 
                 fill = mobility),
             shape = 21,
             size = 5) +
  scale_fill_gradientn("Mobility", colors = pals::parula(100)) +
  xlab(expression(sin(theta))) +
  ylab(expression(cos(theta))) +
  coord_equal() +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16))

# Histograms
orientation$mob_bin <- plyr::round_any(orientation$mobility, 200)
# data between 0 & 100 (0) - 100 & 300 (200) - 300 & 500 (400) etc.
orientation[, .N, by = mob_bin][order(N)]

ggplot(data = orientation[!is.na(mobility)],
       aes(x = Bins, fill = as.factor(mob_bin)),
       color = 'black') +
  geom_bar(color = 'black') +
  scale_x_continuous(breaks = seq(0, length(seq(0, 360, w))-1, 1),
                     labels = seq(0, 360, w)) +
  scale_y_continuous() +
  scale_fill_manual("Mobility", values = pals::parula(9)) +
  coord_polar(start = -.2) +
  ylab("Count") +
  facet_wrap(~as.factor(mob_bin), scales = "free_y") +
  theme_light() +
  theme(axis.title.x = element_blank(),
        legend.position = "none")

```

