---
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r, warning=FALSE, message=FALSE, include=FALSE}
library(data.table)
library(ggplot2)
library(FactoMineR)
library(circular)
library(CircStats)
library(bpnreg)
library(BIADconnect)
conn <- init.conn()
knitr::opts_chunk$set(warning = FALSE, message = FALSE,
                      root.dir = "~/Library/CloudStorage/OneDrive-UniversityofCopenhagen/Graves/")
```

# Circular statistics

```{r stats data}
#### DATA ####
orientation <- fread("./Data/BurialOrientation.csv", na.strings = "")
```

## Test: circular statistics with burial sides

### Mean and variance

```{r stats test}
#### CIRCULAR MEAN AND VARIANCE ####
## Separate data based on burial side
left <- orientation[, .(Num, BurialOrientationMean, BurialOrientationMin, BurialOrientationMax), by = "BurialSide_left"][BurialSide_left == 1]
left[, BurialSide := "Left"]
right <- orientation[, .(Num, BurialOrientationMean, BurialOrientationMin, BurialOrientationMax), by = "BurialSide_right"][BurialSide_right == 1]
right[, BurialSide := "Right"]
back <- orientation[, .(Num, BurialOrientationMean, BurialOrientationMin, BurialOrientationMax), by = "BurialSide_back"][BurialSide_back == 1]
back[, BurialSide := "Back"]

# Bind together in a new table
sides <- rbind(left[, -1], right[, -1], back[, -1])

# Example data in degrees
sides <- sides[!is.na(BurialOrientationMean)]
left_angles <- sides[BurialSide == "Left", BurialOrientationMean]
right_angles <- sides[BurialSide == "Right", BurialOrientationMean]
back_angles <- sides[BurialSide == "Back", BurialOrientationMean]

# Convert to circular objects
left_circ <- circular(left_angles, units = "degrees", type = "angles", rotation = "clock")
right_circ <- circular(right_angles, units = "degrees", type = "angles", rotation = "clock")
back_circ <- circular(back_angles, units = "degrees", type = "angles", rotation = "clock")

# Mean direction
mean_left <- mean.circular(left_circ)
mean_right <- mean.circular(right_circ)
mean_back <- mean.circular(back_circ)

# Variance direction
var_left <- var.circular(left_circ)
var_right <- var.circular(right_circ)
var_back <- var.circular(back_circ)

# Resultant length (measure of concentration)
rho_left <- rho.circular(left_circ)
rho_right <- rho.circular(right_circ)
rho_back <- rho.circular(back_circ)

## Table
stats <- data.table(Mean = c(mean_left, mean_right, mean_back),
                    Variance = c(var_left, var_right, var_back),
                    Rho = c(rho_left, rho_right, rho_back),
                    Side = c("Left", "Right", "Back"))
stats
```

### Angular distance

Here we create a function to calculate the distance between two angles, taking into account that angles can cross zero.

```{r stats angdist}
# Function to calculate smallest angular distance
angular_distance <- function(a, b) {
  diff <- abs(a - b) %% 360
  return(min(diff, 360 - diff))
}
# Example
angular_distance(350, 10)
angular_distance(190, 0)

# Calculate distances
dist_left_right <- angular_distance(as.numeric(mean_left), as.numeric(mean_right))
dist_left_back <- angular_distance(as.numeric(mean_left), as.numeric(mean_back))
dist_right_back <- angular_distance(as.numeric(mean_right), as.numeric(mean_back))

print(dist_left_right)
print(dist_left_back)
print(dist_right_back)
```

### Visualization

```{r stats viz}
## Circular means
sides_means <- c(mean_left, mean_right, mean_back)
names(sides_means) <- c("Left", "Right", "Back")
d <- list()
for(a in 1:length(sides_means)){
  d[[a]] <- sapply(1:length(sides_means), function(x){
    angular_distance(sides_means[a], sides_means[x])
  })
}
sides_mat <- do.call(cbind, d)
colnames(sides_mat) <- rownames(sides_mat) <- names(sides_means)
sides_mat[lower.tri(sides_mat)] <- NA
sides_dists <- melt(sides_mat)

# Plot burial orientations per country
sides_means_dt <- data.table(Sides = names(sides_means), Mean = sides_means)
sides_means_dt[, Mean := ifelse(Mean < 0, Mean + 360, Mean)]
ggplot(data = sides, aes(x = BurialSide, y = BurialOrientationMean, fill = BurialSide)) +
  geom_violin(width = 1, drop = F) +
  geom_point(data = sides_means_dt, aes(x = Sides, y = Mean), size = 2, shape = 23, fill = "white") +
  scale_fill_manual(values = pals::brewer.set2(3)) +
  ylab("Burial orientation") +
  theme_light() +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Plot pairwise angular distances
ggplot(data = sides_dists[!is.na(sides_dists$value),]) +
  geom_tile(aes(Var1, Var2, fill = value), color = "black") +
  scale_fill_gradientn("Angular distance", colours = rev(pals::brewer.rdylbu(100)), 
                       na.value = "transparent",
                       limits = c(0,180)) + # maximum distance between angles?
                       # guide = guide_colorbar(barwidth = 1, barheight = 15)) +
  geom_text(aes(Var1, Var2, label = round(value, 1)), size = 10, size.unit = "pt") +
  scale_x_discrete(position = "top") +
  ggtitle("Pairwise distance between angles means") +
  coord_equal(expand = F) +
  theme_classic() +
  theme(plot.title = element_text(face = "bold", size = 16),
        axis.text.x = element_text(angle = 45, hjust = 0),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "none",
        axis.text = element_text(size = 14))
```

### P-value

Now we shuffle the values of angles for each category and compute the differences in circular means and do a histogram where we plot the differences in circular means against the true difference. Do it for each of the three pairs. Make sure that each group has the same amount of observations.

```{r stats pval}
n <- min(c(length(left_circ), length(right_circ), length(back_circ)))

diffs <- lapply(1:1000, function(x){
  l <- sample(left_circ, n)
  r <- sample(right_circ, n)
  b <- sample(back_circ, n)
  mean_l <- mean.circular(l)
  mean_r <- mean.circular(r)
  mean_b <- mean.circular(b)
  dist_lr <- angular_distance(as.numeric(mean_l), as.numeric(mean_r))
  dist_lb <- angular_distance(as.numeric(mean_l), as.numeric(mean_b))
  dist_br <- angular_distance(as.numeric(mean_b), as.numeric(mean_r))
  return(list(dist_lr, dist_lb, dist_br))
})

diffs_lr <- unlist(sapply(diffs, "[", 1))
diffs_lb <- unlist(sapply(diffs, "[", 2))
diffs_br <- unlist(sapply(diffs, "[", 3))
diffs <- list(diffs_lr, diffs_lb, diffs_br)

## Plot histograms
hist(diffs_lr, breaks = 100, main = "Difference between left and right orientations", xlab = "Angular distance"); abline(v = dist_left_right, col= "red", lwd = 3)

hist(diffs_lb, breaks = 100, main = "Difference between left and back orientations", xlab = "Angular distance"); abline(v = dist_left_back, col= "red", lwd = 3)

hist(diffs_br, breaks = 100, main = "Difference between back and right orientations", xlab = "Angular distance"); abline(v = dist_right_back, col= "red", lwd = 3)

## Calculate p-value
# P = (X + 1) / (N + 1)
obs <- list(dist_left_right, dist_left_back, dist_right_back)
pval <- lapply(1:length(diffs), function(x){
  X <- sum(abs(diffs[[x]]) >= abs(obs[[x]]))
  N <- length(diffs[[x]])
  return((X + 1) / (N + 1))
})
names(pval) <- c("left_right", "left_back", "back_right")
pval
```

### Similarity

We can also calculate a similarity index that takes into account the distance between the means of the different categories.

```{r stats similarity}
# Similarity index
similarity_index <- function(d) {
  return(1 - (d / 180))
}

sim_left_right <- similarity_index(dist_left_right)
sim_left_back <- similarity_index(dist_left_back)
sim_right_back <- similarity_index(dist_right_back)

print(sim_left_right)
print(sim_left_back)
print(sim_right_back)
```

### Watson-William test and circular ANOVA

```{r stats anova}
## Watson-Williams test
# Combine data
all_angles <- c(left_angles, right_angles, back_angles) #right_angles, 
group <- factor(c(rep("left", length(left_angles)),
                  rep("right", length(right_angles)),
                  rep("back", length(back_angles))))

## Mardia-Watson-Wheeler test (non-parametric)
mww_result1 <- watson.two.test(left_circ, right_circ)  # pairwise example
mww_result2 <- watson.two.test(left_circ, back_circ)
mww_result3 <- watson.two.test(right_circ, back_circ)
print(mww_result1)
print(mww_result2)
print(mww_result3)

## Watson-Williams test
watson.williams.test(circular(all_angles, units = "degrees", type = "angles", rotation = "clock"), group)

## Circular ANOVA 
aov.circular(circular(all_angles, units = "degrees", type = "angles", rotation = "clock"), group)
```

Now that we have tested different statistics using the left, right, and back angles, we can do the same for other categories, like periods, cultures, and countries.

## Country

```{r stats country, fig.dim=c(12,10)}
## Circular means
countries <- split(orientation[!(Country %in% c("Armenia", "Kazakhstan"))][, .(BurialOrientationMean, Country)], by = "Country")
countries_means <- lapply(countries, function(x){
  mean.circular(circular(x$BurialOrientationMean, units = "degrees", type = "angles", rotation = "clock"))
})
countries_means <- unlist(countries_means)
countries_means <- countries_means[gtools::mixedsort(names(countries_means))]

d <- list()
for(a in 1:length(countries_means)){
  d[[a]] <- sapply(1:length(countries_means), function(x){
    angular_distance(countries_means[a], countries_means[x])
  })
}
countries_mat <- do.call(cbind, d)
colnames(countries_mat) <- rownames(countries_mat) <- names(countries_means)
countries_mat[lower.tri(countries_mat)] <- NA
countries_dists <- setDT(reshape2::melt(countries_mat))

# Plot burial orientations per country
countries_means_dt <- data.table(Country = names(countries_means), Mean = countries_means)
countries_means_dt[, Mean := ifelse(Mean < 0, Mean + 360, Mean)]

ggplot(data = orientation[!(Country %in% c("Armenia", "Kazakhstan"))], aes(x = Country, y = BurialOrientationMean, fill = Country)) +
  geom_violin(width = 1.5, drop = F) +
  geom_point(data = countries_means_dt, aes(x = Country, y = Mean), size = 2, shape = 23, fill = "white") +
  scale_fill_manual(values = pals::brewer.set2(19)) +
  ylab("Burial orientation") +
  theme_light() +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Plot pairwise angular distances
countries_dists <- countries_dists[!is.na(countries_dists$value),]

ggplot(data = countries_dists) +
  geom_tile(aes(Var1, Var2, fill = value), color = "black") +
  scale_fill_gradientn("Angular distance", colours = rev(pals::brewer.rdylbu(100)), 
                       na.value = "transparent",
                       limits = c(0,180)) + # maximum distance between angles?
  geom_text(aes(as.factor(Var1), as.factor(Var2), label = round(value, 1)), size = 10, size.unit = "pt") +
  scale_x_discrete(position = "top") +
  scale_y_discrete(position = "left") +
  ggtitle("Pairwise distance between angles means") +
  coord_equal(expand = F) +
  theme_classic() +
  theme(plot.title = element_text(face = "bold", size = 16),
        axis.text.x = element_text(angle = 45, hjust = 0),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "none",
        axis.text = element_text(size = 14),
        plot.margin = margin(l = 0, t = 10, b = 10, r = 0))

```

## Period

```{r stats period, fig.dim=c(10, 8)}
period_order <- c("Mesolithic",
                  "Mesolithic/Neolithic",
                  "Neolithic",
                  "Eneolithic",
                  "Eneolithic/Bronze Age",
                  "Bronze Age",
                  "Bronze Age/Iron Age",
                  "Iron Age")
orientation$Period <- factor(orientation$Period,
                             levels = period_order)

## Circular means
period <- split(orientation[Period != "Iron Age"]
                [, .(BurialOrientationMean, Period)], 
                by = "Period")
period_means <- lapply(period, function(x){
  mean.circular(circular(x$BurialOrientationMean, 
                         units = "degrees", 
                         type = "angles", 
                         rotation = "clock"))
})
period_means <- unlist(period_means)
period_means <- period_means[period_order]
d <- list()
for(a in 1:length(period_means)){
  d[[a]] <- sapply(1:length(period_means), function(x){
    angular_distance(period_means[a], period_means[x])
  })
}
period_mat <- do.call(cbind, d)
colnames(period_mat) <- rownames(period_mat) <- names(period_means)
period_mat[lower.tri(period_mat)] <- NA
period_dists <- setDT(reshape2::melt(period_mat))

# Plot burial orientations per period
period_means_dt <- data.table(Period = names(period_means), Mean = period_means)
period_means_dt[, Mean := ifelse(Mean < 0, Mean + 360, Mean)]

ggplot(data = orientation[Period != "Iron Age"], aes(x = Period, y = BurialOrientationMean, fill = Period)) +
  geom_violin(width = 1, drop = F) +
  geom_point(data = period_means_dt[Period != "Iron Age"], aes(x = Period, y = Mean), size = 2, shape = 23, fill = "white") +
  scale_fill_manual(values = pals::brewer.set2(8)) +
  ylab("Burial orientation") +
  theme_light() +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Plot pairwise angular distances
ggplot(data = period_dists[!is.na(period_dists$value),]) +
  geom_tile(aes(Var1, Var2, fill = value), color = "black") +
  scale_fill_gradientn("Angular distance", 
                       colours = rev(pals::brewer.rdylbu(100)), 
                       na.value = "transparent",
                       limits = c(0, 180)) +
  geom_text(aes(Var1, Var2, label = round(value, 1)), 
            size = 10, size.unit = "pt") +
  scale_x_discrete(position = "top") +
  ggtitle("Pairwise distance between angles means") +
  coord_equal(expand = F) +
  theme_classic() +
  theme(plot.title = element_text(face = "bold", size = 16),
        axis.text.x = element_text(angle = 45, hjust = 0),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "none",
        axis.text = element_text(size = 14))

```

## Culture

```{r stats culture, fig.dim=c(15,10)}
# Plot only the 10 most frequent cultures
culture_keep <- orientation[!is.na(Culture)][, .N, by = "Culture"][order(N, decreasing = T)][, Culture][1:10]

## Circular means
culture <- split(orientation[Culture %in% culture_keep]
                 [, .(BurialOrientationMean, Culture)], 
                 by = "Culture")
culture_means <- lapply(culture, function(x){
  mean.circular(circular(x$BurialOrientationMean, 
                         units = "degrees", 
                         type = "angles", 
                         rotation = "clock"),
                na.rm = T)
})
culture_means <- unlist(culture_means)
d <- list()
for(a in 1:length(culture_means)){
  d[[a]] <- sapply(1:length(culture_means), function(x){
    angular_distance(culture_means[a], culture_means[x])
  })
}
culture_mat <- do.call(cbind, d)
colnames(culture_mat) <- rownames(culture_mat) <- names(culture_means)
culture_mat[lower.tri(culture_mat)] <- NA
culture_dists <- setDT(reshape2::melt(culture_mat))

# Plot burial orientations per period
culture_means_dt <- data.table(Culture = names(culture_means), 
                               Mean = culture_means)
culture_means_dt[, Mean := ifelse(Mean < 0, Mean + 360, Mean)]
culture_means_dt
ggplot(data = orientation[Culture %in% culture_keep], 
       aes(x = Culture, y = BurialOrientationMean, 
           fill = Culture)) +
  geom_violin(width = 1, drop = F) +
  geom_point(data = culture_means_dt, aes(x = Culture, y = Mean), 
             size = 2, shape = 23, fill = "white") +
  scale_fill_manual(values = pals::brewer.set2(10)) +
  ylab("Burial orientation") +
  theme_light() +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Plot pairwise angular distances
ggplot(data = culture_dists[!is.na(culture_dists$value),]) +
  geom_tile(aes(Var1, Var2, fill = value), color = "black") +
  scale_fill_gradientn("Angular distance", 
                       colours = rev(pals::brewer.rdylbu(100)), 
                       na.value = "transparent",
                       limits = c(0, 180)) +
  geom_text(aes(Var1, Var2, label = round(value, 1)), 
            size = 10, size.unit = "pt") +
  scale_x_discrete(position = "top") +
  ggtitle("Pairwise distance between angles means") +
  coord_equal(expand = F) +
  theme_classic() +
  theme(plot.title = element_text(face = "bold", size = 16),
        axis.text.x = element_text(angle = 45, hjust = 0),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "none",
        axis.text = element_text(size = 14))

```

### Difference between Corded Ware and Bell Beaker cultures

Here we are going to statistically test for differences in the orientations of Corded Ware and Bell Beaker cultures.

First of all, we look at the summary statistics for the two cultures. We can see that the means (mu) are quite different but the Corded Ware orientations are spread across the whole circle (there is very low concentration). This can be seen both by the rho value (mean resultant length) that is close to 0, and by the variance (var) value that is close to 1 ($var = 1 - rho$). The standard deviation runs from 0 to infinity, where higher values indicate a larger dispersion.

We can now try and check for differences in mean direction and concentration compared to a uniform distribution.

```{r stats cw-bb}
## Orientations
cordedware <- culture[["Corded Ware/Single Grave/Battle Axe/Fatyanovo"]]$BurialOrientationMean
cordedware <- circular(cordedware,
                      type = "angles",
                      units = "degrees",
                      rotation = "clock")

bellbeaker <- culture[["Bell Beaker complex"]]$BurialOrientationMean
bellbeaker <- circular(bellbeaker,
                      type = "angles",
                      units = "degrees",
                      rotation = "clock")

## Summary statistics of the two cultures
summary.dat <- data.frame(culture = c("Corded Ware", "Bell Beaker"),
                          mu = c(mean.circular(cordedware),
                                 mean.circular(bellbeaker)),
                          rho = c(rho.circular(cordedware),
                                  rho.circular(bellbeaker)),
                          var = c(var.circular(cordedware),
                                  var.circular(bellbeaker)),
                          sd = c(sd.circular(cordedware),
                                 sd.circular(bellbeaker)))
summary.dat

## Calculate mean direction and concentration from uniform distribution
cw_unif <- lapply(1:1000, function(x){
  r <- abs(rcircularuniform(n = length(cordedware),
                            control.circular = list(
                              units = "degrees",
                              rotation = "clock")))
  m <- mean.circular(r)
  k <- rho.circular(r)
  return(list(m, k))
})
  
bb_unif <- lapply(1:1000, function(x){
  r <- abs(rcircularuniform(n = length(bellbeaker),
                            control.circular = list(
                              units = "degrees",
                              rotation = "clock")))
  m <- mean.circular(r)
  k <- rho.circular(r)
  return(list(m, k))
})

cw_means <- sapply(cw_unif, "[", 1)
cw_rho <- sapply(cw_unif, "[", 2)

bb_means <- sapply(bb_unif, "[", 1)
bb_rho <- sapply(bb_unif, "[", 2)

hist(as.numeric(cw_means), breaks = 100, main = "Corded Ware", xlab = "Mean direction (mu)"); abline(v = mean.circular(cordedware), col = "red")
hist(as.numeric(cw_rho), breaks = 100, main = "Corded Ware", xlab = "Concentration (rho)", xlim = c(0,1)); abline(v = rho.circular(cordedware), col = "red")

hist(as.numeric(bb_means), breaks = 100, main = "Bell Beaker", xlab = "Mean direction (mu)"); abline(v = mean.circular(bellbeaker), col = "red")
hist(as.numeric(bb_rho), breaks = 100, main = "Bell Beaker", xlab = "Concentration (rho)", xlim = c(0,1)); abline(v = rho.circular(bellbeaker), col = "red")

# Similarity index (without folding over the horizontal axis)
obs <- culture_dists[Var2 == "Corded Ware/Single Grave/Battle Axe/Fatyanovo" & Var1 == "Bell Beaker complex"]$value
diff <- similarity_index(obs)
diff
paste0((1-round(diff,2))*100, "% different")
```

Now for modelling the differences between the two cultures, we can try using the `bpnreg` R package. This package performs a Bayesian circular regression model based on the projected normal distribution. The angles must be in radians. This is equivalent to an ANOVA.

In this model, the Bell Beaker is the reference culture. Thus, the value of the intercept for the categorical variables represents the mean of the Bell Beaker orientation.

Further reading:

-   <https://www.frontiersin.org/journals/psychology/articles/10.3389/fpsyg.2018.02040/full>

-   <https://bpspsychub.onlinelibrary.wiley.com/doi/full/10.1111/bmsp.12108>

-   <https://dspace.library.uu.nl/bitstream/handle/1874/361637/1_s2.0_S0022249616300827_main.pdf?sequence=1>

```{r stats cw-bb bpnreg, results='hide'}
# Data for regression
circ.dat <- data.frame(Angles = c(as.numeric(cordedware)*pi/180,
                                  as.numeric(bellbeaker)*pi/180),
                       Culture = c(rep("Corded Ware", length(cordedware)),
                                   rep("Bell Beaker", length(bellbeaker))))
circ.dat$Culture <- as.factor(circ.dat$Culture)

fit <- bpnr(Angles ~ 1 + Culture, data = circ.dat, 
            its = 10000, burn = 100, n.lag = 3, seed = 1234)
```
```{r stats cw-bb bpnreg2}
# Means
aov.dt <- setDT(data.frame(coef_circ(fit, "categorical", "radians")$Means))
aov.dt[, Culture := as.factor(c("Bell Beaker", "Corded Ware"))]

# Plot
plot.circular(c(aov.dt$LB[1], aov.dt$UB[1]), pch = 16, 
              rotation = "clock", zero = pi/2)
points.circular(c(aov.dt$LB[2], aov.dt$UB[2]), pch = 5, 
                rotation = "clock", zero = pi/2)
points.circular(aov.dt$mean[1], pch = 16, col = "red", 
                rotation = "clock", zero = pi/2)
points.circular(aov.dt$mean[2], pch = 18, col = "blue", 
                rotation = "clock", zero = pi/2)

aov.dt[, .(Culture, deg(.SD)), .SDcols = c("mean", "mode", "sd", "LB", "UB")]

# Differences
diff.dt <- setDT(data.frame(coef_circ(fit, "categorical", units = "degrees")$Differences))
diff.dt[, Difference := "Bell Beaker/Corded Ware"]
diff.dt[, .(Difference, mean, mode, sd, LB, UB)]

# Calculate angular distance between means
angular_distance(aov.dt$mean[1]*180/pi, aov.dt$mean[2]*180/pi)
```

In theory the HPDs of the two cultures overlap so the difference in means shouldn't be significant. When plotting the results, we can see that the two means are quite different and the HPDs overlap only because the spread of the Corded Ware orientations is extremely large. If we look at the differences table, the LB and UP of the HPD for the difference doesn't cross zero, so the difference between the two cultures is significant.

We can also try these other two tests that according to this paper seem to perform well with circular data, without falling too much into Type-1 and Type-2 errors.

<https://doi.org/10.1038/s41598-021-99299-5>

```{r stats watson.two manova}
## Watson's U2 test
watson.two(cordedware, bellbeaker)

## MANOVA
circ.dat$Sin <- sin(circ.dat$Angles)
circ.dat$Cos <- cos(circ.dat$Angles)
maov <- manova(cbind(Sin, Cos) ~ 1 + Culture, circ.dat)
summary(maov)
```

So when the Watson's U^2^ test has a small p-value or a large test statistic, it means that the two groups are **NOT** homogeneious, so they are significantly different.

### Difference between Lažňany and Únětice cultures

Now let's double check that this interpretation is correct by comparing two other very similar cultures, e.g., Lažňany and Únětice. Únětice has a larger spread than Lažňany, but the means have a similar direction.

```{r stats lz-un bpnreg}
laznany <- circular(culture[["Lažňany"]]$BurialOrientationMean,
                    type = "angles",
                    units = "degrees",
                    rotation = "clock")
unetice <- circular(culture[["Únětice"]]$BurialOrientationMean,
                    type = "angles",
                    units = "degrees",
                    rotation = "clock")

summary.dat2 <- data.frame(culture = c("Lažňany", "Únětice"),
                          mu = c(mean.circular(laznany),
                                 mean.circular(unetice)),
                          rho = c(rho.circular(laznany),
                                  rho.circular(unetice)),
                          var = c(var.circular(laznany),
                                  var.circular(unetice)),
                          sd = c(sd.circular(laznany),
                                 sd.circular(unetice)))
summary.dat2
```
```{r stats lz-un bpnreg2, results = 'hide'}
# Data for regression
circ.dat2 <- data.frame(Angles = c(as.numeric(laznany)*pi/180,
                                  as.numeric(unetice)*pi/180),
                       Culture = c(rep("Lažňany", length(laznany)),
                                   rep("Únětice", length(unetice))))
circ.dat2$Culture <- as.factor(circ.dat2$Culture)

fit2 <- bpnr(Angles ~ 1 + Culture, data = circ.dat2, 
            its = 10000, burn = 100, n.lag = 3, seed = 1234)
```
```{r stats lz-un bpnreg3}
# Means
aov.dt2 <- setDT(data.frame(coef_circ(fit2, "categorical", "radians")$Means))
aov.dt2[, Culture := as.factor(c("Lažňany", "Únětice"))]

# Plot
plot.circular(c(aov.dt2$LB[1], aov.dt2$UB[1]), pch = 16, 
              rotation = "clock", zero = pi/2)
points.circular(c(aov.dt2$LB[2], aov.dt2$UB[2]), pch = 5, 
                rotation = "clock", zero = pi/2)
points.circular(aov.dt2$mean[1], pch = 16, col = "red", 
                rotation = "clock", zero = pi/2)
points.circular(aov.dt2$mean[2], pch = 18, col = "blue", 
                rotation = "clock", zero = pi/2)

aov.dt2[, .(Culture, deg(.SD)), .SDcols = c("mean", "mode", "sd", "LB", "UB")]

# Differences
diff.dt2 <- setDT(data.frame(coef_circ(fit2, "categorical", units = "degrees")$Differences))
diff.dt2[, Difference := "Lažňany/Únětice"]
diff.dt2[, .(Difference, mean, mode, sd, LB, UB)]

# Calculate distance between means
angular_distance(aov.dt2$mean[1]*180/pi, aov.dt2$mean[2]*180/pi)

## Watson's U2 test
watson.two(laznany, unetice)

## MANOVA
circ.dat2$Sin <- sin(circ.dat2$Angles)
circ.dat2$Cos <- cos(circ.dat2$Angles)
maov2 <- manova(cbind(Sin, Cos) ~ 1 + Culture, circ.dat2)
summary(maov2)
```

In this case we can see that the HPDs for the two cultures completely overlap and their means differ by only \~3 degrees. Furthermore, the LB and UB of the HPD for the difference crosses 0. Therefore, they are not significantly different from one another.

The Watson's U^2^ test also confirms this interpretation, as the p-value is greater than 0.10 and the test statistic is small.

The MANOVA test seems to be incorrect in this case.

```{r stats cultures all, echo=FALSE, eval=F}
circ.dat.all <- lapply(culture, function(x){
  x$BurialOrientationMean <- x$BurialOrientationMean*pi/180
  return(x)
})
circ.dat.all <- rbindlist(circ.dat.all)
circ.dat.all$Culture <- factor(circ.dat.all$Culture, 
                               labels = c("Linearbandkeramik",
                                          "Lažňany",
                                          "Sopot",
                                          "Bell Beaker",
                                          "Únětice",
                                          "Corded Ware",
                                          "Baden",
                                          "Dnieper-Donets",
                                          "Trichterbecher",
                                          "Baltic Meso-Neolithic"),
                               levels = unique(circ.dat.all$Culture))

# Model
fit.all <- bpnr(BurialOrientationMean ~ 1 + Culture, 
                data = circ.dat.all, its = 10000, burn = 1000, 
                n.lag = 3, seed = 1234)

# Means
aov.dt.all <- setDT(data.frame(coef_circ(fit.all, "categorical", "radians")$Means))
aov.dt.all[, Culture := rownames(coef_circ(fit.all, "categorical", "radians")$Means)]
c1 <- sapply(strsplit(aov.dt.all$Culture, "Culture"), "[", 2)
c2 <- sapply(strsplit(aov.dt.all$Culture, "Culture"), "[", 3)
c1[is.na(c1)] <- c2[is.na(c2)] <- "Linearbandkeramik"
aov.dt.all[, Culture := paste0(c1, "/", c2)]
aov.dt.all[, .(Culture, deg(.SD)), .SDcols = c("mean", "mode", "sd", "LB", "UB")]

# Differences
diff.dt.all <- setDT(data.frame(coef_circ(fit.all, "categorical", units = "degrees")$Differences))
diff.dt.all[, Difference := rownames(coef_circ(fit.all, "categorical", units = "degrees")$Differences)]
c1 <- sapply(strsplit(diff.dt.all$Difference, "Culture"), "[", 2)
c2 <- sapply(strsplit(diff.dt.all$Difference, "Culture"), "[", 3)
c2[is.na(c2)] <- "Linearbandkeramik"
diff.dt.all[, Difference := paste0(c1, "/", c2)]
diff.dt.all[, .(Difference, mean, mode, sd, LB, UB)]
diff.dt.all[, Significant := ifelse(between(0, LB, UB), "No", "Yes")]
```

## Sex

```{r stats sex}
# Group sex
orientation[, Sex := ifelse(Sex < 0.5, 0, ifelse(Sex > 0.5, 1, Sex))]

## Circular means
sex <- split(orientation[, .(BurialOrientationMean, Sex = as.factor(Sex))], by = "Sex")
sex_means <- lapply(sex, function(x){
  mean.circular(circular(x$BurialOrientationMean, 
                         units = "degrees", 
                         type = "angles", 
                         rotation = "clock"))
})
sex_means <- unlist(sex_means)
sex_means <- sex_means[gtools::mixedsort(names(sex_means))]
d <- list()
for(a in 1:length(sex_means)){
  d[[a]] <- sapply(1:length(sex_means), function(x){
    angular_distance(sex_means[a], sex_means[x])
  })
}
sex_mat <- do.call(cbind, d)
colnames(sex_mat) <- rownames(sex_mat) <- paste0("Prob_", names(sex_means))
sex_mat[lower.tri(sex_mat)] <- NA
sex_mat
sex_dists <- reshape2::melt(sex_mat)

# Plot burial orientations per period
sex_means_dt <- data.table(Sex = names(sex_means), Mean = sex_means)
sex_means_dt[, Mean := ifelse(Mean < 0, Mean + 360, Mean)]
ggplot(data = orientation, 
       aes(x = as.factor(Sex), y = BurialOrientationMean, 
           fill = as.factor(Sex))) +
  geom_violin(width = 1, drop = F) +
  geom_point(data = sex_means_dt, aes(x = Sex, y = Mean), 
             size = 2, shape = 23, fill = "white") +
  scale_fill_manual(values = rev(pals::brewer.piyg(3))) +
  ylab("Burial orientation") +
  theme_light() +
  theme(legend.position = "none",
        axis.title.x = element_blank())

# Plot pairwise angular distances
ggplot(data = sex_dists[!is.na(sex_dists$value),]) +
  geom_tile(aes(Var1, Var2, fill = value), color = "black") +
  scale_fill_gradientn("Angular distance", colours = rev(pals::brewer.rdylbu(100)), 
                       na.value = "transparent",
                       limits = c(0, 180)) +
   geom_text(aes(Var1, Var2, label = round(value, 1)), 
            size = 10, size.unit = "pt") +
  scale_x_discrete(position = "top") +
  ggtitle("Pairwise distance between angles means") +
  coord_equal(expand = F) +
  theme_classic() +
  theme(plot.title = element_text(face = "bold", size = 16),
        axis.text.x = element_text(angle = 45, hjust = 0),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "none",
        axis.text = element_text(size = 14))
```
```{r stats sex bpnreg, results = "hide"}
## Bayesian circular regression
sex.dt <- orientation[, .(Angles = BurialOrientationMean, Sex)]
sex.dt <- sex.dt[!is.na(Angles)]
sex.dt$Sex <- factor(sex.dt$Sex, levels = c("0", "0.5", "1"))

fit.sex <- bpnr(Angles ~ 1 + Sex, data = sex.dt, 
            its = 10000, burn = 100, n.lag = 3, seed = 1234)
```
```{r stats sex bpnreg2}
# Means
sex.aov.dt <- setDT(data.frame(coef_circ(fit.sex, "categorical", "radians")$Means))[-4,]
sex.aov.dt[, Sex := levels(sex.dt$Sex)]

# Plot
plot.circular(c(sex.aov.dt$LB[1], sex.aov.dt$UB[1]), pch = 1, 
              col = 'red', rotation = "clock", zero = pi/2)
points.circular(c(sex.aov.dt$LB[2], sex.aov.dt$UB[2]), pch = 5, 
                col = 'blue', rotation = "clock", zero = pi/2)
points.circular(c(sex.aov.dt$LB[3], sex.aov.dt$UB[3]), pch = 2, 
                col = 'green', rotation = "clock", zero = pi/2)
points.circular(sex.aov.dt$mean[1], pch = 16, col = "red", 
                rotation = "clock", zero = pi/2)
points.circular(sex.aov.dt$mean[2], pch = 18, col = "blue", 
                cex = 1.5, rotation = "clock", zero = pi/2)
points.circular(sex.aov.dt$mean[3], pch = 17, col = "green", 
                rotation = "clock", zero = pi/2)

sex.aov.dt[, .(Sex, deg(.SD)), .SDcols = c("mean", "mode", "sd", "LB", "UB")]

# Differences
diff.dt <- setDT(data.frame(coef_circ(fit.sex, "categorical", units = "degrees")$Differences))
diff.dt[, Difference := c("0/0.5",
                          "0/1",
                          "0.5/1")]
diff.dt[, .(Difference, mean, mode, sd, LB, UB)]
```

## Circular linear regression models

For data that is not categorical (e.g., ancestry and mobility values), we have to use a regression model that is specific to angular data.

This regression models both the mean direction and the dispersion simultaneously. The model assumes that a circular response variable $\theta$ has a von Mises distribution with concentration parameter $\kappa$, and mean direction ($\mu$) related to a vector of linear predictor variables according to the relationship: $\mu + 2*atan(\beta*x)$, where $\mu$ and $\beta$ are unknown parameters, $\beta$ being a vector of regression coefficients, and $2*atan()$ being the link function.

The von Mises distribution is a function of $\mu$ and $\kappa$, where $\mu$ is the mean direction and $\kappa$ is the concentration parameter. A value $\kappa = 0$ corresponds to the circular uniform distribution; for $\kappa > 0$, the density is unimodal and symmetrical about $\mu$, and increasingly concentrated as $\kappa$ increases. For $\kappa \ge 2$ the density at the antipode $\mu + \pi$ is effectively zero, and the distribution is well approximated by a normal distribution with variance $1/\kappa$.

We can start with an initial value for $\beta$, to calculate $\mu$ and $\kappa$. These estimates can then be used to update the $\beta$ value, and then to update $\mu$ and $\kappa$, analogously to the ML equations for an independent, identically distributed (i.i.d.) von Mises sample. This cycle is repeated until convergence. If $\beta = 0$, we have a null model where all observations cluster around a common mean direction $\mu$. However, as the elements of $\beta$ become infinite, the use of any monotone link function also leads to a null model with mean direction $\mu ± \pi$. In such cases, the likelihood will have a peak in the vicinity of zero, and attain a similar magnitude as the elements of $\beta$ tend to $± \infty$. **The optimization process is simplified if the x are centered at their means**.

**So should we scale our ancestry values?** (Whether we order the values or scale them, model results don't seem to change)

Further reads: <https://www.jstor.org/stable/2532334?seq=2>; <http://iopscience.iop.org/article/10.1088/1742-6596/1988/1/012108>

The `circular` package, has a `lm.circular` function that can be used for circular and linear data (c-l). We are going to test this function using the ancestry values associated to the grave individuals.

### CHG

```{r stats chg, fig.dim=c(10,10)}
y <- circular(rad(orientation[!is.na(CHG)]$BurialOrientationMean), units = "radians", type = "angles", rotation = "clock")
x <- cbind(orientation[!is.na(CHG)]$CHG, rep(1,length(orientation[!is.na(CHG)]$CHG)))
colnames(x) <- c("CHG", "Intercept")
mu <- as.numeric(mean.circular(y))
reg.chg <- lm.circular(y = y, x = x, type = "c-l", init = c(0,1))
reg.chg
```

We can see that CHG has a significant positive relationship with the mean direction of the burials. This test was made by using all the data as is. But what if we repeat the analysis by averaging the CHG value of the individuals that have the same burial orientation? Similarly to what we did to plot them.

```{r stats chg2}
## Based on average value across individuals with the same angle
dim_chg <- orientation[, mean(CHG, na.rm = T), by = c("BurialOrientationMean_sin", "BurialOrientationMean_cos")]
dim_chg <- dim_chg[!is.nan(V1),]
names(dim_chg)[3] <- "CHG"
# Get the angle from sin and cos (in radians)
dim_chg[, BurialOrientationMean := sapply(1:nrow(dim_chg), function(x) atan2(BurialOrientationMean_sin[x], BurialOrientationMean_cos[x]))]

# New variables for the linear regression model
y2 <- circular(dim_chg$BurialOrientationMean, units = "radians", type = "angles", rotation = "clock")
x2 <- cbind(dim_chg$CHG, rep(1, nrow(dim_chg)))
colnames(x2) <- c("CHG", "Intercept")
mu <- as.numeric(mean.circular(y2))
m <- mean(x2[,1])

# Model
reg.chg2 <- lm.circular(y = y2, x = x2, type = "c-l", init = c(0,1))
reg.chg2

# Plot
mu.df <- data.frame(mu = c(mu, reg.chg2$mu), type = c("Observed", "Estimated"))

ggplot() +
  geom_point(data = dim_chg,
             aes(BurialOrientationMean_sin, BurialOrientationMean_cos, fill = CHG),
             shape = 21,
             size = 5,
             ) +
  geom_point(data = mu.df, aes(sin(mu), cos(mu), shape = type), size = 8) +
  scale_fill_gradientn("CHG", colors = pals::parula(100)) +
  scale_shape_manual(expression(mu), values = c(7, 8)) +
  xlab(expression(sin(theta))) +
  ylab(expression(cos(theta))) +
  coord_equal() +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16))
```

### WHG

```{r stats whg, fig.dim = c(10,10)}
y <- circular(rad(orientation[!is.na(WHG)]$BurialOrientationMean), units = "radians", type = "angles", rotation = "clock")
x <- cbind(orientation[!is.na(WHG)]$WHG, rep(1,length(orientation[!is.na(WHG)]$WHG)))
colnames(x) <- c("WHG", "Intercept")
mu <- as.numeric(mean.circular(y))
reg.whg <- lm.circular(y = y, x = x, type = "c-l", init = c(0,1))
reg.whg
```

### EHG

```{r stats ehg}
y <- circular(rad(orientation[!is.na(EHG)]$BurialOrientationMean), units = "radians", type = "angles", rotation = "clock")
x <- cbind(orientation[!is.na(EHG)]$EHG, rep(1,length(orientation[!is.na(EHG)]$EHG)))
colnames(x) <- c("EHG", "Intercept")
mu <- as.numeric(mean.circular(y))
reg.ehg <- lm.circular(y = y, x = x, type = "c-l", init = c(0,1))
reg.ehg
```

### LVN

```{r stats lvn}
y <- circular(rad(orientation[!is.na(LVN)]$BurialOrientationMean), units = "radians", type = "angles", rotation = "clock")
x <- cbind(orientation[!is.na(LVN)]$LVN, rep(1,length(orientation[!is.na(LVN)]$LVN)))
colnames(x) <- c("LVN", "Intercept")
mu <- as.numeric(mean.circular(y))
reg.lvn <- lm.circular(y = y, x = x, type = "c-l", init = c(0,1))
reg.lvn
```

### Mobility

```{r stats mobility}
y <- circular(rad(orientation[!is.na(mobility)]$BurialOrientationMean), units = "radians", type = "angles", rotation = "clock")
x <- cbind(orientation[!is.na(mobility)]$mobility, rep(1,length(orientation[!is.na(mobility)]$mobility)))
colnames(x) <- c("Mobility", "Intercept")
mu <- as.numeric(mean.circular(y))
reg.mob <- lm.circular(y = y, x = x, type = "c-l", init = c(0,1))
reg.mob
```
