---
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r, warning=FALSE, message=FALSE, include=FALSE}
library(data.table)
library(DT)
library(ggplot2)
library(FactoMineR)
library(circular)
library(CircStats)
library(circlize)
library(bpnreg)
library(BIADconnect)
library(webshot)
conn <- init.conn()
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

# Circular statistics

```{r stats data}
#### DATA ####
orientation <- fread("./Data/BurialOrientation.csv", na.strings = "")
```

## Test: circular statistics with burial sides

### Mean and variance

```{r stats test}
#### CIRCULAR MEAN AND VARIANCE ####
## Separate data based on burial side
left <- orientation[, .(Num, BurialOrientationMean, BurialOrientationMin, BurialOrientationMax), by = "BurialSide_left"][BurialSide_left == 1]
left[, BurialSide := "Left"]
right <- orientation[, .(Num, BurialOrientationMean, BurialOrientationMin, BurialOrientationMax), by = "BurialSide_right"][BurialSide_right == 1]
right[, BurialSide := "Right"]
back <- orientation[, .(Num, BurialOrientationMean, BurialOrientationMin, BurialOrientationMax), by = "BurialSide_back"][BurialSide_back == 1]
back[, BurialSide := "Back"]

# Bind together in a new table
sides <- rbind(left[, -1], right[, -1], back[, -1])

# Example data in degrees
sides <- sides[!is.na(BurialOrientationMean)]
left_angles <- sides[BurialSide == "Left", BurialOrientationMean]
right_angles <- sides[BurialSide == "Right", BurialOrientationMean]
back_angles <- sides[BurialSide == "Back", BurialOrientationMean]

# Convert to circular objects
left_circ <- circular(left_angles, units = "degrees", type = "angles", rotation = "clock")
right_circ <- circular(right_angles, units = "degrees", type = "angles", rotation = "clock")
back_circ <- circular(back_angles, units = "degrees", type = "angles", rotation = "clock")

# Mean direction
mean_left <- mean.circular(left_circ)
mean_right <- mean.circular(right_circ)
mean_back <- mean.circular(back_circ)

# Variance direction
var_left <- var.circular(left_circ)
var_right <- var.circular(right_circ)
var_back <- var.circular(back_circ)

# Resultant length (measure of concentration)
rho_left <- rho.circular(left_circ)
rho_right <- rho.circular(right_circ)
rho_back <- rho.circular(back_circ)

## Table
stats <- data.table(Mean = c(mean_left, mean_right, mean_back),
                    Variance = c(var_left, var_right, var_back),
                    Rho = c(rho_left, rho_right, rho_back),
                    Side = c("Left", "Right", "Back"))
stats
```

### Angular distance

Here we create a function to calculate the distance between two angles, taking into account that angles can cross zero.

```{r stats angdist}
# Function to calculate smallest angular distance
angular_distance <- function(a, b) {
  diff <- abs(a - b) %% 360
  return(min(diff, 360 - diff))
}
# Example
angular_distance(350, 10)
angular_distance(190, 0)

# Calculate distances
dist_left_right <- angular_distance(as.numeric(mean_left), as.numeric(mean_right))
dist_left_back <- angular_distance(as.numeric(mean_left), as.numeric(mean_back))
dist_right_back <- angular_distance(as.numeric(mean_right), as.numeric(mean_back))

print(dist_left_right)
print(dist_left_back)
print(dist_right_back)
```

### Visualization

```{r stats viz}
## Circular means
sides_means <- c(mean_left, mean_right, mean_back)
names(sides_means) <- c("Left", "Right", "Back")
d <- list()
for(a in 1:length(sides_means)){
  d[[a]] <- sapply(1:length(sides_means), function(x){
    angular_distance(sides_means[a], sides_means[x])
  })
}
sides_mat <- do.call(cbind, d)
colnames(sides_mat) <- rownames(sides_mat) <- names(sides_means)
sides_mat[lower.tri(sides_mat)] <- NA
sides_dists <- reshape2::melt(sides_mat)

# Plot burial orientations per country
sides_means_dt <- data.table(Sides = names(sides_means), Mean = sides_means)
sides_means_dt[, Mean := ifelse(Mean < 0, Mean + 360, Mean)]
ggplot(data = sides, aes(x = BurialSide, y = BurialOrientationMean, fill = BurialSide)) +
  geom_violin(width = 1, drop = F) +
  geom_point(data = sides_means_dt, aes(x = Sides, y = Mean), size = 2, shape = 23, fill = "white") +
  scale_fill_manual(values = pals::brewer.set2(3)) +
  ylab("Burial orientation") +
  theme_light() +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Plot pairwise angular distances
ggplot(data = sides_dists[!is.na(sides_dists$value),]) +
  geom_tile(aes(Var1, Var2, fill = value), color = "black") +
  scale_fill_gradientn("Angular distance", colours = rev(pals::brewer.rdylbu(100)), 
                       na.value = "transparent",
                       limits = c(0,180)) + # maximum distance between angles?
                       # guide = guide_colorbar(barwidth = 1, barheight = 15)) +
  geom_text(aes(Var1, Var2, label = round(value, 1)), size = 16, size.unit = "pt") +
  scale_x_discrete(position = "top") +
  ggtitle("Burial side") +
  coord_equal(expand = F) +
  theme_classic() +
  theme(plot.title = element_text(face = "bold", size = 18),
        plot.title.position = "plot",
        plot.margin = margin(l = 0, t = 5, b = 20, r = 30),
        axis.text.x = element_text(angle = 45, hjust = 0),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "none",
        axis.text = element_text(size = 14))
```

### P-value

Now we shuffle the values of angles for each category and compute the differences in circular means and do a histogram where we plot the differences in circular means against the true difference. Do it for each of the three pairs. Make sure that each group has the same amount of observations.

```{r stats pval}
n <- min(c(length(left_circ), length(right_circ), length(back_circ)))

diffs <- lapply(1:1000, function(x){
  l <- sample(left_circ, n)
  r <- sample(right_circ, n)
  b <- sample(back_circ, n)
  mean_l <- mean.circular(l)
  mean_r <- mean.circular(r)
  mean_b <- mean.circular(b)
  dist_lr <- angular_distance(as.numeric(mean_l), as.numeric(mean_r))
  dist_lb <- angular_distance(as.numeric(mean_l), as.numeric(mean_b))
  dist_br <- angular_distance(as.numeric(mean_b), as.numeric(mean_r))
  return(list(dist_lr, dist_lb, dist_br))
})

diffs_lr <- unlist(sapply(diffs, "[", 1))
diffs_lb <- unlist(sapply(diffs, "[", 2))
diffs_br <- unlist(sapply(diffs, "[", 3))
diffs <- list(diffs_lr, diffs_lb, diffs_br)

## Plot histograms
hist(diffs_lr, breaks = 100, main = "Difference between left and right orientations", xlab = "Angular distance"); abline(v = dist_left_right, col= "red", lwd = 3)

hist(diffs_lb, breaks = 100, main = "Difference between left and back orientations", xlab = "Angular distance"); abline(v = dist_left_back, col= "red", lwd = 3)

hist(diffs_br, breaks = 100, main = "Difference between back and right orientations", xlab = "Angular distance"); abline(v = dist_right_back, col= "red", lwd = 3)

## Calculate p-value
# P = (X + 1) / (N + 1)
obs <- list(dist_left_right, dist_left_back, dist_right_back)
pval <- lapply(1:length(diffs), function(x){
  X <- sum(abs(diffs[[x]]) >= abs(obs[[x]]))
  N <- length(diffs[[x]])
  return((X + 1) / (N + 1))
})
names(pval) <- c("left_right", "left_back", "back_right")
pval
```

### Similarity

We can also calculate a similarity index that takes into account the distance between the means of the different categories.

```{r stats similarity}
# Similarity index
similarity_index <- function(d) {
  return(1 - (d / 180))
}

sim_left_right <- similarity_index(dist_left_right)
sim_left_back <- similarity_index(dist_left_back)
sim_right_back <- similarity_index(dist_right_back)

print(sim_left_right)
print(sim_left_back)
print(sim_right_back)
```

### Watson-William test and circular ANOVA

```{r stats anova}
## Watson-Williams test
# Combine data
all_angles <- c(left_angles, right_angles, back_angles) #right_angles, 
group <- factor(c(rep("left", length(left_angles)),
                  rep("right", length(right_angles)),
                  rep("back", length(back_angles))))

## Mardia-Watson-Wheeler test (non-parametric)
mww_result1 <- watson.two.test(left_circ, right_circ)  # pairwise example
mww_result2 <- watson.two.test(left_circ, back_circ)
mww_result3 <- watson.two.test(right_circ, back_circ)
print(mww_result1)
print(mww_result2)
print(mww_result3)

## Watson-Williams test
watson.williams.test(circular(all_angles, units = "degrees", type = "angles", rotation = "clock"), group)

## Circular ANOVA 
aov.circular(circular(all_angles, units = "degrees", type = "angles", rotation = "clock"), group)
```

Now that we have tested different statistics using the left, right, and back angles, we can do the same for other categories, like periods, cultures, and countries.

## Country

```{r stats country, fig.dim=c(12,10)}
## Circular means
countries <- split(orientation[!(Country %in% c("Armenia", "Kazakhstan"))][, .(BurialOrientationMean, Country)], by = "Country")
countries_means <- lapply(countries, function(x){
  mean.circular(circular(x$BurialOrientationMean, units = "degrees", type = "angles", rotation = "clock"))
})
countries_means <- unlist(countries_means)
countries_means <- countries_means[gtools::mixedsort(names(countries_means))]

d <- list()
for(a in 1:length(countries_means)){
  d[[a]] <- sapply(1:length(countries_means), function(x){
    angular_distance(countries_means[a], countries_means[x])
  })
}
countries_mat <- do.call(cbind, d)
colnames(countries_mat) <- rownames(countries_mat) <- names(countries_means)
countries_mat[lower.tri(countries_mat)] <- NA
countries_dists <- setDT(reshape2::melt(countries_mat))

# Plot burial orientations per country
countries_means_dt <- data.table(Country = names(countries_means), Mean = countries_means)
countries_means_dt[, Mean := ifelse(Mean < 0, Mean + 360, Mean)]

ggplot(data = orientation[!(Country %in% c("Armenia", "Kazakhstan"))], aes(x = Country, y = BurialOrientationMean, fill = Country)) +
  geom_violin(width = 1.5, drop = F) +
  geom_point(data = countries_means_dt, aes(x = Country, y = Mean), size = 2, shape = 23, fill = "white") +
  scale_fill_manual(values = pals::brewer.set2(19)) +
  ylab("Burial orientation") +
  theme_light() +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Plot pairwise angular distances
countries_dists <- countries_dists[!is.na(countries_dists$value),]

ggplot(data = countries_dists) +
  geom_tile(aes(Var1, Var2, fill = value), color = "black") +
  scale_fill_gradientn("Angular distance", colours = rev(pals::brewer.rdylbu(100)), 
                       na.value = "transparent",
                       limits = c(0,180)) + # maximum distance between angles?
  geom_text(aes(as.factor(Var1), as.factor(Var2), label = round(value, 1)), size = 10, size.unit = "pt") +
  scale_x_discrete(position = "top") +
  scale_y_discrete(position = "left") +
  ggtitle("Country") +
  coord_equal(expand = F) +
  theme_classic() +
  theme(plot.title = element_text(face = "bold", size = 18),
        plot.title.position = "plot",
        axis.text.x = element_text(angle = 45, hjust = 0),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "none",
        axis.text = element_text(size = 14),
        plot.margin = margin(l = 30, t = 5, b = 20, r = 0))

```

## Period

```{r stats period, fig.dim=c(10, 8)}
period_order <- c("Mesolithic",
                  "Mesolithic/Neolithic",
                  "Neolithic",
                  "Eneolithic",
                  "Eneolithic/Bronze Age",
                  "Bronze Age",
                  "Bronze Age/Iron Age",
                  "Iron Age")
orientation$Period <- factor(orientation$Period,
                             levels = period_order)

## Circular means
period <- split(orientation[Period != "Iron Age"]
                [, .(BurialOrientationMean, Period)], 
                by = "Period")
period_means <- lapply(period, function(x){
  mean.circular(circular(x$BurialOrientationMean, 
                         units = "degrees", 
                         type = "angles", 
                         rotation = "clock"))
})
period_means <- unlist(period_means)
period_means <- period_means[period_order]
d <- list()
for(a in 1:length(period_means)){
  d[[a]] <- sapply(1:length(period_means), function(x){
    angular_distance(period_means[a], period_means[x])
  })
}
period_mat <- do.call(cbind, d)
colnames(period_mat) <- rownames(period_mat) <- names(period_means)
period_mat[lower.tri(period_mat)] <- NA
period_dists <- setDT(reshape2::melt(period_mat))

# Plot burial orientations per period
period_means_dt <- data.table(Period = names(period_means), Mean = period_means)
period_means_dt[, Mean := ifelse(Mean < 0, Mean + 360, Mean)]

ggplot(data = orientation[Period != "Iron Age"], aes(x = Period, y = BurialOrientationMean, fill = Period)) +
  geom_violin(width = 1, drop = F) +
  geom_point(data = period_means_dt[Period != "Iron Age"], aes(x = Period, y = Mean), size = 2, shape = 23, fill = "white") +
  scale_fill_manual(values = pals::brewer.set2(8)) +
  ylab("Burial orientation") +
  theme_light() +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Plot pairwise angular distances
ggplot(data = period_dists[!is.na(period_dists$value),]) +
  geom_tile(aes(Var1, Var2, fill = value), color = "black") +
  scale_fill_gradientn("Angular distance", 
                       colours = rev(pals::brewer.rdylbu(100)), 
                       na.value = "transparent",
                       limits = c(0, 180)) +
  geom_text(aes(Var1, Var2, label = round(value, 1)), 
            size = 10, size.unit = "pt") +
  scale_x_discrete(position = "top") +
  ggtitle("Period") +
  coord_equal(expand = F) +
  theme_classic() +
  theme(plot.title = element_text(face = "bold", size = 18),
        plot.title.position = "plot",
        plot.margin = margin(l = 0, t = 20, b = 0, r = 30),
        axis.text.x = element_text(angle = 45, hjust = 0),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "none",
        axis.text = element_text(size = 14))

```

## Culture

```{r stats culture, fig.dim=c(15,10)}
# Plot only the 10 most frequent cultures
culture_keep <- orientation[!is.na(Culture)][, .N, by = "Culture"][order(N, decreasing = T)][, Culture][1:10]

## Circular means
culture <- orientation[Culture %in% culture_keep][, .(BurialOrientationMean, Culture)]
culture$Culture <- factor(culture$Culture, 
                               labels = c("Linearbandkeramik",
                                          "Lažňany",
                                          "Sopot",
                                          "Bell Beaker",
                                          "Únětice",
                                          "Corded Ware",
                                          "Baden",
                                          "Dnieper-Donets",
                                          "Trichterbecher",
                                          "Baltic Meso-Neolithic"),
                               levels = unique(culture$Culture))

culture <- split(culture, by = "Culture")
culture_means <- lapply(culture, function(x){
  mean.circular(circular(x$BurialOrientationMean, 
                         units = "degrees", 
                         type = "angles", 
                         rotation = "clock"),
                na.rm = T)
})
culture_means <- unlist(culture_means)
d <- list()
for(a in 1:length(culture_means)){
  d[[a]] <- sapply(1:length(culture_means), function(x){
    angular_distance(culture_means[a], culture_means[x])
  })
}
culture_mat <- do.call(cbind, d)
colnames(culture_mat) <- rownames(culture_mat) <- names(culture_means)
culture_mat[lower.tri(culture_mat)] <- NA
culture_dists <- setDT(reshape2::melt(culture_mat))

# Plot burial orientations per period
culture_means_dt <- data.table(Culture = names(culture_means), 
                               Mean = culture_means)
culture_means_dt[, Mean := ifelse(Mean < 0, Mean + 360, Mean)]
culture_means_dt
ggplot(data = orientation[Culture %in% culture_keep], 
       aes(x = Culture, y = BurialOrientationMean, 
           fill = Culture)) +
  geom_violin(width = 1, drop = F) +
  geom_point(data = culture_means_dt, aes(x = Culture, y = Mean), 
             size = 2, shape = 23, fill = "white") +
  scale_fill_manual(values = pals::brewer.set2(10)) +
  ylab("Burial orientation") +
  theme_light() +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Plot pairwise angular distances
ggplot(data = culture_dists[!is.na(culture_dists$value),]) +
  geom_tile(aes(Var1, Var2, fill = value), color = "black") +
  scale_fill_gradientn("Angular distance", 
                       colours = rev(pals::brewer.rdylbu(100)), 
                       na.value = "transparent",
                       limits = c(0, 180)) +
  geom_text(aes(Var1, Var2, label = round(value, 1)), 
            size = 10, size.unit = "pt") +
  scale_x_discrete(position = "top") +
  ggtitle("Culture") +
  coord_equal(expand = F) +
  theme_classic() +
  theme(plot.title = element_text(face = "bold", size = 18),
        plot.title.position = "plot",
        plot.margin = margin(l = 50, t = 20, b = 0, r = 0),
        axis.text.x = element_text(angle = 45, hjust = 0),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "none",
        axis.text = element_text(size = 14))

```

### Difference between Corded Ware and Bell Beaker cultures

Here we are going to statistically test for differences in the orientations of Corded Ware and Bell Beaker cultures.

First of all, we look at the summary statistics for the two cultures. We can see that the means (mu) are quite different but the Corded Ware orientations are spread across the whole circle (there is very low concentration). This can be seen both by the rho value (mean resultant length) that is close to 0, and by the variance (var) value that is close to 1 ($var = 1 - rho$). The standard deviation runs from 0 to infinity, where higher values indicate a larger dispersion.

We can now try and check for differences in mean direction and concentration compared to a uniform distribution.

```{r stats cw-bb}
## Orientations
cordedware <- culture[["Corded Ware/Single Grave/Battle Axe/Fatyanovo"]]$BurialOrientationMean
cordedware <- circular(cordedware,
                      type = "angles",
                      units = "degrees",
                      rotation = "clock")

bellbeaker <- culture[["Bell Beaker complex"]]$BurialOrientationMean
bellbeaker <- circular(bellbeaker,
                      type = "angles",
                      units = "degrees",
                      rotation = "clock")

## Summary statistics of the two cultures
summary.dat <- data.frame(culture = c("Corded Ware", "Bell Beaker"),
                          mu = c(mean.circular(cordedware),
                                 mean.circular(bellbeaker)),
                          rho = c(rho.circular(cordedware),
                                  rho.circular(bellbeaker)),
                          var = c(var.circular(cordedware),
                                  var.circular(bellbeaker)),
                          sd = c(sd.circular(cordedware),
                                 sd.circular(bellbeaker)))
summary.dat

## Calculate mean direction and concentration from uniform distribution
cw_unif <- lapply(1:1000, function(x){
  r <- rcircularuniform(n = length(cordedware),
                            control.circular = list(
                              units = "degrees",
                              rotation = "clock"))
  m <- mean.circular(r)
  k <- rho.circular(r)
  return(list(m, k))
})
  
bb_unif <- lapply(1:1000, function(x){
  r <- rcircularuniform(n = length(bellbeaker),
                            control.circular = list(
                              units = "degrees",
                              rotation = "clock"))
  m <- mean.circular(r)
  k <- rho.circular(r)
  return(list(m, k))
})

cw_means <- sapply(cw_unif, "[", 1)
cw_rho <- sapply(cw_unif, "[", 2)

bb_means <- sapply(bb_unif, "[", 1)
bb_rho <- sapply(bb_unif, "[", 2)

hist(as.numeric(cw_means), breaks = 100, main = "Corded Ware", xlab = "Mean direction (mu)"); abline(v = mean.circular(cordedware), col = "red")
hist(as.numeric(cw_rho), breaks = 100, main = "Corded Ware", xlab = "Concentration (rho)", xlim = c(0,1)); abline(v = rho.circular(cordedware), col = "red")

hist(as.numeric(bb_means), breaks = 100, main = "Bell Beaker", xlab = "Mean direction (mu)"); abline(v = mean.circular(bellbeaker), col = "red")
hist(as.numeric(bb_rho), breaks = 100, main = "Bell Beaker", xlab = "Concentration (rho)", xlim = c(0,1)); abline(v = rho.circular(bellbeaker), col = "red")

# Similarity index (without folding over the horizontal axis)
obs <- culture_dists[Var2 == "Corded Ware/Single Grave/Battle Axe/Fatyanovo" & Var1 == "Bell Beaker complex"]$value
diff <- similarity_index(obs)
diff
paste0((1-round(diff,2))*100, "% different")
```

Now for modelling the differences between the two cultures, we can try using the `bpnreg` R package. This package performs a Bayesian circular regression model based on the projected normal distribution. The angles must be in radians. This is equivalent to an ANOVA.

In this model, the Bell Beaker is the reference culture. Thus, the value of the intercept for the categorical variables represents the mean of the Bell Beaker orientation.

Further reading:

-   <https://www.frontiersin.org/journals/psychology/articles/10.3389/fpsyg.2018.02040/full>

-   <https://bpspsychub.onlinelibrary.wiley.com/doi/full/10.1111/bmsp.12108>

-   <https://dspace.library.uu.nl/bitstream/handle/1874/361637/1_s2.0_S0022249616300827_main.pdf?sequence=1>

```{r stats cw-bb bpnreg, results='hide'}
# Data for regression
circ.dat <- data.frame(Angles = c(as.numeric(cordedware)*pi/180,
                                  as.numeric(bellbeaker)*pi/180),
                       Culture = c(rep("Corded Ware", length(cordedware)),
                                   rep("Bell Beaker", length(bellbeaker))))
circ.dat$Culture <- as.factor(circ.dat$Culture)

fit <- bpnr(Angles ~ 1 + Culture, data = circ.dat, 
            its = 10000, burn = 100, n.lag = 3, seed = 1234)
```

```{r stats cw-bb bpnreg2}
# Means
aov.dt <- setDT(data.frame(coef_circ(fit, "categorical", "radians")$Means))
aov.dt[, Culture := as.factor(c("Bell Beaker", "Corded Ware"))]

# Plot
plot.circular(c(aov.dt$LB[1], aov.dt$UB[1]), pch = 16, 
              rotation = "clock", zero = pi/2)
points.circular(c(aov.dt$LB[2], aov.dt$UB[2]), pch = 5, 
                rotation = "clock", zero = pi/2)
points.circular(aov.dt$mean[1], pch = 16, col = "red", 
                rotation = "clock", zero = pi/2)
points.circular(aov.dt$mean[2], pch = 18, col = "blue", 
                rotation = "clock", zero = pi/2)

aov.dt[, .(Culture, deg(.SD)), .SDcols = c("mean", "mode", "sd", "LB", "UB")]

# Differences
diff.dt <- setDT(data.frame(coef_circ(fit, "categorical", units = "degrees")$Differences))
diff.dt[, Difference := "Bell Beaker/Corded Ware"]
diff.dt[, .(Difference, mean, mode, sd, LB, UB)]

# Calculate angular distance between means
angular_distance(aov.dt$mean[1]*180/pi, aov.dt$mean[2]*180/pi)
```

In theory the HPDs of the two cultures overlap so the difference in means shouldn't be significant. When plotting the results, we can see that the two means are quite different and the HPDs overlap only because the spread of the Corded Ware orientations is extremely large. If we look at the differences table, the LB and UP of the HPD for the difference doesn't cross zero, so the difference between the two cultures is significant.

We can also try these other two tests that according to this paper seem to perform well with circular data, without falling too much into Type-1 and Type-2 errors.

<https://doi.org/10.1038/s41598-021-99299-5>

```{r stats watson.two}
## Watson's U2 test
watson.two(cordedware, bellbeaker)
```

So when the Watson's U^2^ test has a small p-value or a large test statistic, it means that the two groups are **NOT** homogeneious, so they are significantly different.

### Difference between Lažňany and Únětice cultures

Now let's double check that this interpretation is correct by comparing two other very similar cultures, e.g., Lažňany and Únětice. Únětice has a larger spread than Lažňany, but the means have a similar direction.

```{r stats lz-un bpnreg}
laznany <- circular(culture[["Lažňany"]]$BurialOrientationMean,
                    type = "angles",
                    units = "degrees",
                    rotation = "clock")
unetice <- circular(culture[["Únětice"]]$BurialOrientationMean,
                    type = "angles",
                    units = "degrees",
                    rotation = "clock")

summary.dat2 <- data.frame(culture = c("Lažňany", "Únětice"),
                          mu = c(mean.circular(laznany),
                                 mean.circular(unetice)),
                          rho = c(rho.circular(laznany),
                                  rho.circular(unetice)),
                          var = c(var.circular(laznany),
                                  var.circular(unetice)),
                          sd = c(sd.circular(laznany),
                                 sd.circular(unetice)))
summary.dat2
```

```{r stats lz-un bpnreg2, results = 'hide'}
# Data for regression
circ.dat2 <- data.frame(Angles = c(as.numeric(laznany)*pi/180,
                                  as.numeric(unetice)*pi/180),
                       Culture = c(rep("Lažňany", length(laznany)),
                                   rep("Únětice", length(unetice))))
circ.dat2$Culture <- as.factor(circ.dat2$Culture)

fit2 <- bpnr(Angles ~ 1 + Culture, data = circ.dat2, 
            its = 10000, burn = 100, n.lag = 3, seed = 1234)
```

```{r stats lz-un bpnreg3}
# Means
aov.dt2 <- setDT(data.frame(coef_circ(fit2, "categorical", "radians")$Means))
aov.dt2[, Culture := as.factor(c("Lažňany", "Únětice"))]

# Plot
plot.circular(c(aov.dt2$LB[1], aov.dt2$UB[1]), pch = 16, 
              rotation = "clock", zero = pi/2)
points.circular(c(aov.dt2$LB[2], aov.dt2$UB[2]), pch = 5, 
                rotation = "clock", zero = pi/2)
points.circular(aov.dt2$mean[1], pch = 16, col = "red", 
                rotation = "clock", zero = pi/2)
points.circular(aov.dt2$mean[2], pch = 18, col = "blue", 
                rotation = "clock", zero = pi/2)

aov.dt2[, .(Culture, deg(.SD)), .SDcols = c("mean", "mode", "sd", "LB", "UB")]

# Differences
diff.dt2 <- setDT(data.frame(coef_circ(fit2, "categorical", units = "degrees")$Differences))
diff.dt2[, Difference := "Lažňany/Únětice"]
diff.dt2[, .(Difference, mean, mode, sd, LB, UB)]

# Calculate distance between means
angular_distance(aov.dt2$mean[1]*180/pi, aov.dt2$mean[2]*180/pi)

## Watson's U2 test
watson.two(laznany, unetice)
```

In this case we can see that the HPDs for the two cultures completely overlap and their means differ by only \~3 degrees. Furthermore, the LB and UB of the HPD for the difference crosses 0. Therefore, they are not significantly different from one another.

The Watson's U^2^ test also confirms this interpretation, as the p-value is greater than 0.10 and the test statistic is small.

### Difference across all cultures

For testing difference across all cultures (groups) we have to rotate the reference culture against which the regression is made. This is because the Intercept in these models is actually the reference culture, so each posterior summary statistic has to be interpreted against the group of reference.

Therefore, to look at the differences table we will have to fit a different model where the reference culture rotates.

But first, we can look at the means table which will always stay the same.

```{r stats cultures all, echo=TRUE, eval=T, results = "hide"}
culture <- split(orientation[Culture %in% culture_keep]
                 [, .(BurialOrientationMean, Culture, Longitude, Latitude, Age)], 
                 by = "Culture")
culture_means <- lapply(culture, function(x){
  mean.circular(circular(x$BurialOrientationMean, 
                         units = "degrees", 
                         type = "angles", 
                         rotation = "clock"),
                na.rm = T)
})
circ.dat.all <- lapply(culture, function(x){
  x$BurialOrientationMean <- x$BurialOrientationMean*pi/180
  return(x)
})
circ.dat.all <- rbindlist(circ.dat.all)
circ.dat.all$Culture <- factor(circ.dat.all$Culture, 
                               labels = c("Linearbandkeramik",
                                          "Lažňany",
                                          "Sopot",
                                          "Bell Beaker",
                                          "Únětice",
                                          "Corded Ware",
                                          "Baden",
                                          "Dnieper-Donets",
                                          "Trichterbecher",
                                          "Baltic Meso-Neolithic"),
                               levels = unique(circ.dat.all$Culture))

# Remove individual without Age
circ.dat.all <- circ.dat.all[!is.na(Age)]

# Model
fit.all <- bpnr(BurialOrientationMean ~ 1 + Culture, 
                data = circ.dat.all, its = 10000, burn = 1000, 
                n.lag = 3, seed = 1234)

# Model fit
fit(fit.all)
```

The table shows the HPD of the means for each group. If the HPDs overlap across groups, the difference between those groups is not significant.

```{r stats cultures all2}
# Means
aov.dt.all <- setDT(data.frame(coef_circ(fit.all, "categorical", "radians")$Means))[1:10,]
aov.dt.all[, Culture := unique(circ.dat.all$Culture)]
# c1 <- sapply(strsplit(aov.dt.all$Culture, "Culture"), "[", 2)
# c2 <- sapply(strsplit(aov.dt.all$Culture, "Culture"), "[", 3)
# c1[is.na(c1)] <- c2[is.na(c2)] <- "Linearbandkeramik"
# aov.dt.all[, Culture := paste0(c1, "/", c2)]
aov.dt.all[, .(Culture, deg(.SD)), .SDcols = c("mean", "mode", "sd", "LB", "UB")]
```

```{r stats cultures all3, echo = FALSE, fig.dim=c(15,15)}
## Visualization with circlize package
culture.means <- deg(aov.dt.all$mean)
culture.means <- ifelse(culture.means < 0, 360 + culture.means, culture.means)
culture.ranges.lb <- deg(aov.dt.all$LB)
culture.ranges.lb <- ifelse(culture.ranges.lb < 0, 360 + culture.ranges.lb, culture.ranges.lb)
culture.ranges.ub <- deg(aov.dt.all$UB)
culture.ranges.ub <- ifelse(culture.ranges.ub < 0, 360 + culture.ranges.ub, culture.ranges.ub)

# params
labs <- as.character(seq(0, 360, 22.5))
labs[17] <- "0"
cols <- rev(pals::ocean.thermal(12))[2:11]
ys <- seq(0, 1, length.out = 10)
texts <- levels(aov.dt.all$Culture)

# plot
circos.par(gap.degree = 0, cell.padding = c(0, 0, 0, 0), start.degree = 90)
circos.initialize("a", xlim = c(0, 360), ring = T)
circos.track(ylim = c(0.75, 1.05), bg.border = NA)
circos.axis(major.at = seq(0, 360, 45), 
            direction = "outside",
            labels.facing = "inside",
            labels.niceFacing = T,
            labels = labs,
            labels.cex = 2)
circos.segments(x0 = culture.ranges.lb[1], 
                y0 = ys[1], 
                x1 = culture.ranges.ub[1], 
                y1 = ys[1], col = cols[1], lwd = 2)
circos.text(x = culture.means[1], 
            y = ys[1] - 0.05,
            labels = texts[1], 
            col = cols[1],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[2], 
                y0 = ys[2], 
                x1 = culture.ranges.ub[2], 
                y1 = ys[2], col = cols[2], 
                lwd = 2)
circos.text(x = culture.means[2], 
            y = ys[2] - 0.05,
            labels = texts[2], 
            col = cols[2],
            facing = "bending.outside",
            cex = 1)
# crosses zero
circos.segments(x0 = culture.ranges.lb[3], 
                y0 = ys[3], x1 = 360, 
                y1 = ys[3], col = cols[3], 
                lwd = 2)
circos.segments(x0 = 0, y0 = ys[3], 
                x1 = culture.ranges.ub[3], 
                y1 = ys[3], col = cols[3], 
                lwd = 2)
circos.text(x = culture.means[3], 
            y = ys[3] - 0.05,
            labels = texts[3], 
            col = cols[3],
            facing = "bending.inside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[4], 
                y0 = ys[4], 
                x1 = culture.ranges.ub[4], 
                y1 = ys[4], col = cols[4], 
                lwd = 2)
circos.text(x = culture.means[4], 
            y = ys[4] - 0.05,
            labels = texts[4], 
            col = cols[4],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[5], 
                y0 = ys[5], 
                x1 = culture.ranges.ub[5], 
                y1 = ys[5], col = cols[5], 
                lwd = 2)
circos.text(x = culture.means[5], 
            y = ys[5] - 0.05,
            labels = texts[5], 
            col = cols[5],
            facing = "bending.outside",
            cex = 1)
# crosses zero
circos.segments(x0 = culture.ranges.lb[6], 
                y0 = ys[6], x1 = 360, 
                y1 = ys[6], col = cols[6], 
                lwd = 2)
circos.segments(x0 = 0, y0 = ys[6], 
                x1 = culture.ranges.ub[6], 
                y1 = ys[6], col = cols[6], 
                lwd = 2)
circos.text(x = culture.means[6], 
            y = ys[6] - 0.05,
            labels = texts[6], 
            col = cols[6],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[7], 
                y0 = ys[7], 
                x1 = culture.ranges.ub[7], 
                y1 = ys[7], col = cols[7], 
                lwd = 2)
circos.text(x = culture.means[7], 
            y = ys[7] - 0.05,
            labels = texts[7], 
            col = cols[7],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[8], 
                y0 = ys[8], 
                x1 = culture.ranges.ub[8], 
                y1 = ys[8], col = cols[8], 
                lwd = 2)
circos.text(x = culture.means[8], 
            y = ys[8] - 0.05,
            labels = texts[8], 
            col = cols[8],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[9], 
                y0 = ys[9], 
                x1 = culture.ranges.ub[9], 
                y1 = ys[9], col = cols[9], 
                lwd = 2)
circos.text(x = culture.means[9], 
            y = ys[9] - 0.05,
            labels = texts[9], 
            col = cols[9],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[10], 
                y0 = ys[10], 
                x1 = culture.ranges.ub[10], 
                y1 = ys[10], col = cols[10], 
                lwd = 2)
circos.text(x = culture.means[10], 
            y = ys[10] - 0.05,
            labels = texts[10], 
            col = cols[10],
            facing = "bending.outside",
            cex = 1)
circos.points(x = culture.means, y = ys, bg = cols, pch = 21)
circos.clear()
```

The plot shows the posterior mean direction and HPD range of each culture. If the mean is outside the ranges of another culture, the difference between those cultures is significant.

We will test this statistically by fitting models with rotating group of reference.

```{r stats differences, results = "hide"}
# Rotate values so that reference culture is always a new one
culture.labels <- as.character(unique(aov.dt.all$Culture))
culture.labels <- lapply(0:(length(culture.labels)-1), function(i) {
  c(culture.labels[(i+1):length(culture.labels)], culture.labels[1:i])
})
culture.labels[[1]] <- culture.labels[[1]][1:10]

# Regression model list
differences.all <- lapply(culture.labels, function(l){
  # data
  circ.dat.all$Culture <- as.character(circ.dat.all$Culture)
  circ.dat.all$Culture <- factor(circ.dat.all$Culture, 
                                 levels = l)
  # model
  fit.m <- bpnr(BurialOrientationMean ~ 1 + Culture, 
                data = circ.dat.all, its = 10000, burn = 1000, 
                n.lag = 3, seed = 1234)
  # differences
  diff.dt <- setDT(data.frame(coef_circ(fit.m, "categorical", units = "degrees")$Differences)[1:9,])
  diff.dt[, Comparison := rownames(coef_circ(fit.m, "categorical", units = "degrees")$Differences)[1:9]]
  c1 <- sapply(strsplit(diff.dt$Comparison, "Culture"), "[", 2)
  c2 <- sapply(strsplit(diff.dt$Comparison, "Culture"), "[", 3)
  c2[is.na(c2)] <- l[1]
  diff.dt[, Comparison := paste0(c2, "/", c1)]
  diff.dt[, .(Comparison, mean, mode, sd, LB, UB)]
  diff.dt[, Overlapping := ifelse(between(0, LB, UB) | LB > UB, "No", "Yes")]
  return(diff.dt)
})
```

```{r stats differences2}
# Remove duplicated comparisons
dt <- rbindlist(differences.all)
dt$Comparison <- sapply(strsplit(dt$Comparison, "/"), function(pair) {
  paste(gtools::mixedsort(pair), collapse = "/")
})
dt <- dt[!duplicated(Comparison)][, .(Comparison, mean, mode, sd, LB, UB, Overlapping)]
dt[, Culture1 := sapply(strsplit(dt$Comparison, "/"), "[", 1)]
dt[, Culture2 := sapply(strsplit(dt$Comparison, "/"), "[", 2)]
dt <- dt[, .(Comparison, Culture1, Culture2, mean, mode, sd, LB, UB, Overlapping)][order(Culture1)]

datatable(dt[order(Culture1)][, .(Culture1, Culture2, mean, mode, sd, LB, UB, Overlapping, Comparison)], options = list(scrollX = TRUE))
```

#### Model with culture, space, and time

Here we are going to add latitude, longitude and age of the burial to the model. We want to test whether culture is indeed important for explaining patterns in burial orientation compared to just the location of the graves in space and time.

```{r stats cultures spatemp, echo=TRUE, eval=T, results = "hide"}
# Model
fit.sptemp <- bpnr(BurialOrientationMean ~ 1 + Culture + Longitude + Latitude + Age, 
                data = circ.dat.all, its = 10000, burn = 1000, 
                n.lag = 3, seed = 1234)
```
```{r stats cultures spatemp2, echo=TRUE, eval=T}
coef_circ(fit.sptemp, type = "continuous")
coef_circ(fit.sptemp, type = "categorical")$Means[1:10,]

# Model fit
fit(fit.sptemp)
```
```{r stats cultures spatemp3, echo = FALSE, fig.dim=c(15,15), eval = F}
## Visualization with circlize package
culture.means <- coef_circ(fit.sptemp, "categorical", "degrees")$Means[1:10,1]
culture.means <- ifelse(culture.means < 0, 360 + culture.means, culture.means)
culture.ranges.lb <- coef_circ(fit.sptemp, "categorical", "degrees")$Means[1:10,4]
culture.ranges.lb <- ifelse(culture.ranges.lb < 0, 360 + culture.ranges.lb, culture.ranges.lb)
culture.ranges.ub <- coef_circ(fit.sptemp, "categorical", "degrees")$Means[1:10,5]
culture.ranges.ub <- ifelse(culture.ranges.ub < 0, 360 + culture.ranges.ub, culture.ranges.ub)

# params
labs <- as.character(seq(0, 360, 22.5))
labs[17] <- "0"
cols <- rev(pals::ocean.thermal(12))[2:11]
ys <- seq(0, 1, length.out = 10)
texts <- levels(aov.dt.all$Culture)

# plot
circos.par(gap.degree = 0, cell.padding = c(0, 0, 0, 0), start.degree = 90)
circos.initialize("a", xlim = c(0, 360), ring = T)
circos.track(ylim = c(0,1), bg.border = NA)
circos.axis(major.at = seq(0, 360, 22.5), 
            direction = "outside",
            labels = labs)
circos.segments(x0 = culture.ranges.lb[1], 
                y0 = ys[1], 
                x1 = culture.ranges.ub[1], 
                y1 = ys[1], col = cols[1], lwd = 2)
circos.text(x = culture.means[1], 
            y = ys[1] - 0.05,
            labels = texts[1], 
            col = cols[1],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[2], 
                y0 = ys[2], 
                x1 = culture.ranges.ub[2], 
                y1 = ys[2], col = cols[2], 
                lwd = 2)
circos.text(x = culture.means[2], 
            y = ys[2] - 0.05,
            labels = texts[2], 
            col = cols[2],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[3], 
                y0 = ys[3], x1 = culture.ranges.ub[3], 
                y1 = ys[3], col = cols[3], 
                lwd = 2)
circos.text(x = culture.means[3], 
            y = ys[3] - 0.05,
            labels = texts[3], 
            col = cols[3],
            facing = "bending.inside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[4], 
                y0 = ys[4], 
                x1 = culture.ranges.ub[4], 
                y1 = ys[4], col = cols[4], 
                lwd = 2)
circos.text(x = culture.means[4], 
            y = ys[4] - 0.05,
            labels = texts[4], 
            col = cols[4],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[5], 
                y0 = ys[5], 
                x1 = culture.ranges.ub[5], 
                y1 = ys[5], col = cols[5], 
                lwd = 2)
circos.text(x = culture.means[5], 
            y = ys[5] - 0.05,
            labels = texts[5], 
            col = cols[5],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[6], 
                y0 = ys[6], x1 = culture.ranges.ub[6], 
                y1 = ys[6], col = cols[6], 
                lwd = 2)
circos.text(x = culture.means[6], 
            y = ys[6] - 0.05,
            labels = texts[6], 
            col = cols[6],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[7], 
                y0 = ys[7], 
                x1 = culture.ranges.ub[7], 
                y1 = ys[7], col = cols[7], 
                lwd = 2)
circos.text(x = culture.means[7], 
            y = ys[7] - 0.05,
            labels = texts[7], 
            col = cols[7],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[8], 
                y0 = ys[8], 
                x1 = culture.ranges.ub[8], 
                y1 = ys[8], col = cols[8], 
                lwd = 2)
circos.text(x = culture.means[8], 
            y = ys[8] - 0.05,
            labels = texts[8], 
            col = cols[8],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[9], 
                y0 = ys[9], 
                x1 = culture.ranges.ub[9], 
                y1 = ys[9], col = cols[9], 
                lwd = 2)
circos.text(x = culture.means[9], 
            y = ys[9] - 0.05,
            labels = texts[9], 
            col = cols[9],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[10], 
                y0 = ys[10], 
                x1 = culture.ranges.ub[10], 
                y1 = ys[10], col = cols[10], 
                lwd = 2)
circos.text(x = culture.means[10], 
            y = ys[10] - 0.05,
            labels = texts[10], 
            col = cols[10],
            facing = "bending.outside",
            cex = 1)
circos.points(x = culture.means, y = ys, bg = cols, pch = 21)
circos.clear()
```

In this case we can see that DIC and WAIC scores for the full model are smaller than the only culture one, meaning that culture added to space and time gives a better fit. 

#### Model with just space and time

Now we repeat by including only space and time, like in a baseline model.

```{r stats spatemp, echo=TRUE, eval=T, results = "hide"}
# Model
fit.sptemp.only <- bpnr(BurialOrientationMean ~ 1 + Longitude + Latitude + Age + Longitude:Latitude, 
                data = circ.dat.all, its = 10000, burn = 1000, 
                n.lag = 3, seed = 1234)
```
```{r stats spatemp2, echo=TRUE, eval=T}
coef_circ(fit.sptemp.only, type = "continuous")

# Model fit
fit(fit.sptemp.only)
```

Based on the DIC and WAIC scores we can see that this model has the worst model fit across all three, with the full model being the most informative. This means that space and time are not enough to explain the patterns in burial orientation and therefore culture is a good predictor.

#### Comparing models fit
```{r stats compare fit}
# Culture only model
fit(fit.all)

# Culture, space and time
fit(fit.sptemp)

# Space and time only
fit(fit.sptemp.only)
```

### Differences across cultures with orientations divided by body position

Here we are going to test for differences in cultures when dividing by body position. This means that for each position (left, right, back), we model the orientations against cultures.

#### Left

```{r stats cultures left, echo=TRUE, eval=T, results = "hide"}
# Add culture information to left data
left <- merge(left, orientation[, .(Num, Culture, Longitude, Latitude, Age)], by = "Num", all.x = T)
nrow(left)
nrow(left[!is.na(Culture)][Culture %in% culture_keep])

# Remove NAs in Culture and keep only 10 most common cultures
culture_left <- left[!is.na(Culture)][!is.na(Age)][Culture %in% culture_keep][, .(BurialOrientationMean, Culture, Longitude, Latitude, Age)]

# Transform to radians
circ.dat.left <- culture_left
circ.dat.left[, Angles := BurialOrientationMean*pi/180]
circ.dat.left$Culture <- factor(circ.dat.left$Culture, 
                               labels = c("Linearbandkeramik",
                                          "Lažňany",
                                          "Sopot",
                                          "Bell Beaker",
                                          "Corded Ware",
                                          "Baden",
                                          "Dnieper-Donets",
                                          "Únětice",
                                          "Trichterbecher",
                                          "Baltic Meso-Neolithic"),
                               levels = unique(circ.dat.left$Culture))

# Model
fit.left <- bpnr(Angles ~ 1 + Culture, 
                data = circ.dat.left, its = 10000, burn = 1000, 
                n.lag = 3, seed = 1234)
```
```{r stats cultures left2}
# Means
aov.dt.left <- setDT(data.frame(coef_circ(fit.left, "categorical", "radians")$Means))[1:10,]
aov.dt.left[, Culture := unique(circ.dat.left$Culture)]
# c1 <- sapply(strsplit(aov.dt.all$Culture, "Culture"), "[", 2)
# c2 <- sapply(strsplit(aov.dt.all$Culture, "Culture"), "[", 3)
# c1[is.na(c1)] <- c2[is.na(c2)] <- "Linearbandkeramik"
# aov.dt.all[, Culture := paste0(c1, "/", c2)]
aov.dt.left[, .(Culture, deg(.SD)), .SDcols = c("mean", "mode", "sd", "LB", "UB")]
```
```{r stats differences left, results = "hide"}
# Rotate values so that reference culture is always a new one
culture.labels <- as.character(unique(aov.dt.left$Culture))
culture.labels <- lapply(0:(length(culture.labels)-1), function(i) {
  c(culture.labels[(i+1):length(culture.labels)], culture.labels[1:i])
})
culture.labels[[1]] <- culture.labels[[1]][1:10]

# Regression model list
differences.left <- lapply(culture.labels, function(l){
  # data
  circ.dat.left$Culture <- as.character(circ.dat.left$Culture)
  circ.dat.left$Culture <- factor(circ.dat.left$Culture, 
                                 levels = l)
  # model
  fit.m <- bpnr(Angles ~ 1 + Culture, 
                data = circ.dat.left, its = 10000, burn = 1000, 
                n.lag = 3, seed = 1234)
  # differences
  diff.dt.left <- setDT(data.frame(coef_circ(fit.m, "categorical", units = "degrees")$Differences)[1:9,])
  diff.dt.left[, Difference := rownames(coef_circ(fit.m, "categorical", units = "degrees")$Differences)[1:9]]
  c1 <- sapply(strsplit(diff.dt.left$Difference, "Culture"), "[", 2)
  c2 <- sapply(strsplit(diff.dt.left$Difference, "Culture"), "[", 3)
  c2[is.na(c2)] <- l[1]
  diff.dt.left[, Difference := paste0(c2, "/", c1)]
  diff.dt.left[, .(Difference, mean, mode, sd, LB, UB)]
  diff.dt.left[, Significant := ifelse(between(0, LB, UB) | LB > UB, "No", "Yes")]
  return(diff.dt.left)
})
```
```{r stats differences left2}
# Remove duplicated comparisons
dt <- rbindlist(differences.left)
dt$Difference <- sapply(strsplit(dt$Difference, "/"), function(pair) {
  paste(gtools::mixedsort(pair), collapse = "/")
})
dt <- dt[!duplicated(Difference)][, .(Difference, mean, mode, sd, LB, UB, Significant)]
dt[, Culture1 := sapply(strsplit(dt$Difference, "/"), "[", 1)]
dt[, Culture2 := sapply(strsplit(dt$Difference, "/"), "[", 2)]
dt <- dt[, .(Difference, Culture1, Culture2, mean, mode, sd, LB, UB, Significant)][order(Culture1)]

datatable(dt[order(Culture1)][, .(Culture1, Culture2, mean, mode, sd, LB, UB, Significant, Difference)], options = list(scrollX = TRUE))
```
```{r stats cultures left3, echo = FALSE, fig.dim=c(15,15)}
## Visualization with circlize package
culture.means <- deg(aov.dt.left$mean)
culture.means <- ifelse(culture.means < 0, 360 + culture.means, culture.means)
culture.ranges.lb <- deg(aov.dt.left$LB)
culture.ranges.lb <- ifelse(culture.ranges.lb < 0, 360 + culture.ranges.lb, culture.ranges.lb)
culture.ranges.ub <- deg(aov.dt.left$UB)
culture.ranges.ub <- ifelse(culture.ranges.ub < 0, 360 + culture.ranges.ub, culture.ranges.ub)

# params
labs <- as.character(seq(0, 360, 22.5))
labs[17] <- "0"
cols <- rev(pals::ocean.thermal(12))[2:11]
ys <- seq(0, 1, length.out = 10)
texts <- levels(aov.dt.left$Culture)

# plot
circos.par(gap.degree = 0, cell.padding = c(0, 0, 0, 0), start.degree = 90)
circos.initialize("a", xlim = c(0, 360), ring = T)
circos.track(ylim = c(0,1), bg.border = NA)
circos.axis(major.at = seq(0, 360, 22.5), 
            direction = "outside",
            labels = labs)
circos.segments(x0 = culture.ranges.lb[1], 
                y0 = ys[1], 
                x1 = culture.ranges.ub[1], 
                y1 = ys[1], col = cols[1], lwd = 2)
circos.text(x = culture.means[1], 
            y = ys[1] - 0.05,
            labels = texts[1], 
            col = cols[1],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[2], 
                y0 = ys[2], 
                x1 = culture.ranges.ub[2], 
                y1 = ys[2], col = cols[2], 
                lwd = 2)
circos.text(x = culture.means[2], 
            y = ys[2] - 0.05,
            labels = texts[2], 
            col = cols[2],
            facing = "bending.outside",
            cex = 1)
# crosses zero
circos.segments(x0 = culture.ranges.lb[3], 
                y0 = ys[3], x1 = 360, 
                y1 = ys[3], col = cols[3], 
                lwd = 2)
circos.segments(x0 = 0, y0 = ys[3], 
                x1 = culture.ranges.ub[3], 
                y1 = ys[3], col = cols[3], 
                lwd = 2)
circos.text(x = culture.means[3], 
            y = ys[3] - 0.05,
            labels = texts[3], 
            col = cols[3],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[4], 
                y0 = ys[4], 
                x1 = culture.ranges.ub[4], 
                y1 = ys[4], col = cols[4], 
                lwd = 2)
circos.text(x = culture.means[4], 
            y = ys[4] - 0.05,
            labels = texts[4], 
            col = cols[4],
            facing = "bending.inside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[5], 
                y0 = ys[5], 
                x1 = culture.ranges.ub[5], 
                y1 = ys[5], col = cols[5], 
                lwd = 2)
circos.text(x = culture.means[5], 
            y = ys[5] - 0.05,
            labels = texts[5], 
            col = cols[5],
            facing = "bending.inside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[6], 
                y0 = ys[6],
                x1 = culture.ranges.ub[6], 
                y1 = ys[6], col = cols[6], 
                lwd = 2)
circos.text(x = culture.means[6], 
            y = ys[6] - 0.05,
            labels = texts[6], 
            col = cols[6],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[7], 
                y0 = ys[7], 
                x1 = culture.ranges.ub[7], 
                y1 = ys[7], col = cols[7], 
                lwd = 2)
circos.text(x = culture.means[7], 
            y = ys[7] - 0.05,
            labels = texts[7], 
            col = cols[7],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[8], 
                y0 = ys[8], 
                x1 = culture.ranges.ub[8], 
                y1 = ys[8], col = cols[8], 
                lwd = 2)
circos.text(x = culture.means[8], 
            y = ys[8] - 0.05,
            labels = texts[8], 
            col = cols[8],
            facing = "bending.inside",
            cex = 1)
# crosses zero
circos.segments(x0 = culture.ranges.lb[9], 
                y0 = ys[9], x1 = 360, 
                y1 = ys[9], col = cols[9], 
                lwd = 2)
circos.segments(x0 = 0, y0 = ys[9], 
                x1 = culture.ranges.ub[9], 
                y1 = ys[9], col = cols[9], 
                lwd = 2)
circos.text(x = culture.means[9], 
            y = ys[9] - 0.05,
            labels = texts[9], 
            col = cols[9],
            facing = "bending.inside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[10], 
                y0 = ys[10], 
                x1 = culture.ranges.ub[10], 
                y1 = ys[10], col = cols[10], 
                lwd = 2)
circos.text(x = culture.means[10], 
            y = ys[10] - 0.05,
            labels = texts[10], 
            col = cols[10],
            facing = "bending.outside",
            cex = 1)
circos.points(x = culture.means, y = ys, bg = cols, pch = 21)
circos.clear()
```

##### Model comparison: full, culture only, spatiotemporal only
```{r stats left comparison, echo=TRUE, results = "hide"}
# Full model
fit.left.full <- bpnr(Angles ~ 1 + Culture + Longitude + Latitude + Age, 
                data = circ.dat.left, its = 10000, burn = 1000, 
                n.lag = 3, seed = 1234)

# Spatiotemporal only
fit.left.sptemp <- bpnr(Angles ~ 1 + Longitude + Latitude + Age, 
                data = circ.dat.left, its = 10000, burn = 1000, 
                n.lag = 3, seed = 1234)
```
```{r stats left comparison2, echo=TRUE, eval=T}
# Full model
coef_circ(fit.left.full, type = "continuous")
coef_circ(fit.left.full, type = "categorical")$Means[1:10,]

# Spatiotemporal only
coef_circ(fit.left.sptemp, type = "continuous")

# Model fit
fit(fit.left)
fit(fit.left.full)
fit(fit.left.sptemp)
```

For the burial orientations with individuals buried on the left side, it seems that the culture only model has the best fit, meaning that culture is an important predictor.

#### Right

```{r stats cultures right, echo=TRUE, eval=T, results = "hide"}
# Add culture information to left data
right <- merge(right, orientation[, .(Num, Culture, Longitude, Latitude, Age)], by = "Num", all.x = T)
nrow(right)
nrow(right[!is.na(Culture)][Culture %in% culture_keep])

# Remove NAs in Culture and keep only 10 most common cultures
culture_right <- right[!is.na(Culture)][!is.na(Age)][Culture %in% culture_keep][, .(BurialOrientationMean, Culture, Longitude, Latitude, Age)]

# Transform to radians
circ.dat.right <- culture_right
circ.dat.right[, Angles := BurialOrientationMean*pi/180]
circ.dat.right$Culture <- factor(circ.dat.right$Culture, 
                               labels = c("Linearbandkeramik",
                                          "Lažňany",
                                          "Sopot",
                                          "Bell Beaker",
                                          "Dnieper-Donets",
                                          "Baden",
                                          "Corded Ware",
                                          "Trichterbecher",
                                          "Únětice",
                                          "Baltic Meso-Neolithic"),
                               levels = unique(circ.dat.right$Culture))

# Model
fit.right <- bpnr(Angles ~ 1 + Culture, 
                data = circ.dat.right, its = 10000, burn = 1000, 
                n.lag = 3, seed = 1234)
```
```{r stats cultures right2}
# Means
aov.dt.right <- setDT(data.frame(coef_circ(fit.right, "categorical", "radians")$Means))[1:10,]
aov.dt.right[, Culture := unique(circ.dat.right$Culture)]
# c1 <- sapply(strsplit(aov.dt.all$Culture, "Culture"), "[", 2)
# c2 <- sapply(strsplit(aov.dt.all$Culture, "Culture"), "[", 3)
# c1[is.na(c1)] <- c2[is.na(c2)] <- "Linearbandkeramik"
# aov.dt.all[, Culture := paste0(c1, "/", c2)]
aov.dt.right[, .(Culture, deg(.SD)), .SDcols = c("mean", "mode", "sd", "LB", "UB")]
```
```{r stats differences right, results = "hide"}
# Rotate values so that reference culture is always a new one
culture.labels <- as.character(unique(aov.dt.right$Culture))
culture.labels <- lapply(0:(length(culture.labels)-1), function(i) {
  c(culture.labels[(i+1):length(culture.labels)], culture.labels[1:i])
})
culture.labels[[1]] <- culture.labels[[1]][1:10]

# Regression model list
differences.right <- lapply(culture.labels, function(l){
  # data
  circ.dat.right$Culture <- as.character(circ.dat.right$Culture)
  circ.dat.right$Culture <- factor(circ.dat.right$Culture, 
                                 levels = l)
  # model
  fit.m <- bpnr(Angles ~ 1 + Culture, 
                data = circ.dat.right, its = 10000, burn = 1000, 
                n.lag = 3, seed = 1234)
  # differences
  diff.dt.right <- setDT(data.frame(coef_circ(fit.m, "categorical", units = "degrees")$Differences)[1:9,])
  diff.dt.right[, Difference := rownames(coef_circ(fit.m, "categorical", units = "degrees")$Differences)[1:9]]
  c1 <- sapply(strsplit(diff.dt.right$Difference, "Culture"), "[", 2)
  c2 <- sapply(strsplit(diff.dt.right$Difference, "Culture"), "[", 3)
  c2[is.na(c2)] <- l[1]
  diff.dt.right[, Difference := paste0(c2, "/", c1)]
  diff.dt.right[, .(Difference, mean, mode, sd, LB, UB)]
  diff.dt.right[, Significant := ifelse(between(0, LB, UB) | LB > UB, "No", "Yes")]
  return(diff.dt.right)
})
```
```{r stats differences right2}
# Remove duplicated comparisons
dt <- rbindlist(differences.right)
dt$Difference <- sapply(strsplit(dt$Difference, "/"), function(pair) {
  paste(gtools::mixedsort(pair), collapse = "/")
})
dt <- dt[!duplicated(Difference)][, .(Difference, mean, mode, sd, LB, UB, Significant)]
dt[, Culture1 := sapply(strsplit(dt$Difference, "/"), "[", 1)]
dt[, Culture2 := sapply(strsplit(dt$Difference, "/"), "[", 2)]
dt <- dt[, .(Difference, Culture1, Culture2, mean, mode, sd, LB, UB, Significant)][order(Culture1)]

datatable(dt[order(Culture1)][, .(Culture1, Culture2, mean, mode, sd, LB, UB, Significant, Difference)], options = list(scrollX = TRUE))
```
```{r stats cultures right3, echo = FALSE, fig.dim=c(15,15)}
## Visualization with circlize package
culture.means <- deg(aov.dt.right$mean)
culture.means <- ifelse(culture.means < 0, 360 + culture.means, culture.means)
culture.ranges.lb <- deg(aov.dt.right$LB)
culture.ranges.lb <- ifelse(culture.ranges.lb < 0, 360 + culture.ranges.lb, culture.ranges.lb)
culture.ranges.ub <- deg(aov.dt.right$UB)
culture.ranges.ub <- ifelse(culture.ranges.ub < 0, 360 + culture.ranges.ub, culture.ranges.ub)

# params
labs <- as.character(seq(0, 360, 22.5))
labs[17] <- "0"
cols <- rev(pals::ocean.thermal(12))[2:11]
ys <- seq(0, 1, length.out = 10)
texts <- levels(aov.dt.right$Culture)

# plot
circos.par(gap.degree = 0, cell.padding = c(0, 0, 0, 0), start.degree = 90)
circos.initialize("a", xlim = c(0, 360), ring = T)
circos.track(ylim = c(0,1), bg.border = NA)
circos.axis(major.at = seq(0, 360, 22.5), 
            direction = "outside",
            labels = labs)
circos.segments(x0 = culture.ranges.lb[1], 
                y0 = ys[1], 
                x1 = culture.ranges.ub[1], 
                y1 = ys[1], col = cols[1], lwd = 2)
circos.text(x = culture.means[1], 
            y = ys[1] - 0.05,
            labels = texts[1], 
            col = cols[1],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[2], 
                y0 = ys[2], 
                x1 = culture.ranges.ub[2], 
                y1 = ys[2], col = cols[2], 
                lwd = 2)
circos.text(x = culture.means[2], 
            y = ys[2] - 0.05,
            labels = texts[2], 
            col = cols[2],
            facing = "bending.outside",
            cex = 1)
# crosses zero
circos.segments(x0 = culture.ranges.lb[3], 
                y0 = ys[3], x1 = 360, 
                y1 = ys[3], col = cols[3], 
                lwd = 2)
circos.segments(x0 = 0, y0 = ys[3], 
                x1 = culture.ranges.ub[3], 
                y1 = ys[3], col = cols[3], 
                lwd = 2)
circos.text(x = culture.means[3], 
            y = ys[3] - 0.05,
            labels = texts[3], 
            col = cols[3],
            facing = "bending.inside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[4], 
                y0 = ys[4], 
                x1 = culture.ranges.ub[4], 
                y1 = ys[4], col = cols[4], 
                lwd = 2)
circos.text(x = culture.means[4], 
            y = ys[4] - 0.05,
            labels = texts[4], 
            col = cols[4],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[5], 
                y0 = ys[5], 
                x1 = culture.ranges.ub[5], 
                y1 = ys[5], col = cols[5], 
                lwd = 2)
circos.text(x = culture.means[5], 
            y = ys[5] - 0.05,
            labels = texts[5], 
            col = cols[5],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[6], 
                y0 = ys[6],
                x1 = culture.ranges.ub[6], 
                y1 = ys[6], col = cols[6], 
                lwd = 2)
circos.text(x = culture.means[6], 
            y = ys[6] - 0.05,
            labels = texts[6], 
            col = cols[6],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[7], 
                y0 = ys[7], 
                x1 = culture.ranges.ub[7], 
                y1 = ys[7], col = cols[7], 
                lwd = 2)
circos.text(x = culture.means[7], 
            y = ys[7] - 0.05,
            labels = texts[7], 
            col = cols[7],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[8], 
                y0 = ys[8], 
                x1 = culture.ranges.ub[8], 
                y1 = ys[8], col = cols[8], 
                lwd = 2)
circos.text(x = culture.means[8], 
            y = ys[8] - 0.05,
            labels = texts[8], 
            col = cols[8],
            facing = "bending.outside",
            cex = 1)
# crosses zero
circos.segments(x0 = culture.ranges.lb[9], 
                y0 = ys[9], 
                x1 = culture.ranges.ub[9], 
                y1 = ys[9], col = cols[9], 
                lwd = 2)
circos.text(x = culture.means[9], 
            y = ys[9] - 0.05,
            labels = texts[9], 
            col = cols[9],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[10], 
                y0 = ys[10], 
                x1 = culture.ranges.ub[10], 
                y1 = ys[10], col = cols[10], 
                lwd = 2)
circos.text(x = culture.means[10], 
            y = ys[10] - 0.05,
            labels = texts[10], 
            col = cols[10],
            facing = "bending.outside",
            cex = 1)
circos.points(x = culture.means, y = ys, bg = cols, pch = 21)
circos.clear()
```

##### Model comparison: full, culture only, spatiotemporal only
```{r stats right comparison, echo=TRUE, results = "hide"}
# Full model
fit.right.full <- bpnr(Angles ~ 1 + Culture + Longitude + Latitude + Age, 
                data = circ.dat.right, its = 10000, burn = 1000, 
                n.lag = 3, seed = 1234)

# Spatiotemporal only
fit.right.sptemp <- bpnr(Angles ~ 1 + Longitude + Latitude + Age, 
                data = circ.dat.right, its = 10000, burn = 1000, 
                n.lag = 3, seed = 1234)
```
```{r stats right comparison2, echo=TRUE, eval=T}
# Full model
coef_circ(fit.right.full, type = "continuous")
coef_circ(fit.right.full, type = "categorical")$Means[1:10,]

# Spatiotemporal only
coef_circ(fit.right.sptemp, type = "continuous")

# Model fit
fit(fit.right)
fit(fit.right.full)
fit(fit.right.sptemp)
```

For the right side, the culture-only model is the one with the best fit. However, the spatiotemporal only model gives very similar scores and the full model is the one with the worst fit. This means that having more covariates in the model produces a worse performance.

#### Back

When filtering for the back position, we have **no** individuals belonging to the Lažňany culture.

```{r stats cultures back, echo=TRUE, eval=T, results = "hide"}
# Add culture information to left data
back <- merge(back, orientation[, .(Num, Culture, Longitude, Latitude, Age)], by = "Num", all.x = T)
nrow(back)
nrow(back[!is.na(Culture)][Culture %in% culture_keep])

# Remove NAs in Culture and keep only 10 most common cultures
culture_back <- back[!is.na(Culture)][!is.na(Age)][Culture %in% culture_keep][, .(BurialOrientationMean, Culture, Longitude, Latitude, Age)]

# Transform to radians
circ.dat.back <- culture_back
circ.dat.back[, Angles := BurialOrientationMean*pi/180]
circ.dat.back$Culture <- factor(circ.dat.back$Culture, 
                               labels = c("Linearbandkeramik",
                                          # "Lažňany",
                                          "Sopot",
                                          "Bell Beaker",
                                          "Únětice",
                                          "Baden",
                                          "Corded Ware",
                                          "Trichterbecher",
                                          "Baltic Meso-Neolithic",
                                          "Dnieper-Donets"),
                               levels = unique(circ.dat.back$Culture))

# Model
fit.back <- bpnr(Angles ~ 1 + Culture, 
                data = circ.dat.back, its = 10000, burn = 1000, 
                n.lag = 3, seed = 1234)
```
```{r stats cultures back2}
# Means
aov.dt.back <- setDT(data.frame(coef_circ(fit.back, "categorical", "radians")$Means))[1:9,]
aov.dt.back[, Culture := unique(circ.dat.back$Culture)]
# c1 <- sapply(strsplit(aov.dt.all$Culture, "Culture"), "[", 2)
# c2 <- sapply(strsplit(aov.dt.all$Culture, "Culture"), "[", 3)
# c1[is.na(c1)] <- c2[is.na(c2)] <- "Linearbandkeramik"
# aov.dt.all[, Culture := paste0(c1, "/", c2)]
aov.dt.back[, .(Culture, deg(.SD)), .SDcols = c("mean", "mode", "sd", "LB", "UB")]
```
```{r stats differences back, results = "hide"}
# Rotate values so that reference culture is always a new one
culture.labels <- as.character(unique(aov.dt.back$Culture))
culture.labels <- lapply(0:(length(culture.labels)-1), function(i) {
  c(culture.labels[(i+1):length(culture.labels)], culture.labels[1:i])
})
culture.labels[[1]] <- culture.labels[[1]][1:9]

# Regression model list
differences.back <- lapply(culture.labels, function(l){
  # data
  circ.dat.back$Culture <- as.character(circ.dat.back$Culture)
  circ.dat.back$Culture <- factor(circ.dat.back$Culture, 
                                 levels = l)
  # model
  fit.m <- bpnr(Angles ~ 1 + Culture, 
                data = circ.dat.back, its = 10000, burn = 1000, 
                n.lag = 3, seed = 1234)
  # differences
  diff.dt.back <- setDT(data.frame(coef_circ(fit.m, "categorical", units = "degrees")$Differences)[1:8,])
  diff.dt.back[, Difference := rownames(coef_circ(fit.m, "categorical", units = "degrees")$Differences)[1:8]]
  c1 <- sapply(strsplit(diff.dt.back$Difference, "Culture"), "[", 2)
  c2 <- sapply(strsplit(diff.dt.back$Difference, "Culture"), "[", 3)
  c2[is.na(c2)] <- l[1]
  diff.dt.back[, Difference := paste0(c2, "/", c1)]
  diff.dt.back[, .(Difference, mean, mode, sd, LB, UB)]
  diff.dt.back[, Significant := ifelse(between(0, LB, UB) | LB > UB, "No", "Yes")]
  return(diff.dt.back)
})
```
```{r stats differences back2}
# Remove duplicated comparisons
dt <- rbindlist(differences.back)
dt$Difference <- sapply(strsplit(dt$Difference, "/"), function(pair) {
  paste(gtools::mixedsort(pair), collapse = "/")
})
dt <- dt[!duplicated(Difference)][, .(Difference, mean, mode, sd, LB, UB, Significant)]
dt[, Culture1 := sapply(strsplit(dt$Difference, "/"), "[", 1)]
dt[, Culture2 := sapply(strsplit(dt$Difference, "/"), "[", 2)]
dt <- dt[, .(Difference, Culture1, Culture2, mean, mode, sd, LB, UB, Significant)][order(Culture1)]

datatable(dt[order(Culture1)][, .(Culture1, Culture2, mean, mode, sd, LB, UB, Significant, Difference)], options = list(scrollX = TRUE))
```
```{r stats cultures back3, echo = FALSE, fig.dim=c(15,15)}
## Visualization with circlize package
culture.means <- deg(aov.dt.back$mean)
culture.means <- ifelse(culture.means < 0, 360 + culture.means, culture.means)
culture.ranges.lb <- deg(aov.dt.back$LB)
culture.ranges.lb <- ifelse(culture.ranges.lb < 0, 360 + culture.ranges.lb, culture.ranges.lb)
culture.ranges.ub <- deg(aov.dt.back$UB)
culture.ranges.ub <- ifelse(culture.ranges.ub < 0, 360 + culture.ranges.ub, culture.ranges.ub)

# params
labs <- as.character(seq(0, 360, 22.5))
labs[17] <- "0"
cols <- rev(pals::ocean.thermal(12))[2:11]
ys <- seq(0, 1, length.out = 9)
texts <- levels(aov.dt.back$Culture)

# plot
circos.par(gap.degree = 0, cell.padding = c(0, 0, 0, 0), start.degree = 90)
circos.initialize("a", xlim = c(0, 360), ring = T)
circos.track(ylim = c(0,1), bg.border = NA)
circos.axis(major.at = seq(0, 360, 22.5), 
            direction = "outside",
            labels = labs)
# cross zero
circos.segments(x0 = culture.ranges.lb[1], 
                y0 = ys[1], 
                x1 = 360, 
                y1 = ys[1], col = cols[1], lwd = 2)
circos.segments(x0 = 0, 
                y0 = ys[1], 
                x1 = culture.ranges.ub[1], 
                y1 = ys[1], col = cols[1], lwd = 2)
circos.text(x = culture.means[1], 
            y = ys[1] - 0.05,
            labels = texts[1], 
            col = cols[1],
            facing = "bending.inside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[2], 
                y0 = ys[2], 
                x1 = culture.ranges.ub[2], 
                y1 = ys[2], col = cols[2], 
                lwd = 2)
circos.text(x = culture.means[2], 
            y = ys[2] - 0.05,
            labels = texts[2], 
            col = cols[2],
            facing = "bending.inside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[3], 
                y0 = ys[3], 
                x1 = culture.ranges.ub[3], 
                y1 = ys[3], col = cols[3], 
                lwd = 2)
circos.text(x = culture.means[3], 
            y = ys[3] - 0.05,
            labels = texts[3], 
            col = cols[3],
            facing = "bending.inside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[4], 
                y0 = ys[4], 
                x1 = culture.ranges.ub[4], 
                y1 = ys[4], col = cols[4], 
                lwd = 2)
circos.text(x = culture.means[4], 
            y = ys[4] - 0.05,
            labels = texts[4], 
            col = cols[4],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[5], 
                y0 = ys[5], 
                x1 = culture.ranges.ub[5], 
                y1 = ys[5], col = cols[5], 
                lwd = 2)
circos.text(x = culture.means[5], 
            y = ys[5] - 0.05,
            labels = texts[5], 
            col = cols[5],
            facing = "bending.outside",
            cex = 1)
# cross zero
circos.segments(x0 = culture.ranges.lb[6], 
                y0 = ys[6],
                x1 = 360, 
                y1 = ys[6], col = cols[6], 
                lwd = 2)
circos.segments(x0 = 0, 
                y0 = ys[6],
                x1 = culture.ranges.ub[6], 
                y1 = ys[6], col = cols[6], 
                lwd = 2)
circos.text(x = culture.means[6], 
            y = ys[6] - 0.05,
            labels = texts[6], 
            col = cols[6],
            facing = "bending.outside",
            cex = 1)
# cross zero
circos.segments(x0 = culture.ranges.lb[7], 
                y0 = ys[7], 
                x1 = 360, 
                y1 = ys[7], col = cols[7], 
                lwd = 2)
circos.segments(x0 = 0, 
                y0 = ys[7], 
                x1 = culture.ranges.ub[7], 
                y1 = ys[7], col = cols[7], 
                lwd = 2)
circos.text(x = culture.means[7], 
            y = ys[7] - 0.05,
            labels = texts[7], 
            col = cols[7],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[8], 
                y0 = ys[8], 
                x1 = culture.ranges.ub[8], 
                y1 = ys[8], col = cols[8], 
                lwd = 2)
circos.text(x = culture.means[8], 
            y = ys[8] - 0.05,
            labels = texts[8], 
            col = cols[8],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[9], 
                y0 = ys[9], 
                x1 = culture.ranges.ub[9], 
                y1 = ys[9], col = cols[9], 
                lwd = 2)
circos.text(x = culture.means[9], 
            y = ys[9] - 0.05,
            labels = texts[9], 
            col = cols[9],
            facing = "bending.outside",
            cex = 1)
circos.points(x = culture.means, y = ys, bg = cols, pch = 21)
circos.clear()
```


```{r stats back comparison, echo=FALSE, results = "hide", eval = FALSE}
# Full model
fit.back.full <- bpnr(Angles ~ 1 + Culture + Longitude + Latitude + Age, 
                data = circ.dat.back, its = 10000, burn = 1000, 
                n.lag = 3, seed = 1234)

# Spatiotemporal only
fit.back.sptemp <- bpnr(Angles ~ 1 + Longitude + Latitude + Age, 
                data = circ.dat.back, its = 10000, burn = 1000, 
                n.lag = 3, seed = 1234)
```
```{r stats back comparison2, echo=FALSE, eval=F}
# Full model
coef_circ(fit.back.full, type = "continuous")
coef_circ(fit.back.full, type = "categorical")$Means[1:10,]

# Spatiotemporal only
coef_circ(fit.back.sptemp, type = "continuous")

# Model fit
fit(fit.back)
fit(fit.back.full)
fit(fit.back.sptemp)
```

## Sex

```{r stats sex}
# Group sex
orientation[, Sex := ifelse(Sex < 0.5, 0, ifelse(Sex > 0.5, 1, Sex))]

## Circular means
sex <- split(orientation[, .(BurialOrientationMean, Sex = as.factor(Sex))], by = "Sex")
sex_means <- lapply(sex, function(x){
  mean.circular(circular(x$BurialOrientationMean, 
                         units = "degrees", 
                         type = "angles", 
                         rotation = "clock"))
})
sex_means <- unlist(sex_means)
sex_means <- sex_means[gtools::mixedsort(names(sex_means))]
d <- list()
for(a in 1:length(sex_means)){
  d[[a]] <- sapply(1:length(sex_means), function(x){
    angular_distance(sex_means[a], sex_means[x])
  })
}
sex_mat <- do.call(cbind, d)
colnames(sex_mat) <- rownames(sex_mat) <- paste0("Prob_", names(sex_means))
sex_mat[lower.tri(sex_mat)] <- NA
sex_mat
sex_dists <- reshape2::melt(sex_mat)

# Plot burial orientations per period
sex_means_dt <- data.table(Sex = names(sex_means), Mean = sex_means)
sex_means_dt[, Mean := ifelse(Mean < 0, Mean + 360, Mean)]

cols <- rev(pals::brewer.piyg(3))
cols[2] <- "grey80"

ggplot(data = orientation, 
       aes(x = as.factor(Sex), y = BurialOrientationMean, 
           fill = as.factor(Sex))) +
  geom_violin(width = 1, drop = F) +
  geom_point(data = sex_means_dt, aes(x = Sex, y = Mean), 
             size = 2, shape = 23, fill = "white") +
  scale_fill_manual(values = cols) +
  ylab("Burial orientation") +
  theme_light() +
  theme(legend.position = "none",
        axis.title.x = element_blank())

# Plot pairwise angular distances
ggplot(data = sex_dists[!is.na(sex_dists$value),]) +
  geom_tile(aes(Var1, Var2, fill = value), color = "black") +
  scale_fill_gradientn("Angular distance", colours = rev(pals::brewer.rdylbu(100)), 
                       na.value = "transparent",
                       limits = c(0, 180)) +
   geom_text(aes(Var1, Var2, label = round(value, 1)), 
            size = 10, size.unit = "pt") +
  scale_x_discrete(position = "top") +
  ggtitle("Pairwise distance between angles means") +
  coord_equal(expand = F) +
  theme_classic() +
  theme(plot.title = element_text(face = "bold", size = 16),
        axis.text.x = element_text(angle = 45, hjust = 0),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "none",
        axis.text = element_text(size = 14))
```

```{r stats sex bpnreg, results = "hide"}
## Bayesian circular regression
sex.dt <- orientation[, .(Angles = rad(BurialOrientationMean), Sex)]
sex.dt <- sex.dt[!is.na(Angles)]
sex.dt$Sex <- factor(sex.dt$Sex, levels = c("0", "0.5", "1"))

fit.sex <- bpnr(Angles ~ 1 + Sex, data = sex.dt, 
            its = 10000, burn = 100, n.lag = 3, seed = 1234)
```

```{r stats sex bpnreg2}
# Means
sex.aov.dt <- setDT(data.frame(coef_circ(fit.sex, "categorical", "radians")$Means))[-4,]
sex.aov.dt[, Sex := levels(sex.dt$Sex)]
sex.aov.dt <- sex.aov.dt[, .(Sex, deg(.SD)), .SDcols = c("mean", "mode", "sd", "LB", "UB")]
sex.aov.dt

# Plot
cols <- rev(pals::brewer.piyg(3))
cols[2] <- "grey80"
ys <- seq(0, 1, 0.5)

circos.par(gap.degree = 0, cell.padding = c(0, 0, 0, 0), start.degree = 90)
circos.initialize("a", xlim = c(0, 360), ring = T)
circos.track(ylim = c(0,1), bg.border = NA)
circos.axis(major.at = seq(0, 360, 22.5), 
            direction = "outside",
            labels = labs)
circos.segments(x0 = sex.aov.dt$LB[1], y0 = ys[1], 
                x1 = sex.aov.dt$UB[1], y1 = ys[1],
                col = cols[1], lwd = 2)
circos.segments(x0 = sex.aov.dt$LB[2], y0 = ys[2], 
                x1 = sex.aov.dt$UB[2], y1 = ys[2],
                col = cols[2], lwd = 2)
circos.segments(x0 = sex.aov.dt$LB[3], y0 = ys[3], 
                x1 = sex.aov.dt$UB[3], y1 = ys[3],
                col = cols[3], lwd = 2)
circos.points(x = sex.aov.dt$mean[1], y = ys[1],
              bg = cols[1], pch = 21)
circos.points(x = sex.aov.dt$mean[2], y = ys[2],
              bg = cols[2], pch = 21)
circos.points(x = sex.aov.dt$mean[3], y = ys[3],
              bg = cols[3], pch = 21)
circos.clear()

# Differences
diff.dt <- setDT(data.frame(coef_circ(fit.sex, "categorical", units = "degrees")$Differences)[1:2,])
diff.dt[, Difference := c("0/0.5",
                          "0/1")]
diff.dt[, .(Difference, mean, mode, sd, LB, UB)]
```

It seems that 0 is different from 0.5 and 1. So we do see a difference between males and females, although this is of ~20 degrees. There is however no difference between females and the undetermined individuals.

## Circular linear regression models

For data that is not categorical (e.g., ancestry and mobility values), we have to use a regression model that is specific to angular data.

This regression models both the mean direction and the dispersion simultaneously. The model assumes that a circular response variable $\theta$ has a von Mises distribution with concentration parameter $\kappa$, and mean direction ($\mu$) related to a vector of linear predictor variables according to the relationship: $\mu + 2*atan(\beta*x)$, where $\mu$ and $\beta$ are unknown parameters, $\beta$ being a vector of regression coefficients, and $2*atan()$ being the link function.

The von Mises distribution is a function of $\mu$ and $\kappa$, where $\mu$ is the mean direction and $\kappa$ is the concentration parameter. A value $\kappa = 0$ corresponds to the circular uniform distribution; for $\kappa > 0$, the density is unimodal and symmetrical about $\mu$, and increasingly concentrated as $\kappa$ increases. For $\kappa \ge 2$ the density at the antipode $\mu + \pi$ is effectively zero, and the distribution is well approximated by a normal distribution with variance $1/\kappa$.

We can start with an initial value for $\beta$, to calculate $\mu$ and $\kappa$. These estimates can then be used to update the $\beta$ value, and then to update $\mu$ and $\kappa$, analogously to the ML equations for an independent, identically distributed (i.i.d.) von Mises sample. This cycle is repeated until convergence. If $\beta = 0$, we have a null model where all observations cluster around a common mean direction $\mu$. However, as the elements of $\beta$ become infinite, the use of any monotone link function also leads to a null model with mean direction $\mu ± \pi$. In such cases, the likelihood will have a peak in the vicinity of zero, and attain a similar magnitude as the elements of $\beta$ tend to $± \infty$. **The optimization process is simplified if the x are centered at their means**.

**So should we scale our ancestry values?** (Whether we order the values or scale them, model results don't seem to change)

Further reads: <https://www.jstor.org/stable/2532334?seq=2>; <http://iopscience.iop.org/article/10.1088/1742-6596/1988/1/012108>

The `circular` package, has a `lm.circular` function that can be used for circular and linear data (c-l). We are going to test this function using the ancestry values associated to the grave individuals.

### CHG

```{r stats chg, fig.dim=c(10,10)}
y <- circular(rad(orientation[!is.na(CHG)]$BurialOrientationMean), units = "radians", type = "angles", rotation = "clock")
x <- cbind(orientation[!is.na(CHG)]$CHG, rep(1,length(orientation[!is.na(CHG)]$CHG)))
colnames(x) <- c("CHG", "Intercept")
mu <- as.numeric(mean.circular(y))
reg.chg <- lm.circular(y = y, x = x, type = "c-l", init = c(0,1))
reg.chg
```

We can see that CHG has a significant positive relationship with the mean direction of the burials. This test was made by using all the data as is. But what if we repeat the analysis by averaging the CHG value of the individuals that have the same burial orientation? Similarly to what we did to plot them.

```{r stats chg2}
## Based on average value across individuals with the same angle
dim_chg <- orientation[, mean(CHG, na.rm = T), by = c("BurialOrientationMean_sin", "BurialOrientationMean_cos")]
dim_chg <- dim_chg[!is.nan(V1),]
names(dim_chg)[3] <- "CHG"
# Get the angle from sin and cos (in radians)
dim_chg[, BurialOrientationMean := sapply(1:nrow(dim_chg), function(x) atan2(BurialOrientationMean_sin[x], BurialOrientationMean_cos[x]))]

# New variables for the linear regression model
y2 <- circular(dim_chg$BurialOrientationMean, units = "radians", type = "angles", rotation = "clock")
x2 <- cbind(dim_chg$CHG, rep(1, nrow(dim_chg)))
colnames(x2) <- c("CHG", "Intercept")
mu <- as.numeric(mean.circular(y2))
m <- mean(x2[,1])

# Model
reg.chg2 <- lm.circular(y = y2, x = x2, type = "c-l", init = c(0,1))
reg.chg2

# Plot
mu.df <- data.frame(mu = c(mu, reg.chg2$mu), type = c("Observed", "Estimated"))

ggplot() +
  geom_point(data = dim_chg,
             aes(BurialOrientationMean_sin, BurialOrientationMean_cos, fill = CHG),
             shape = 21,
             size = 5,
             ) +
  geom_point(data = mu.df, aes(sin(mu), cos(mu), shape = type), size = 8) +
  scale_fill_gradientn("CHG", colors = pals::parula(100)) +
  scale_shape_manual(expression(mu), values = c(7, 8)) +
  xlab(expression(sin(theta))) +
  ylab(expression(cos(theta))) +
  coord_equal() +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16))
```

### WHG

```{r stats whg, fig.dim = c(10,10)}
y <- circular(rad(orientation[!is.na(WHG)]$BurialOrientationMean), units = "radians", type = "angles", rotation = "clock")
x <- cbind(orientation[!is.na(WHG)]$WHG, rep(1,length(orientation[!is.na(WHG)]$WHG)))
colnames(x) <- c("WHG", "Intercept")
mu <- as.numeric(mean.circular(y))
reg.whg <- lm.circular(y = y, x = x, type = "c-l", init = c(0,1))
reg.whg
```

### EHG

```{r stats ehg}
y <- circular(rad(orientation[!is.na(EHG)]$BurialOrientationMean), units = "radians", type = "angles", rotation = "clock")
x <- cbind(orientation[!is.na(EHG)]$EHG, rep(1,length(orientation[!is.na(EHG)]$EHG)))
colnames(x) <- c("EHG", "Intercept")
mu <- as.numeric(mean.circular(y))
reg.ehg <- lm.circular(y = y, x = x, type = "c-l", init = c(0,1))
reg.ehg
```

### LVN

```{r stats lvn}
y <- circular(rad(orientation[!is.na(LVN)]$BurialOrientationMean), units = "radians", type = "angles", rotation = "clock")
x <- cbind(orientation[!is.na(LVN)]$LVN, rep(1,length(orientation[!is.na(LVN)]$LVN)))
colnames(x) <- c("LVN", "Intercept")
mu <- as.numeric(mean.circular(y))
reg.lvn <- lm.circular(y = y, x = x, type = "c-l", init = c(0,1))
reg.lvn
```

### Mobility

```{r stats mobility}
y <- circular(rad(orientation[!is.na(mobility)]$BurialOrientationMean), units = "radians", type = "angles", rotation = "clock")
x <- cbind(orientation[!is.na(mobility)]$mobility, rep(1,length(orientation[!is.na(mobility)]$mobility)))
colnames(x) <- c("Mobility", "Intercept")
mu <- as.numeric(mean.circular(y))
reg.mob <- lm.circular(y = y, x = x, type = "c-l", init = c(0,1))
reg.mob
```

## Bayesian circular regression with continuous data in `bpnreg`

We can also try using the same Bayesian regression model that we used for the categories.

For continuous explanatory variables we get a set of parameters, *bc*, *SAM* and *AS*, describing its effect on the circle. The coefficient *bc* represents the slope of the circular regression line at the inflection point. However, this may not be a representative effect for each dataset, as the inflection point can lie in the extremes of the data or even completely outside the range of the predictor x. Therefore, two additional circular coefficients were developed by Cremers et al. (2018), the slope at the mean *SAM* and the average slope *AS.* The coefficient *SAM* represents the slope of the circular regression line at the average of the predictor and the coefficients *AS* represents the average slope over all values of x.

If the HPD intervals for all three circular coefficients (*bc, SAM and AS*) include 0, we do not find evidence that at the inflection point, at the average predictor value and on average the covariate influences the angular error.

<https://www.frontiersin.org/journals/psychology/articles/10.3389/fpsyg.2018.02040/full#B5> <https://bpspsychub.onlinelibrary.wiley.com/doi/10.1111/bmsp.12108>

### Full model: space, time, culture, 3 ancestries

In the culture models we did above, we saw that culture does seem to be an important predictor of burial orientation. We can now add the three ancestries (CHG, EHG, LVN), the geographic location and time to see if they also are good predictors of orientation. We start with the full model and then we will remove covariates to see which model has the best fit (DIC or WAIC score) to the data.

We need to consider that we only have a subset of data with full information about culture and ancestry, so the number of individuals included in the model are less compared to the regressions we did with culture. This will likely change the posterior densities of the coefficients for the cultures, so the circular plot will probably look different than the one above, even for the culture only model.

```{r stats bpnreg continuous, results = "hide"}
ance.dt <- orientation[, .(Angles = rad(BurialOrientationMean),
                          WHG, CHG, EHG, LVN, Longitude, Latitude, Age, Culture)]
ance.dt <- ance.dt[Culture %in% culture_keep]
ance.dt <- ance.dt[!is.na(WHG)]
ance.dt <- ance.dt[!is.na(Age)]
ance.dt <- ance.dt[, .(Angles, scale(.SD), Culture), .SDcols = c("WHG", "EHG", "CHG", "LVN", "Longitude", "Latitude", "Age")]
ance.dt$Culture <- factor(ance.dt$Culture, 
                               labels = c("Linearbandkeramik",
                                          "Lažňany",
                                          "Sopot",
                                          "Bell Beaker",
                                          "Únětice",
                                          "Corded Ware",
                                          "Baden",
                                          "Dnieper-Donets",
                                          "Trichterbecher",
                                          "Baltic Meso-Neolithic"),
                               levels = unique(ance.dt$Culture))

fit.full <- bpnr(Angles ~ 1 + CHG + EHG + LVN + Latitude + Longitude + Age + Culture, data = ance.dt, its = 10000, burn = 1000, 
                n.lag = 3, seed = 1234)
```

Based on the table for linear coefficients, it seems that none of the ancestries have an effect on the two components (sine, cosine) of the angles. Now to determine whether there is an effect on the circle, we have to look at the posterior HPDs of the circular coefficients, specifically bc, AS, SAM and SSDO. These will tell us if there is an accuracy or location effect.

```{r stats bpnreg continuous2}
fit.full
```

When looking at the table, we can see that the HPDs for *bc* cross zero for all ancestries. This is also true for the HPDs for *SSDO*. If the HPD interval of the *SSDO* of a predictor does not include 0, the bivariate regression line does not run through the origin and therefore there is a location effect. Based on this and our HPDs for *SSDO*, we can conclude that **we do not have a location effect**. The effect that we see on the two components might therefore be an accuracy effect, an effect on the spread of the angles around the circles and not on the mean direction. *SAM* is the slope of the regression curve at the mean of the predictor data and the HPDs are outside zero for CHG. This means that at the average CHG, with 1 unit increase we have an increase in the spread around the circle of 2˚, respectively. When looking at the average slope at each predictor data point (*AS*), the HPD is outside zero only for CHG. This means that on average, a unit increase in CHG results in an increase of 2.5˚ in the spread around the circle.

### Space, time, culture

We should also run a model that does not contain WHG.

```{r bpnreg sptc, results = "hide"}
fit.sptc <- bpnr(Angles ~ 1 + Latitude + Longitude + Age + Culture, data = ance.dt, its = 10000, burn = 1000, n.lag = 3, seed = 1234)
```
```{r bpnreg sptc2}
fit.sptc
```

### Space, time, ancestry
```{r bpnreg spta, results = "hide"}
fit.spta <- bpnr(Angles ~ 1 + CHG + EHG + LVN + Latitude + Longitude + Age, data = ance.dt, its = 10000, burn = 1000, n.lag = 3, seed = 1234)
```
```{r bpnreg spta2}
fit.spta
```

### Culture only model
```{r bpnreg cult, results = "hide"}
fit.cult <- bpnr(Angles ~ 1 + Culture, data = ance.dt, its = 10000, burn = 1000, n.lag = 3, seed = 1234)
```
```{r bpnreg cult2}
fit.cult
```

### Ancestry only model
```{r bpnreg ance, results = "hide"}
fit.ance <- bpnr(Angles ~ 1 + CHG + EHG + LVN, data = ance.dt, its = 10000, burn = 1000, n.lag = 3, seed = 1234)
```
```{r bpnreg ance2}
fit.ance
```

In the ancestry only model, so when we do not include space, time or culture, we can see that the model predicts an effect of CHG on the first component. When looking at the effect in the circle, the model shows an effect of CHG on *bc* and of all ancestries on *SSDO*. This indicates a location effect. However, we still have to determine if this model has a good fit to the data, or if other covariates are better at explaining orientation.

### Space-time only model
```{r bpnreg spt, results = "hide"}
fit.spt <- bpnr(Angles ~ 1 + Longitude + Latitude + Age, data = ance.dt, its = 10000, burn = 1000, n.lag = 3, seed = 1234)
```
```{r bpnreg spt2}
fit.spt
```

### Models fit
```{r bpnreg fit}
fit(fit.full)
fit(fit.sptc)
fit(fit.spta)
fit(fit.cult)
fit(fit.ance)
fit(fit.spt)

fit.list<- lapply(list(fit.full, fit.sptc, fit.spta, fit.cult, fit.ance, fit.spt), fit)
fit.dic <- lapply(fit.list, function(x){x[2,1]})
fit.dt <- setDT(as.data.frame(do.call(rbind, fit.dic)))
fit.dt[, Model := c("Space + time + ancestry + culture",
                    "Space + time + culture",
                    "Space + time + ancestry",
                    "Culture only",
                    "Ancestry only",
                    "Space + time")]
fit.dt[, WAIC := do.call(rbind, lapply(fit.list, function(x){x[4,1]}))]

ggplot(data = fit.dt) +
  geom_point(aes(x = Model, y = V1), size = 3) +
  geom_line(aes(x = Model, y = V1, group = 1)) +
  ylab("DIC") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank())

ggplot(data = fit.dt) +
  geom_point(aes(x = Model, y = WAIC), size = 3) +
  geom_line(aes(x = Model, y = WAIC, group = 1)) +
  ylab("WAIC") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank())
```

Based on both DIC and WAIC scores, we can see that the best models are always the ones that include culture. This means that culture seems to be the most important predictor of burial orientation, compared to ancestry or just the spatiotemporal location.

### Visualize cultures

Here we are going to visualize the differences between ancestries for this subset of data, i.e. grave individuals for which we have a value of ancestry and for which we know their culture and burial orientation.

```{r bpnreg cultures only, fig.dim = c(15,15)}
# Means
results.dt <- setDT(data.frame(coef_circ(fit.cult, "categorical", "radians")$Means))[1:10,]
results.dt[, Culture := levels(ance.dt$Culture)]
# c1 <- sapply(strsplit(aov.dt.all$Culture, "Culture"), "[", 2)
# c2 <- sapply(strsplit(aov.dt.all$Culture, "Culture"), "[", 3)
# c1[is.na(c1)] <- c2[is.na(c2)] <- "Linearbandkeramik"
# aov.dt.all[, Culture := paste0(c1, "/", c2)]
results.dt[, .(Culture, deg(.SD)), .SDcols = c("mean", "mode", "sd", "LB", "UB")]

## Visualization with circlize package
culture.means <- deg(results.dt$mean)
culture.means <- ifelse(culture.means < 0, 360 + culture.means, culture.means)
culture.ranges.lb <- deg(results.dt$LB)
culture.ranges.lb <- ifelse(culture.ranges.lb < 0, 360 + culture.ranges.lb, culture.ranges.lb)
culture.ranges.ub <- deg(results.dt$UB)
culture.ranges.ub <- ifelse(culture.ranges.ub < 0, 360 + culture.ranges.ub, culture.ranges.ub)

# params
labs <- as.character(seq(0, 360, 22.5))
labs[17] <- "0"
cols <- rev(pals::ocean.thermal(12))[2:11]
ys <- seq(0, 1, length.out = 10)
texts <- results.dt$Culture

# plot
circos.par(gap.degree = 0, cell.padding = c(0, 0, 0, 0), start.degree = 90)
circos.initialize("a", xlim = c(0, 360), ring = T)
circos.track(ylim = c(0.75, 1.05), bg.border = NA)
circos.axis(major.at = seq(0, 360, 22.5), 
            direction = "outside",
            labels = labs)
circos.segments(x0 = culture.ranges.lb[1], 
                y0 = ys[1], 
                x1 = culture.ranges.ub[1], 
                y1 = ys[1], col = cols[1], lwd = 2)
circos.text(x = culture.means[1], 
            y = ys[1] - 0.05,
            labels = texts[1], 
            col = cols[1],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[2], 
                y0 = ys[2], 
                x1 = culture.ranges.ub[2], 
                y1 = ys[2], col = cols[2], 
                lwd = 2)
circos.text(x = culture.means[2], 
            y = ys[2] - 0.05,
            labels = texts[2], 
            col = cols[2],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[3], 
                y0 = ys[3], 
                x1 = culture.ranges.ub[3], 
                y1 = ys[3], col = cols[3], 
                lwd = 2)
circos.text(x = culture.means[3], 
            y = ys[3] - 0.05,
            labels = texts[3], 
            col = cols[3],
            facing = "bending.inside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[4], 
                y0 = ys[4], 
                x1 = culture.ranges.ub[4], 
                y1 = ys[4], col = cols[4], 
                lwd = 2)
circos.text(x = culture.means[4], 
            y = ys[4] - 0.05,
            labels = texts[4], 
            col = cols[4],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[5], 
                y0 = ys[5], 
                x1 = culture.ranges.ub[5], 
                y1 = ys[5], col = cols[5], 
                lwd = 2)
circos.text(x = culture.means[5], 
            y = ys[5] - 0.05,
            labels = texts[5], 
            col = cols[5],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[6], 
                y0 = ys[6], 
                x1 = culture.ranges.ub[6], 
                y1 = ys[6], col = cols[6], 
                lwd = 2)
circos.text(x = culture.means[6], 
            y = ys[6] - 0.05,
            labels = texts[6], 
            col = cols[6],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[7], 
                y0 = ys[7], 
                x1 = culture.ranges.ub[7], 
                y1 = ys[7], col = cols[7], 
                lwd = 2)
circos.text(x = culture.means[7], 
            y = ys[7] - 0.05,
            labels = texts[7], 
            col = cols[7],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[8], 
                y0 = ys[8], 
                x1 = culture.ranges.ub[8], 
                y1 = ys[8], col = cols[8], 
                lwd = 2)
circos.text(x = culture.means[8], 
            y = ys[8] - 0.05,
            labels = texts[8], 
            col = cols[8],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[9], 
                y0 = ys[9], 
                x1 = culture.ranges.ub[9], 
                y1 = ys[9], col = cols[9], 
                lwd = 2)
circos.text(x = culture.means[9], 
            y = ys[9] - 0.05,
            labels = texts[9], 
            col = cols[9],
            facing = "bending.outside",
            cex = 1)
circos.segments(x0 = culture.ranges.lb[10], 
                y0 = ys[10], 
                x1 = culture.ranges.ub[10], 
                y1 = ys[10], col = cols[10], 
                lwd = 2)
circos.text(x = culture.means[10], 
            y = ys[10] - 0.05,
            labels = texts[10], 
            col = cols[10],
            facing = "bending.outside",
            cex = 1)
circos.points(x = culture.means, y = ys, bg = cols, pch = 21)
circos.clear()
```

```{r culture only table, results = "hide"}
# Rotate values so that reference culture is always a new one
culture.labels <- as.character(unique(ance.dt$Culture))
culture.labels <- lapply(0:(length(culture.labels)-1), function(i) {
  c(culture.labels[(i+1):length(culture.labels)], culture.labels[1:i])
})
culture.labels[[1]] <- culture.labels[[1]][1:10]

# Regression model list
differences.all <- lapply(culture.labels, function(l){
  # data
  ance.dt$Culture <- as.character(ance.dt$Culture)
  ance.dt$Culture <- factor(ance.dt$Culture, 
                                 levels = l)
  # model
  fit.m <- bpnr(Angles ~ 1 + Culture, 
                data = ance.dt, its = 10000, burn = 1000, 
                n.lag = 3, seed = 1234)
  # differences
  diff.dt <- setDT(data.frame(coef_circ(fit.m, "categorical", units = "degrees")$Differences)[1:9,])
  diff.dt[, Comparison := rownames(coef_circ(fit.m, "categorical", units = "degrees")$Differences)[1:9]]
  c1 <- sapply(strsplit(diff.dt$Comparison, "Culture"), "[", 2)
  c2 <- sapply(strsplit(diff.dt$Comparison, "Culture"), "[", 3)
  c2[is.na(c2)] <- l[1]
  diff.dt[, Comparison := paste0(c2, "/", c1)]
  diff.dt[, .(Comparison, mean, mode, sd, LB, UB)]
  diff.dt[, Different := ifelse(between(0, LB, UB) | LB > UB, "No", "Yes")]
  return(diff.dt)
})
```

```{r stats differences2}
# Remove duplicated comparisons
dt <- rbindlist(differences.all)
dt$Comparison <- sapply(strsplit(dt$Comparison, "/"), function(pair) {
  paste(gtools::mixedsort(pair), collapse = "/")
})
dt <- dt[!duplicated(Comparison)][, .(Comparison, mean, mode, sd, LB, UB, Overlapping)]
dt[, Culture1 := sapply(strsplit(dt$Comparison, "/"), "[", 1)]
dt[, Culture2 := sapply(strsplit(dt$Comparison, "/"), "[", 2)]
dt <- dt[, .(Comparison, Culture1, Culture2, mean, mode, sd, LB, UB, Overlapping)][order(Culture1)]
dt <- dt[, .(Comparison, Culture1, Culture2, round(.SD, 2), Overlapping), .SDcols = c("mean", "mode", "sd", "LB", "UB")]

fwrite(dt, "~/Library/CloudStorage/OneDrive-UniversityofCopenhagen/Projects/Graves/Results/BurialOrientationDifferences_culture_ance.csv")
```

### Percentage of explained variance by the "Culture only" model

We are going to try calculate the percentage of explained variance by the model with the best fit to the data, the "Culture only" model. To do so, we must manually compute the circular variance of the predicted values and divide it by the circular variance of the observed data. Because *bpnreg* fits an ANOVA via a Projected Normal linear model onto a bivariate latent space, standard linear \(R^{2}\) formulas do not work directly. We need to extract the posterior samples of the predicted circular means, convert them to circular variance, and compare them to the observed data.

```{r variance explained1}
# --- 1. Reconstruct the Model Matrix (Design Matrix) ---
# Create the linear design matrix (e.g., intercept and dummy-coded group factors)
XI <- fit.cult$mm$XI
XII <- fit.cult$mm$XII # 952 rows x K predictors

# --- 2. Extract Posterior Samples for Component I and II ---
# These are matrices of size: (MCMC iterations) x (K predictors)
B1 <- fit.cult$beta1 
B2 <- fit.cult$beta2

# --- 3. Compute Latent Linear Predictions ---
# Matrix multiplication yields: (952 individuals) x (MCMC iterations)
pred_I <- XI %*% t(B1)
pred_II <- XII %*% t(B2)

# --- 4. Transform Latent Space to Circular Predicted Angles ---
# atan2(y, x) transforms the bivariate space back onto the 0 to 2pi circular surface
pred_angles <- atan2(pred_Linear_II, pred_Linear_I)

# Adjust negative angles to be strictly between 0 and 2*pi
pred_angles <- ifelse(pred_angles < 0, pred_angles + 2 * pi, pred_angles)

# --- 5. Calculate the Circular Variance R² ---
# 'fit$theta' holds your actual observed radian outcome vector (952 rows)
observed_data <- circular(ance.dt$Angles, units = "radians", rotation = "clock")
total_var     <- var.circular(observed_data)

# Convert our manually generated prediction matrix into a circular object
pred_circular_matrix <- circular(pred_angles, units = "radians", rotation = "clock")

# Compute the R² distribution across columns (each iteration)
r2_posterior <- apply(pred_circular_matrix, 2, function(col) {
  var.circular(col) / total_var
})

# --- 6. Summarize Results ---
mean_r2 <- mean(r2_posterior) * 100
hpd_r2  <- quantile(r2_posterior, probs = c(0.025, 0.975)) * 100

# Print the final percentage of explained variance
cat("Percentage of Explained Variance (Mean R²):", round(mean_r2, 2), "%\n")
cat("95% Credible Interval:", round(hpd_r2[1], 2), "% to", round(hpd_r2[2], 2), "%\n")
```

We can also use the Gelman's Bayesian \(R^{2}\) framework on the Bayesian circular regression, but to do so we have to apply it to each latent linear components (I and II) individually. This is because *bpnreg* uses an embedding approach called the Projected Normal distribution. This means it takes your circular outcome and projects it onto a bivariate, 2D latent linear space (Component I and Component II). Gelman's \(R^{2}\) relies on linear combinations of variance (\(\text{Var}_{\text{fit}} / (\text{Var}_{\text{fit}} + \text{Var}_{\text{res}})\)), thus we cannot apply it directly to the circular surface angles. 

```{r variance explained2}

n_samples <- nrow(B1) # 10000

# Initialize vectors to track Gelman R² for each MCMC iteration
r2_component_I  <- numeric(n_samples)
r2_component_II <- numeric(n_samples)

# --- Compute Gelman's Variance Ratio Across MCMC Iterations ---
for (s in 1:n_samples) {
  
  # Linear predictions for this specific posterior slice (952 individual values)
  mu_I  <- XI %*% B1[s, ]
  mu_II <- XII %*% B2[s, ]
  
  # Calculate the sample variance of the model-fit values (Var_fit)
  var_fit_I  <- var(as.numeric(mu_I))
  var_fit_II <- var(as.numeric(mu_II))
  
  # Standard Projected Normal models fix the residual error variance to 1.0
  var_res_latent <- 1.0
  
  # Apply Gelman's Formula: Var_fit / (Var_fit + Var_res)
  r2_component_I[s]  <- var_fit_I  / (var_fit_I  + var_res_latent)
  r2_component_II[s] <- var_fit_II / (var_fit_II + var_res_latent)
}

# --- Combine and Summarize Latent R² ---
# Total Latent R² is the average explained variance across both axes
r2_total_gelman <- (r2_component_I + r2_component_II) / 2

# Package the Bayesian Credible Intervals
gelman_summary <- data.frame(
  Metric = c("Component I R-squared", "Component II R-squared", "Overall Latent Gelman R-squared"),
  Mean_Percentage = c(mean(r2_component_I), mean(r2_component_II), mean(r2_total_gelman)) * 100,
  Lower_95_CI     = c(quantile(r2_component_I, 0.025), quantile(r2_component_II, 0.025), quantile(r2_total_gelman, 0.025)) * 100,
  Upper_95_CI     = c(quantile(r2_component_I, 0.975), quantile(r2_component_II, 0.975), quantile(r2_total_gelman, 0.975)) * 100
)

print(round(gelman_summary, 2))
```

