---
title: "Circular Analysis"
author: "Elisabetta Canteri"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r, warning=FALSE, message=FALSE, include=FALSE}
library(data.table)
library(ggplot2)
library(FactoMineR)
library(circular)
library(CircStats)
library(bpnreg)
library(BIADconnect)
conn <- init.conn()
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

# DATA

```{r}
#### DATA ####
graves <- fread("/Users/msb290/Library/CloudStorage/OneDrive-UniversityofCopenhagen/Graves/Data/Burial.csv", na.strings = "")
graves$DepositionType <- as.factor(graves$DepositionType)
graves$BodyPositioning <- as.factor(graves$BodyPositioning)
graves$BurialSide <- as.factor(graves$BurialSide)
graves <- graves[order(IndividualID)]
str(graves)

# Sites table (BIAD)
sites <- setDT(query.database("SELECT * FROM Sites", conn))

# Culture and Period
culture <- fread("/Users/msb290/Library/CloudStorage/OneDrive-UniversityofCopenhagen/Graves/Data/burial rites_culture_period.csv", na.strings = "\\N")
culture <- culture[!duplicated(IndividualID)]

## Add info to graves table
graves <- merge(graves, sites[, .(SiteID, Country)], by = "SiteID", all.x = T)
rm(sites)
graves <- merge(graves, culture, by = "IndividualID", all.x = T)

## Remove outlier
graves <- graves[!is.na(YearBP)]
graves <- graves[YearBP < 15000]
summary(graves$YearBP)
hist(graves$YearBP, breaks = 100)

## Create data table
orientation <- graves[, .(Num, BurialOrientationMin, BurialOrientationMax, BurialSide_back, BurialSide_left, BurialSide_right)]
```

```{r}
## We can't just average between the min and max angles because for some angles the average will be 0, so we will not get burials facing south.
## We create a function to transform the angles that works around this issue.

#----------------------------------------------------------------------
find.mid.deg <- function(min.deg, max.deg){
  
  difference <- max.deg - min.deg
  i.accute <- difference<180
  i.obtuse <- difference>180
  i.bad <- difference==180
  
  mid <- numeric()
  mid[i.accute] <- max.deg[i.accute] - difference[i.accute]/2
  mid[i.obtuse] <- max.deg[i.obtuse] + (360 - difference[i.obtuse])/2
  mid[i.bad] <- NA
  
  i <- mid>=360
  mid[i] <- mid[i]-360
  
  i <- mid<0
  mid[i] <- mid[i]+360           
  return(mid)}
#----------------------------------------------------------------------
```

```{r}
# Remove NAs in both min and max orientation columns
orientation <- orientation[!(is.na(BurialOrientationMin) & is.na(BurialOrientationMax))]

# Create a table for records that have a single angle value (in BurialOrientationMin)
single_angles <- orientation[is.na(BurialOrientationMax)]
single_angles[, BurialOrientationMean := BurialOrientationMin]

# Remove from overall table
orientation <- orientation[!(is.na(BurialOrientationMin) | is.na(BurialOrientationMax))]

# Average across min and max 
## Using Adrian's function
orientation[, BurialOrientationMean := find.mid.deg(BurialOrientationMin, BurialOrientationMax)]

## Using circular package
means <- round(sapply(1:nrow(orientation), function(x) mean.circular(circular(c(orientation$BurialOrientationMin[x], orientation$BurialOrientationMax[x]), type = "angles", units = "degrees", rotation = "clock"))),2)
means <- sapply(1:length(means), function(x) ifelse(means[x] < 0, 360 + means[x], means[x]))

## Check differences
orientation$BurialOrientationMin[698]; orientation$BurialOrientationMax[698]
orientation$BurialOrientationMean[698]; means[698]

## In this example, we see that the mean.circular function takes into account that 270 is the min angle. Do we think this is the correct way to go?

orientation[, BurialOrientationMean := means]

# Bind tables together
orientation <- rbind(orientation, single_angles)
orientation <- orientation[order(Num)]

# Calculate cosine and sine from radians
orientation[, BurialOrientationMean_cos := cos(BurialOrientationMean*pi/180)]
orientation[, BurialOrientationMin_cos := cos(BurialOrientationMin*pi/180)]
orientation[, BurialOrientationMax_cos := cos(BurialOrientationMax*pi/180)]
orientation[, BurialOrientationMean_sin := sin(BurialOrientationMean*pi/180)]
orientation[, BurialOrientationMin_sin := sin(BurialOrientationMin*pi/180)]
orientation[, BurialOrientationMax_sin := sin(BurialOrientationMax*pi/180)]

# Add Country, Culture, Period and Ancestry info
orientation$Country <- graves[Num %in% orientation$Num]$Country
orientation$Period <- graves[Num %in% orientation$Num]$Period
orientation$Culture <- graves[Num %in% orientation$Num]$Culture
orientation$Sex <- graves[Num %in% orientation$Num]$Sex
orientation$Age <- graves[Num %in% orientation$Num]$gaussianModelMu
orientation$EHG <- graves[Num %in% orientation$Num]$ANCE2_v2
orientation$WHG <- graves[Num %in% orientation$Num]$ANCE4_v2
orientation$LVN <- graves[Num %in% orientation$Num]$ANCE6_v2
orientation$CHG <- graves[Num %in% orientation$Num]$ANCE8_v2
orientation$mobility <- graves[Num %in% orientation$Num]$mobility_MDS_250y_retrospective

```

# TEST STATISTICS WITH BURIAL SIDES

```{r}
#### CIRCULAR MEAN AND VARIANCE ####
## Separate data based on burial side
left <- orientation[, .(Num, BurialOrientationMean, BurialOrientationMin, BurialOrientationMax), by = "BurialSide_left"][BurialSide_left == 1]
left[, BurialSide := "Left"]
right <- orientation[, .(Num, BurialOrientationMean, BurialOrientationMin, BurialOrientationMax), by = "BurialSide_right"][BurialSide_right == 1]
right[, BurialSide := "Right"]
back <- orientation[, .(Num, BurialOrientationMean, BurialOrientationMin, BurialOrientationMax), by = "BurialSide_back"][BurialSide_back == 1]
back[, BurialSide := "Back"]

# Bind together in a new table
sides <- rbind(left[, -1], right[, -1], back[, -1])

# Example data in degrees
sides <- sides[!is.na(BurialOrientationMean)]
left_angles <- sides[BurialSide == "Left", BurialOrientationMean]
right_angles <- sides[BurialSide == "Right", BurialOrientationMean]
back_angles <- sides[BurialSide == "Back", BurialOrientationMean]

# Convert to circular objects
left_circ <- circular(left_angles, units = "degrees", type = "angles", rotation = "clock")
right_circ <- circular(right_angles, units = "degrees", type = "angles", rotation = "clock")
back_circ <- circular(back_angles, units = "degrees", type = "angles", rotation = "clock")

# Mean direction
mean_left <- mean.circular(left_circ)
mean_right <- mean.circular(right_circ)
mean_back <- mean.circular(back_circ)
as.numeric(deg(mean_left))

# Variance direction
var_left <- var.circular(left_circ)
var_right <- var.circular(right_circ)
var_back <- var.circular(back_circ)

# Resultant length (measure of concentration)
rho_left <- rho.circular(left_circ)
rho_right <- rho.circular(right_circ)
rho_back <- rho.circular(back_circ)

as.numeric(deg(mean_left)); as.numeric(deg(mean_right)); as.numeric(deg(mean_back))
as.numeric(deg(var_left)); as.numeric(deg(var_right)); as.numeric(deg(var_back))
as.numeric(deg(rho_left)); as.numeric(deg(rho_right)); as.numeric(deg(rho_back))
```

```{r}
# Function to calculate smallest angular distance
angular_distance <- function(a, b) {
  diff <- abs(a - b) %% 360
  return(min(diff, 360 - diff))
}
# Example
angular_distance(350, 10)
angular_distance(190, 0)

# Calculate distances
dist_left_right <- angular_distance(as.numeric(mean_left), as.numeric(mean_right))
dist_left_back <- angular_distance(as.numeric(mean_left), as.numeric(mean_back))
dist_right_back <- angular_distance(as.numeric(mean_right), as.numeric(mean_back))

print(dist_left_right)
print(dist_left_back)
print(dist_right_back)
```

```{r}
## Circular means
sides_means <- c(mean_left, mean_right, mean_back)
names(sides_means) <- c("Left", "Right", "Back")
d <- list()
for(a in 1:length(sides_means)){
  d[[a]] <- sapply(1:length(sides_means), function(x){
    angular_distance(sides_means[a], sides_means[x])
  })
}
sides_mat <- do.call(cbind, d)
colnames(sides_mat) <- rownames(sides_mat) <- names(sides_means)
sides_mat[lower.tri(sides_mat)] <- NA
sides_dists <- melt(sides_mat)

# Plot burial orientations per country
sides_means_dt <- data.table(Sides = names(sides_means), Mean = sides_means)
sides_means_dt[, Mean := ifelse(Mean < 0, Mean + 360, Mean)]
ggplot(data = sides, aes(x = BurialSide, y = BurialOrientationMean, fill = BurialSide)) +
  geom_violin(width = 1, drop = F) +
  geom_point(data = sides_means_dt, aes(x = Sides, y = Mean), size = 2, shape = 23, fill = "white") +
  scale_fill_manual(values = pals::brewer.set2(3)) +
  ylab("Burial orientation") +
  theme_light() +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Plot pairwise angular distances
ggplot(data = sides_dists[!is.na(sides_dists$value),]) +
  geom_tile(aes(Var1, Var2, fill = value), color = "black") +
  scale_fill_gradientn("Angular distance", colours = rev(pals::brewer.rdylbu(100)), 
                       na.value = "transparent",
                       limits = c(0,180)) + # maximum distance between angles?
                       # guide = guide_colorbar(barwidth = 1, barheight = 15)) +
  geom_text(aes(Var1, Var2, label = round(value, 1)), size = 10, size.unit = "pt") +
  scale_x_discrete(position = "top") +
  ggtitle("Pairwise distance between angles means") +
  coord_equal(expand = F) +
  theme_classic() +
  theme(plot.title = element_text(face = "bold", size = 16),
        axis.text.x = element_text(angle = 45, hjust = 0),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "none",
        axis.text = element_text(size = 14))
```

Shuffle the values of angles for each category and compute the differences in circular means and do a histogram where we plot the differences in circular means against the true difference. Do it for each of the three pairs. Make sure that each group has the same amount of observations. 

```{r}
n <- min(c(length(left_circ), length(right_circ), length(back_circ)))

diffs <- lapply(1:1000, function(x){
  l <- sample(left_circ, n)
  r <- sample(right_circ, n)
  b <- sample(back_circ, n)
  mean_l <- mean.circular(l)
  mean_r <- mean.circular(r)
  mean_b <- mean.circular(b)
  dist_lr <- angular_distance(as.numeric(mean_l), as.numeric(mean_r))
  dist_lb <- angular_distance(as.numeric(mean_l), as.numeric(mean_b))
  dist_br <- angular_distance(as.numeric(mean_b), as.numeric(mean_r))
  return(list(dist_lr, dist_lb, dist_br))
})

diffs_lr <- unlist(sapply(diffs, "[", 1))
diffs_lb <- unlist(sapply(diffs, "[", 2))
diffs_br <- unlist(sapply(diffs, "[", 3))
diffs <- list(diffs_lr, diffs_lb, diffs_br)

## Plot histograms
hist(diffs_lr, breaks = 100, main = "Difference between left and right orientations", xlab = "Angular distance"); abline(v = dist_left_right, col= "red", lwd = 3)

hist(diffs_lb, breaks = 100, main = "Difference between left and back orientations", xlab = "Angular distance"); abline(v = dist_left_back, col= "red", lwd = 3)

hist(diffs_br, breaks = 100, main = "Difference between back and right orientations", xlab = "Angular distance"); abline(v = dist_right_back, col= "red", lwd = 3)

## Calculate p-value
# P = (X + 1) / (N + 1)
obs <- list(dist_left_right, dist_left_back, dist_right_back)
pval <- lapply(1:length(diffs), function(x){
  X <- sum(abs(diffs[[x]]) >= abs(obs[[x]]))
  N <- length(diffs[[x]])
  return((X + 1) / (N + 1))
})
names(pval) <- c("left_right", "left_back", "back_right")
pval
```

```{r}
# Similarity index
similarity_index <- function(d) {
  return(1 - (d / 180))
}

sim_left_right <- similarity_index(dist_left_right)
sim_left_back <- similarity_index(dist_left_back)
sim_right_back <- similarity_index(dist_right_back)

print(sim_left_right)
print(sim_left_back)
print(sim_right_back)
```

```{r}
## Watson-Williams test
# Combine data
all_angles <- c(left_angles, right_angles, back_angles) #right_angles, 
group <- factor(c(rep("left", length(left_angles)),
                  rep("right", length(right_angles)),
                  rep("back", length(back_angles))))

## Mardia-Watson-Wheeler test (non-parametric)
mww_result1 <- watson.two.test(left_circ, right_circ)  # pairwise example
mww_result2 <- watson.two.test(left_circ, back_circ)
mww_result3 <- watson.two.test(right_circ, back_circ)
print(mww_result1)
print(mww_result2)
print(mww_result3)

## Watson-Williams test
watson.williams.test(circular(all_angles, units = "degrees", type = "angles", rotation = "clock"), group)

## Circular ANOVA 
aov.circular(circular(all_angles, units = "degrees", type = "angles", rotation = "clock"), group)
```

Now that we have tested different statistics using the left, right, and back angles, we can do the same for other categories, like periods, cultures, and countries.

# COUNTRY
```{r}
orientation$Country <- factor(orientation$Country, levels = c("Netherlands",
                                                              "France",
                                                              "Germany",
                                                              "Denmark",
                                                              "Sweden",
                                                              "Latvia",
                                                              "Poland",
                                                              "Czechia / Czech Republic",
                                                              "Slovakia",
                                                              "Austria",
                                                              "Hungary",
                                                              "Croatia",
                                                              "Romania",
                                                              "Serbia",
                                                              "Bulgaria",
                                                              "Ukraine",
                                                              "Russia",
                                                              "Armenia",
                                                              "Kazakhstan"))

## Circular means
countries <- split(orientation[, .(BurialOrientationMean, Country)], by = "Country")
countries_means <- lapply(countries, function(x){
  mean.circular(circular(x$BurialOrientationMean, units = "degrees", type = "angles", rotation = "clock"))
})
countries_means <- unlist(countries_means)
d <- list()
for(a in 1:length(countries_means)){
  d[[a]] <- sapply(1:length(countries_means), function(x){
    angular_distance(countries_means[a], countries_means[x])
  })
}
countries_mat <- do.call(cbind, d)
colnames(countries_mat) <- rownames(countries_mat) <- names(countries_means)
countries_mat[lower.tri(countries_mat)] <- NA
countries_dists <- melt(countries_mat)

# Plot burial orientations per country
countries_means_dt <- data.table(Country = names(countries_means), Mean = countries_means)
countries_means_dt[, Mean := ifelse(Mean < 0, Mean + 360, Mean)]
ggplot(data = orientation, aes(x = Country, y = BurialOrientationMean, fill = Country)) +
  geom_violin(width = 1.5, drop = F) +
  geom_point(data = countries_means_dt, aes(x = Country, y = Mean), size = 2, shape = 23, fill = "white") +
  scale_fill_manual(values = pals::brewer.set2(19)) +
  ylab("Burial orientation") +
  theme_light() +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Plot pairwise angular distances
ggplot(data = countries_dists[!is.na(countries_dists$value),]) +
  geom_tile(aes(Var1, Var2, fill = value), color = "black") +
  scale_fill_gradientn("Angular distance", colours = rev(pals::brewer.rdylbu(100)), 
                       na.value = "transparent",
                       limits = c(0,180)) + # maximum distance between angles?
                       # guide = guide_colorbar(barwidth = 1, barheight = 15)) +
  geom_text(aes(Var1, Var2, label = round(value, 1)), size = 10, size.unit = "pt") +
  scale_x_discrete(position = "top") +
  ggtitle("Pairwise distance between angles means") +
  coord_equal(expand = F) +
  theme_classic() +
  theme(plot.title = element_text(face = "bold", size = 16),
        axis.text.x = element_text(angle = 45, hjust = 0),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "none",
        axis.text = element_text(size = 14))


## Circular ANOVA 
aov_country <- aov.circular(circular(orientation$BurialOrientationMean, units = "degrees", type = "angles", rotation = "clock"), orientation$Country)
aov_country
aov_country$mu
aov_country$kappa
aov_country$rho
aov_country$p.value
```

# PERIOD
```{r}
orientation$Period <- factor(orientation$Period, levels = c("Mesolithic",
                                                          "Mesolithic/Neolithic",
                                                          "Neolithic",
                                                          "Eneolithic",
                                                          "Eneolithic/Bronze Age",
                                                          "Bronze Age",
                                                          "Bronze Age/Iron Age",
                                                          "Iron Age"))

## Circular means
period <- split(orientation[, .(BurialOrientationMean, Period)], by = "Period")
period_means <- lapply(period, function(x){
  mean.circular(circular(rad(x$BurialOrientationMean), units = "radians", type = "angles", rotation = "clock"))
})
period_means <- unlist(period_means)
d <- list()
for(a in 1:length(period_means)){
  d[[a]] <- sapply(1:length(period_means), function(x){
    angular_distance(period_means[a], period_means[x])
  })
}
period_mat <- do.call(cbind, d)
colnames(period_mat) <- rownames(period_mat) <- names(period_means)
period_mat[lower.tri(period_mat)] <- NA
period_dists <- melt(period_mat)

# Plot burial orientations per period
period_means_dt <- data.table(Period = names(period_means), Mean = period_means)
period_means_dt[, Mean := ifelse(Mean < 0, Mean + 360, Mean)]
ggplot(data = orientation, aes(x = Period, y = BurialOrientationMean, fill = Period)) +
  geom_violin(width = 1, drop = F) +
  geom_point(data = period_means_dt, aes(x = Period, y = Mean), size = 2, shape = 23, fill = "white") +
  scale_fill_manual(values = pals::brewer.set2(8)) +
  ylab("Burial orientation") +
  theme_light() +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Plot pairwise angular distances
ggplot(data = period_dists[!is.na(period_dists$value),]) +
  geom_tile(aes(Var1, Var2, fill = value), color = "black") +
  scale_fill_gradientn("Angular distance", colours = rev(pals::brewer.rdylbu(100)), 
                       na.value = "transparent",
                       limits = c(0, 180)) +
                       # guide = guide_colorbar(barwidth = 1, barheight = 15)) +
  geom_text(aes(Var1, Var2, label = round(value, 1)), size = 10, size.unit = "pt") +
  scale_x_discrete(position = "top") +
  ggtitle("Pairwise distance between angles means") +
  coord_equal(expand = F) +
  theme_classic() +
  theme(plot.title = element_text(face = "bold", size = 16),
        axis.text.x = element_text(angle = 45, hjust = 0),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "none",
        axis.text = element_text(size = 14))


## Circular ANOVA 
aov_period <- aov.circular(circular(orientation$BurialOrientationMean, units = "degrees", type = "angles", rotation = "clock"), orientation$Period)
aov_period
aov_period$mu
aov_period$kappa
aov_period$rho
aov_period$p.value
```

# CULTURE
```{r}
# Plot only the 10 most frequent cultures
culture_keep <- orientation[!is.na(Culture)][, .N, by = "Culture"][order(N, decreasing = T)][, Culture][1:10]

## Circular means
culture <- split(orientation[Culture %in% culture_keep][, .(BurialOrientationMean, Culture)], by = "Culture")
culture_means <- lapply(culture, function(x){
  mean.circular(circular(x$BurialOrientationMean, units = "degrees", type = "angles", rotation = "clock"))
})
culture_means <- unlist(culture_means)
d <- list()
for(a in 1:length(culture_means)){
  d[[a]] <- sapply(1:length(culture_means), function(x){
    angular_distance(culture_means[a], culture_means[x])
  })
}
culture_mat <- do.call(cbind, d)
colnames(culture_mat) <- rownames(culture_mat) <- names(culture_means)
culture_mat[lower.tri(culture_mat)] <- NA
culture_dists <- melt(culture_mat)

# Plot burial orientations per period
culture_means_dt <- data.table(Culture = names(culture_means), Mean = culture_means)
culture_means_dt[, Mean := ifelse(Mean < 0, Mean + 360, Mean)]
ggplot(data = orientation[Culture %in% culture_keep], 
       aes(x = Culture, y = BurialOrientationMean, 
           fill = Culture)) +
  geom_violin(width = 1, drop = F) +
  geom_point(data = culture_means_dt, aes(x = Culture, y = Mean), size = 2, shape = 23, fill = "white") +
  scale_fill_manual(values = pals::brewer.set2(10)) +
  ylab("Burial orientation") +
  theme_light() +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Plot pairwise angular distances
ggplot(data = culture_dists[!is.na(culture_dists$value),]) +
  geom_tile(aes(Var1, Var2, fill = value), color = "black") +
  scale_fill_gradientn("Angular distance", colours = rev(pals::brewer.rdylbu(100)), 
                       na.value = "transparent",
                       limits = c(0, 180)) +
                       # guide = guide_colorbar(barwidth = 1, barheight = 15)) +
  geom_text(aes(Var1, Var2, label = round(value, 1)), size = 10, size.unit = "pt") +
  scale_x_discrete(position = "top") +
  ggtitle("Pairwise distance between angles means") +
  coord_equal(expand = F) +
  theme_classic() +
  theme(plot.title = element_text(face = "bold", size = 16),
        axis.text.x = element_text(angle = 45, hjust = 0),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "none",
        axis.text = element_text(size = 14))


## Circular ANOVA 
aov_culture <- aov.circular(circular(orientation[Culture %in% culture_keep]$BurialOrientationMean, units = "degrees", type = "angles", rotation = "clock"), orientation[Culture %in% culture_keep]$Culture)
aov_culture
aov_culture$mu
aov_culture$kappa
aov_culture$rho
aov_culture$p.value
```

# SEX
```{r}
## Circular means
sex <- split(orientation[, .(BurialOrientationMean, Sex = as.factor(Sex))], by = "Sex")
sex_means <- lapply(sex, function(x){
  mean.circular(circular(x$BurialOrientationMean, units = "degrees", type = "angles", rotation = "clock"))
})
sex_means <- unlist(sex_means)
d <- list()
for(a in 1:length(sex_means)){
  d[[a]] <- sapply(1:length(sex_means), function(x){
    angular_distance(sex_means[a], sex_means[x])
  })
}
sex_mat <- do.call(cbind, d)
colnames(sex_mat) <- rownames(sex_mat) <- paste0("Prob_", names(sex_means))
sex_mat[lower.tri(sex_mat)] <- NA
sex_mat
sex_dists <- reshape2::melt(sex_mat)

# Plot burial orientations per period
sex_means_dt <- data.table(Sex = names(sex_means), Mean = sex_means)
sex_means_dt[, Mean := ifelse(Mean < 0, Mean + 360, Mean)]
ggplot(data = orientation, 
       aes(x = as.factor(Sex), y = BurialOrientationMean, 
           fill = as.factor(Sex))) +
  geom_violin(width = 1, drop = F) +
  geom_point(data = sex_means_dt, aes(x = Sex, y = Mean), size = 2, shape = 23, fill = "white") +
  scale_fill_manual(values = pals::brewer.set2(10)) +
  ylab("Burial orientation") +
  theme_light() +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Plot pairwise angular distances
ggplot(data = sex_dists[!is.na(sex_dists$value),]) +
  geom_tile(aes(Var1, Var2, fill = value), color = "black") +
  scale_fill_gradientn("Angular distance", colours = rev(pals::brewer.rdylbu(100)), 
                       na.value = "transparent",
                       limits = c(0, 180)) +
                       # guide = guide_colorbar(barwidth = 1, barheight = 15)) +
  geom_text(aes(Var1, Var2, label = round(value, 1)), 
            size = 10, size.unit = "pt") +
  scale_x_discrete(position = "top") +
  ggtitle("Pairwise distance between angles means") +
  coord_equal(expand = F) +
  theme_classic() +
  theme(plot.title = element_text(face = "bold", size = 16),
        axis.text.x = element_text(angle = 45, hjust = 0),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "none",
        axis.text = element_text(size = 14))


## Circular ANOVA 
aov_sex <- aov.circular(circular(orientation$BurialOrientationMean, units = "degrees", type = "angles", rotation = "clock"), orientation[, as.factor(Sex)])
aov_sex
aov_sex$mu
aov_sex$kappa
aov_sex$rho
aov_sex$p.value

# Only using sex probabilities (0, 0.5, 1)
aov_sex2 <- aov.circular(circular(orientation[Sex %in% c(0, 0.5, 1)]$BurialOrientationMean, units = "degrees", type = "angles", rotation = "clock"), orientation[Sex %in% c(0, 0.5, 1)][, as.factor(Sex)])
aov_sex2
aov_sex2$mu
aov_sex2$kappa
aov_sex2$rho
aov_sex2$p.value
```

# CIRCULAR LINEAR REGRESSIONS WITH ANCESTRY
## WHG
```{r}
y <- rad(circular(orientation[!is.na(WHG)]$BurialOrientationMean, units = "degrees", type = "angles", rotation = "clock"))
x <- cbind(orientation[!is.na(WHG)]$WHG, rep(1,length(orientation[!is.na(WHG)]$WHG)))
colnames(x) <- c("WHG", "Intercept")
reg <- lm.circular(y = y, x = x, type = "c-l", init = c(0,0))
summary(reg)
```

