---
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r, warning=FALSE, message=FALSE, include=FALSE}
library(data.table)
library(ggplot2)
library(FactoMineR)
library(circular)
library(CircStats)
library(bpnreg)
library(BIADconnect)
conn <- init.conn()
knitr::opts_chunk$set(warning = FALSE, message = FALSE,
                      root.dir = "~/Library/CloudStorage/OneDrive-UniversityofCopenhagen/Graves/")
```

# Circular statistics

```{r stats data}
#### DATA ####
orientation <- fread("./Data/BurialOrientation.csv")
```

## Test: circular statistics with burial sides

### Mean and variance

```{r stats test}
#### CIRCULAR MEAN AND VARIANCE ####
## Separate data based on burial side
left <- orientation[, .(Num, BurialOrientationMean, BurialOrientationMin, BurialOrientationMax), by = "BurialSide_left"][BurialSide_left == 1]
left[, BurialSide := "Left"]
right <- orientation[, .(Num, BurialOrientationMean, BurialOrientationMin, BurialOrientationMax), by = "BurialSide_right"][BurialSide_right == 1]
right[, BurialSide := "Right"]
back <- orientation[, .(Num, BurialOrientationMean, BurialOrientationMin, BurialOrientationMax), by = "BurialSide_back"][BurialSide_back == 1]
back[, BurialSide := "Back"]

# Bind together in a new table
sides <- rbind(left[, -1], right[, -1], back[, -1])

# Example data in degrees
sides <- sides[!is.na(BurialOrientationMean)]
left_angles <- sides[BurialSide == "Left", BurialOrientationMean]
right_angles <- sides[BurialSide == "Right", BurialOrientationMean]
back_angles <- sides[BurialSide == "Back", BurialOrientationMean]

# Convert to circular objects
left_circ <- circular(left_angles, units = "degrees", type = "angles", rotation = "clock")
right_circ <- circular(right_angles, units = "degrees", type = "angles", rotation = "clock")
back_circ <- circular(back_angles, units = "degrees", type = "angles", rotation = "clock")

# Mean direction
mean_left <- mean.circular(left_circ)
mean_right <- mean.circular(right_circ)
mean_back <- mean.circular(back_circ)

# Variance direction
var_left <- var.circular(left_circ)
var_right <- var.circular(right_circ)
var_back <- var.circular(back_circ)

# Resultant length (measure of concentration)
rho_left <- rho.circular(left_circ)
rho_right <- rho.circular(right_circ)
rho_back <- rho.circular(back_circ)

## Table
stats <- data.table(Mean = c(mean_left, mean_right, mean_back),
                    Variance = c(var_left, var_right, var_back),
                    Rho = c(rho_left, rho_right, rho_back),
                    Side = c("Left", "Right", "Back"))
stats
```

### Angular distance

Here we create a function to calculate the distance between two angles, taking into account that angles can cross zero.

```{r stats angdist}
# Function to calculate smallest angular distance
angular_distance <- function(a, b) {
  diff <- abs(a - b) %% 360
  return(min(diff, 360 - diff))
}
# Example
angular_distance(350, 10)
angular_distance(190, 0)

# Calculate distances
dist_left_right <- angular_distance(as.numeric(mean_left), as.numeric(mean_right))
dist_left_back <- angular_distance(as.numeric(mean_left), as.numeric(mean_back))
dist_right_back <- angular_distance(as.numeric(mean_right), as.numeric(mean_back))

print(dist_left_right)
print(dist_left_back)
print(dist_right_back)
```

### Visualization

```{r stats viz}
## Circular means
sides_means <- c(mean_left, mean_right, mean_back)
names(sides_means) <- c("Left", "Right", "Back")
d <- list()
for(a in 1:length(sides_means)){
  d[[a]] <- sapply(1:length(sides_means), function(x){
    angular_distance(sides_means[a], sides_means[x])
  })
}
sides_mat <- do.call(cbind, d)
colnames(sides_mat) <- rownames(sides_mat) <- names(sides_means)
sides_mat[lower.tri(sides_mat)] <- NA
sides_dists <- melt(sides_mat)

# Plot burial orientations per country
sides_means_dt <- data.table(Sides = names(sides_means), Mean = sides_means)
sides_means_dt[, Mean := ifelse(Mean < 0, Mean + 360, Mean)]
ggplot(data = sides, aes(x = BurialSide, y = BurialOrientationMean, fill = BurialSide)) +
  geom_violin(width = 1, drop = F) +
  geom_point(data = sides_means_dt, aes(x = Sides, y = Mean), size = 2, shape = 23, fill = "white") +
  scale_fill_manual(values = pals::brewer.set2(3)) +
  ylab("Burial orientation") +
  theme_light() +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Plot pairwise angular distances
ggplot(data = sides_dists[!is.na(sides_dists$value),]) +
  geom_tile(aes(Var1, Var2, fill = value), color = "black") +
  scale_fill_gradientn("Angular distance", colours = rev(pals::brewer.rdylbu(100)), 
                       na.value = "transparent",
                       limits = c(0,180)) + # maximum distance between angles?
                       # guide = guide_colorbar(barwidth = 1, barheight = 15)) +
  geom_text(aes(Var1, Var2, label = round(value, 1)), size = 10, size.unit = "pt") +
  scale_x_discrete(position = "top") +
  ggtitle("Pairwise distance between angles means") +
  coord_equal(expand = F) +
  theme_classic() +
  theme(plot.title = element_text(face = "bold", size = 16),
        axis.text.x = element_text(angle = 45, hjust = 0),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "none",
        axis.text = element_text(size = 14))
```

### P-value

Now we shuffle the values of angles for each category and compute the differences in circular means and do a histogram where we plot the differences in circular means against the true difference. Do it for each of the three pairs. Make sure that each group has the same amount of observations.

```{r stats pval}
n <- min(c(length(left_circ), length(right_circ), length(back_circ)))

diffs <- lapply(1:1000, function(x){
  l <- sample(left_circ, n)
  r <- sample(right_circ, n)
  b <- sample(back_circ, n)
  mean_l <- mean.circular(l)
  mean_r <- mean.circular(r)
  mean_b <- mean.circular(b)
  dist_lr <- angular_distance(as.numeric(mean_l), as.numeric(mean_r))
  dist_lb <- angular_distance(as.numeric(mean_l), as.numeric(mean_b))
  dist_br <- angular_distance(as.numeric(mean_b), as.numeric(mean_r))
  return(list(dist_lr, dist_lb, dist_br))
})

diffs_lr <- unlist(sapply(diffs, "[", 1))
diffs_lb <- unlist(sapply(diffs, "[", 2))
diffs_br <- unlist(sapply(diffs, "[", 3))
diffs <- list(diffs_lr, diffs_lb, diffs_br)

## Plot histograms
hist(diffs_lr, breaks = 100, main = "Difference between left and right orientations", xlab = "Angular distance"); abline(v = dist_left_right, col= "red", lwd = 3)

hist(diffs_lb, breaks = 100, main = "Difference between left and back orientations", xlab = "Angular distance"); abline(v = dist_left_back, col= "red", lwd = 3)

hist(diffs_br, breaks = 100, main = "Difference between back and right orientations", xlab = "Angular distance"); abline(v = dist_right_back, col= "red", lwd = 3)

## Calculate p-value
# P = (X + 1) / (N + 1)
obs <- list(dist_left_right, dist_left_back, dist_right_back)
pval <- lapply(1:length(diffs), function(x){
  X <- sum(abs(diffs[[x]]) >= abs(obs[[x]]))
  N <- length(diffs[[x]])
  return((X + 1) / (N + 1))
})
names(pval) <- c("left_right", "left_back", "back_right")
pval
```

### Similarity

We can also calculate a similarity index that takes into account the distance between the means of the different categories.

```{r stats similarity}
# Similarity index
similarity_index <- function(d) {
  return(1 - (d / 180))
}

sim_left_right <- similarity_index(dist_left_right)
sim_left_back <- similarity_index(dist_left_back)
sim_right_back <- similarity_index(dist_right_back)

print(sim_left_right)
print(sim_left_back)
print(sim_right_back)
```

### Watson-William test and circular ANOVA

```{r stats anova}
## Watson-Williams test
# Combine data
all_angles <- c(left_angles, right_angles, back_angles) #right_angles, 
group <- factor(c(rep("left", length(left_angles)),
                  rep("right", length(right_angles)),
                  rep("back", length(back_angles))))

## Mardia-Watson-Wheeler test (non-parametric)
mww_result1 <- watson.two.test(left_circ, right_circ)  # pairwise example
mww_result2 <- watson.two.test(left_circ, back_circ)
mww_result3 <- watson.two.test(right_circ, back_circ)
print(mww_result1)
print(mww_result2)
print(mww_result3)

## Watson-Williams test
watson.williams.test(circular(all_angles, units = "degrees", type = "angles", rotation = "clock"), group)

## Circular ANOVA 
aov.circular(circular(all_angles, units = "degrees", type = "angles", rotation = "clock"), group)
```

Now that we have tested different statistics using the left, right, and back angles, we can do the same for other categories, like periods, cultures, and countries.

## Country

```{r stats country, fig.dim=c(12,10)}
## Circular means
countries <- split(orientation[!(Country %in% c("Armenia", "Kazakhstan"))][, .(BurialOrientationMean, Country)], by = "Country")
countries_means <- lapply(countries, function(x){
  mean.circular(circular(x$BurialOrientationMean, units = "degrees", type = "angles", rotation = "clock"))
})
countries_means <- unlist(countries_means)
countries_means <- countries_means[gtools::mixedsort(names(countries_means))]

d <- list()
for(a in 1:length(countries_means)){
  d[[a]] <- sapply(1:length(countries_means), function(x){
    angular_distance(countries_means[a], countries_means[x])
  })
}
countries_mat <- do.call(cbind, d)
colnames(countries_mat) <- rownames(countries_mat) <- names(countries_means)
countries_mat[lower.tri(countries_mat)] <- NA
countries_dists <- setDT(reshape2::melt(countries_mat))

# Plot burial orientations per country
countries_means_dt <- data.table(Country = names(countries_means), Mean = countries_means)
countries_means_dt[, Mean := ifelse(Mean < 0, Mean + 360, Mean)]

ggplot(data = orientation[!(Country %in% c("Armenia", "Kazakhstan"))], aes(x = Country, y = BurialOrientationMean, fill = Country)) +
  geom_violin(width = 1.5, drop = F) +
  geom_point(data = countries_means_dt, aes(x = Country, y = Mean), size = 2, shape = 23, fill = "white") +
  scale_fill_manual(values = pals::brewer.set2(19)) +
  ylab("Burial orientation") +
  theme_light() +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Plot pairwise angular distances
countries_dists <- countries_dists[!is.na(countries_dists$value),]

ggplot(data = countries_dists) +
  geom_tile(aes(Var1, Var2, fill = value), color = "black") +
  scale_fill_gradientn("Angular distance", colours = rev(pals::brewer.rdylbu(100)), 
                       na.value = "transparent",
                       limits = c(0,180)) + # maximum distance between angles?
  geom_text(aes(as.factor(Var1), as.factor(Var2), label = round(value, 1)), size = 10, size.unit = "pt") +
  scale_x_discrete(position = "top") +
  scale_y_discrete(position = "left") +
  ggtitle("Pairwise distance between angles means") +
  coord_equal(expand = F) +
  theme_classic() +
  theme(plot.title = element_text(face = "bold", size = 16),
        axis.text.x = element_text(angle = 45, hjust = 0),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "none",
        axis.text = element_text(size = 14),
        plot.margin = margin(l = 0, t = 10, b = 10, r = 0))


## Circular ANOVA 
aov_country <- aov.circular(circular(orientation$BurialOrientationMean, units = "degrees", type = "angles", rotation = "clock"), orientation$Country)
aov_country
aov_country$mu
aov_country$kappa
aov_country$rho
aov_country$p.value
```

## Period

```{r stats period, fig.dim=c(10, 8)}
period_order <- c("Mesolithic",
                  "Mesolithic/Neolithic",
                  "Neolithic",
                  "Eneolithic",
                  "Eneolithic/Bronze Age",
                  "Bronze Age",
                  "Bronze Age/Iron Age",
                  "Iron Age")
orientation$Period <- factor(orientation$Period,
                             levels = period_order)

## Circular means
period <- split(orientation[Period != "Iron Age"]
                [, .(BurialOrientationMean, Period)], 
                by = "Period")
period_means <- lapply(period, function(x){
  mean.circular(circular(x$BurialOrientationMean, 
                         units = "degrees", 
                         type = "angles", 
                         rotation = "clock"))
})
period_means <- unlist(period_means)
period_means <- period_means[period_order]
d <- list()
for(a in 1:length(period_means)){
  d[[a]] <- sapply(1:length(period_means), function(x){
    angular_distance(period_means[a], period_means[x])
  })
}
period_mat <- do.call(cbind, d)
colnames(period_mat) <- rownames(period_mat) <- names(period_means)
period_mat[lower.tri(period_mat)] <- NA
period_dists <- setDT(reshape2::melt(period_mat))

# Plot burial orientations per period
period_means_dt <- data.table(Period = names(period_means), Mean = period_means)
period_means_dt[, Mean := ifelse(Mean < 0, Mean + 360, Mean)]

ggplot(data = orientation[Period != "Iron Age"], aes(x = Period, y = BurialOrientationMean, fill = Period)) +
  geom_violin(width = 1, drop = F) +
  geom_point(data = period_means_dt[Period != "Iron Age"], aes(x = Period, y = Mean), size = 2, shape = 23, fill = "white") +
  scale_fill_manual(values = pals::brewer.set2(8)) +
  ylab("Burial orientation") +
  theme_light() +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Plot pairwise angular distances
ggplot(data = period_dists[!is.na(period_dists$value),]) +
  geom_tile(aes(Var1, Var2, fill = value), color = "black") +
  scale_fill_gradientn("Angular distance", 
                       colours = rev(pals::brewer.rdylbu(100)), 
                       na.value = "transparent",
                       limits = c(0, 180)) +
  geom_text(aes(Var1, Var2, label = round(value, 1)), 
            size = 10, size.unit = "pt") +
  scale_x_discrete(position = "top") +
  ggtitle("Pairwise distance between angles means") +
  coord_equal(expand = F) +
  theme_classic() +
  theme(plot.title = element_text(face = "bold", size = 16),
        axis.text.x = element_text(angle = 45, hjust = 0),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "none",
        axis.text = element_text(size = 14))


## Circular ANOVA 
aov_period <- aov.circular(circular(orientation$BurialOrientationMean, units = "degrees", type = "angles", rotation = "clock"), orientation$Period)
aov_period
aov_period$mu
aov_period$kappa
aov_period$rho
aov_period$p.value
```

## Culture

```{r stats culture, fig.dim=c(15,10)}
# Plot only the 10 most frequent cultures
culture_keep <- orientation[!is.na(Culture)][, .N, by = "Culture"][order(N, decreasing = T)][, Culture][1:10]

## Circular means
culture <- split(orientation[Culture %in% culture_keep]
                 [, .(BurialOrientationMean, Culture)], 
                 by = "Culture")
culture_means <- lapply(culture, function(x){
  mean.circular(circular(x$BurialOrientationMean, 
                         units = "degrees", 
                         type = "angles", 
                         rotation = "clock"),
                na.rm = T)
})
culture_means <- unlist(culture_means)
d <- list()
for(a in 1:length(culture_means)){
  d[[a]] <- sapply(1:length(culture_means), function(x){
    angular_distance(culture_means[a], culture_means[x])
  })
}
culture_mat <- do.call(cbind, d)
colnames(culture_mat) <- rownames(culture_mat) <- names(culture_means)
culture_mat[lower.tri(culture_mat)] <- NA
culture_dists <- setDT(reshape2::melt(culture_mat))

# Plot burial orientations per period
culture_means_dt <- data.table(Culture = names(culture_means), 
                               Mean = culture_means)
culture_means_dt[, Mean := ifelse(Mean < 0, Mean + 360, Mean)]
culture_means_dt
ggplot(data = orientation[Culture %in% culture_keep], 
       aes(x = Culture, y = BurialOrientationMean, 
           fill = Culture)) +
  geom_violin(width = 1, drop = F) +
  geom_point(data = culture_means_dt, aes(x = Culture, y = Mean), 
             size = 2, shape = 23, fill = "white") +
  scale_fill_manual(values = pals::brewer.set2(10)) +
  ylab("Burial orientation") +
  theme_light() +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# Plot pairwise angular distances
ggplot(data = culture_dists[!is.na(culture_dists$value),]) +
  geom_tile(aes(Var1, Var2, fill = value), color = "black") +
  scale_fill_gradientn("Angular distance", 
                       colours = rev(pals::brewer.rdylbu(100)), 
                       na.value = "transparent",
                       limits = c(0, 180)) +
  geom_text(aes(Var1, Var2, label = round(value, 1)), 
            size = 10, size.unit = "pt") +
  scale_x_discrete(position = "top") +
  ggtitle("Pairwise distance between angles means") +
  coord_equal(expand = F) +
  theme_classic() +
  theme(plot.title = element_text(face = "bold", size = 16),
        axis.text.x = element_text(angle = 45, hjust = 0),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "none",
        axis.text = element_text(size = 14))


## Circular ANOVA 
aov_culture <- aov.circular(circular(orientation[Culture %in% culture_keep]$BurialOrientationMean, units = "degrees", type = "angles", rotation = "clock"), orientation[Culture %in% culture_keep]$Culture)
aov_culture
aov_culture$mu
aov_culture$kappa
aov_culture$rho
aov_culture$p.value
```

### Difference between Corded Ware and Bell Beaker cultures

We are going to fold the angles along the horizontal axis (E-W) so that and angle of 135 degrees is equal to 45 degrees and an angle of 225 degrees is equal to 315 degrees (-45 degrees).

An angle of 270 degrees is now -90 degrees.

```{r stats cw-bb}
## Fold angles on the horizontal axis
cordedware <- culture[["Corded Ware/Single Grave/Battle Axe/Fatyanovo"]]$BurialOrientationMean
cordedware <- ifelse(between(cordedware, 90, 269), 180-cordedware, ifelse(between(cordedware, 270, 360), -(360-cordedware), cordedware))
cordedware <- circular(cordedware,
                      type = "angles",
                      units = "degrees",
                      rotation = "clock")
bellbeaker <- culture[["Bell Beaker complex"]]$BurialOrientationMean
bellbeaker <- ifelse(between(bellbeaker, 90, 269), 180-bellbeaker, ifelse(between(bellbeaker, 270, 360), -(360-bellbeaker), bellbeaker))
bellbeaker <- circular(bellbeaker,
                      type = "angles",
                      units = "degrees",
                      rotation = "clock")

# Shuffle angles
n <- min(c(length(cordedware), length(bellbeaker)))
dists <- sapply(1:1000, function(x){
  cw <- sample(cordedware, n)
  bb <- sample(bellbeaker, n)
  mean_cw <- mean.circular(cw)
  mean_bb <- mean.circular(bb)
  dist_cwbb <- angular_distance(as.numeric(mean_cw), as.numeric(mean_bb))
  return(dist_cwbb)
})

## Plot histograms
true_dist <- angular_distance(as.numeric(mean.circular(cordedware)), as.numeric(mean.circular(bellbeaker)))
hist(dists, breaks = 100, main = "Angular distances between Corded Ware and Bell Beaker orientations", xlab = "Angular distance"); abline(v = true_dist, col= "red", lwd = 3)

## Calculate p-value
# P = (X + 1) / (N + 1)
X <- sum(abs(dists) >= abs(true_dist))
N <- length(dists)
pval <- (X + 1) / (N + 1)
pval

# Similarity index (without folding over the horizontal axis)
obs <- culture_dists[Var2 == "Corded Ware/Single Grave/Battle Axe/Fatyanovo" & Var1 == "Bell Beaker complex"]$value
diff <- similarity_index(obs)
diff
paste0((1-round(diff,2))*100, "% different")
```

**THE P-VALUE TEST HAS ISSUES. CALCULATING THE DISTANCE BETWEEN MEANS DOES NOT GIVE THE DESIRED RESULT. REVIEW WITH FERNANDO HOW TO DO IT**

## Sex

```{r stats sex}
# Group sex
orientation[, Sex := ifelse(Sex < 0.5, 0, ifelse(Sex > 0.5, 1, Sex))]

## Circular means
sex <- split(orientation[, .(BurialOrientationMean, Sex = as.factor(Sex))], by = "Sex")
sex_means <- lapply(sex, function(x){
  mean.circular(circular(x$BurialOrientationMean, 
                         units = "degrees", 
                         type = "angles", 
                         rotation = "clock"))
})
sex_means <- unlist(sex_means)
sex_means <- sex_means[gtools::mixedsort(names(sex_means))]
d <- list()
for(a in 1:length(sex_means)){
  d[[a]] <- sapply(1:length(sex_means), function(x){
    angular_distance(sex_means[a], sex_means[x])
  })
}
sex_mat <- do.call(cbind, d)
colnames(sex_mat) <- rownames(sex_mat) <- paste0("Prob_", names(sex_means))
sex_mat[lower.tri(sex_mat)] <- NA
sex_mat
sex_dists <- reshape2::melt(sex_mat)

# Plot burial orientations per period
sex_means_dt <- data.table(Sex = names(sex_means), Mean = sex_means)
sex_means_dt[, Mean := ifelse(Mean < 0, Mean + 360, Mean)]
ggplot(data = orientation, 
       aes(x = as.factor(Sex), y = BurialOrientationMean, 
           fill = as.factor(Sex))) +
  geom_violin(width = 1, drop = F) +
  geom_point(data = sex_means_dt, aes(x = Sex, y = Mean), 
             size = 2, shape = 23, fill = "white") +
  scale_fill_manual(values = rev(pals::brewer.piyg(3))) +
  ylab("Burial orientation") +
  theme_light() +
  theme(legend.position = "none",
        axis.title.x = element_blank())

# Plot pairwise angular distances
ggplot(data = sex_dists[!is.na(sex_dists$value),]) +
  geom_tile(aes(Var1, Var2, fill = value), color = "black") +
  scale_fill_gradientn("Angular distance", colours = rev(pals::brewer.rdylbu(100)), 
                       na.value = "transparent",
                       limits = c(0, 180)) +
   geom_text(aes(Var1, Var2, label = round(value, 1)), 
            size = 10, size.unit = "pt") +
  scale_x_discrete(position = "top") +
  ggtitle("Pairwise distance between angles means") +
  coord_equal(expand = F) +
  theme_classic() +
  theme(plot.title = element_text(face = "bold", size = 16),
        axis.text.x = element_text(angle = 45, hjust = 0),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "none",
        axis.text = element_text(size = 14))

# Only using sex probabilities (0, 0.5, 1)
aov_sex <- aov.circular(circular(orientation[Sex %in% c(0, 0.5, 1)]$BurialOrientationMean, units = "degrees", type = "angles", rotation = "clock"), orientation[Sex %in% c(0, 0.5, 1)][, as.factor(Sex)])
aov_sex
aov_sex$mu
aov_sex$kappa
aov_sex$rho
aov_sex$p.value
```

## Circular linear regression models

For data that is not categorical (e.g., ancestry and mobility values), we have to use a regression model that is specific to angular data.

This regression models both the mean direction and the dispersion simultaneously. The model assumes that a circular response variable $\theta$ has a von Mises distribution with concentration parameter $\kappa$, and mean direction ($\mu$) related to a vector of linear predictor variables according to the relationship: $\mu + 2*atan(\beta*x)$, where $\mu$ and $\beta$ are unknown parameters, $\beta$ being a vector of regression coefficients, and $2*atan()$ being the link function.

The von Mises distribution is a function of $\mu$ and $\kappa$, where $\mu$ is the mean direction and $\kappa$ is the concentration parameter. A value $\kappa = 0$ corresponds to the circular uniform distribution; for $\kappa > 0$, the density is unimodal and symmetrical about $\mu$, and increasingly concentrated as $\kappa$ increases. For $\kappa \ge 2$ the density at the antipode $\mu + \pi$ is effectively zero, and the distribution is well approximated by a normal distribution with variance $1/\kappa$.

We can start with an initial value for $\beta$, to calculate $\mu$ and $\kappa$. These estimates can then be used to update the $\beta$ value, and then to update $\mu$ and $\kappa$, analogously to the ML equations for an independent, identically distributed (i.i.d.) von Mises sample. This cycle is repeated until convergence. If $\beta = 0$, we have a null model where all observations cluster around a common mean direction $\mu$. However, as the elements of $\beta$ become infinite, the use of any monotone link function also leads to a null model with mean direction $\mu ± \pi$. In such cases, the likelihood will have a peak in the vicinity of zero, and attain a similar magnitude as the elements of $\beta$ tend to $± \infty$. **The optimization process is simplified if the x are centered at their means**. 

**So should we scale our ancestry values?** (Whether we order the values or scale them, model results don't seem to change)

Further reads: <https://www.jstor.org/stable/2532334?seq=2>; <http://iopscience.iop.org/article/10.1088/1742-6596/1988/1/012108>

The `circular` package, has a `lm.circular` function that can be used for circular and linear data (c-l). We are going to test this function using the ancestry values associated to the grave individuals.

### CHG

```{r stats chg, fig.dim=c(10,10)}
y <- circular(rad(orientation[!is.na(CHG)]$BurialOrientationMean), units = "radians", type = "angles", rotation = "clock")
x <- cbind(orientation[!is.na(CHG)]$CHG, rep(1,length(orientation[!is.na(CHG)]$CHG)))
colnames(x) <- c("CHG", "Intercept")
mu <- as.numeric(mean.circular(y))
reg.chg <- lm.circular(y = y, x = x, type = "c-l", init = c(0,1))
reg.chg
```

We can see that CHG has a significant positive relationship with the mean direction of the burials. This test was made by using all the data as is. But what if we repeat the analysis by averaging the CHG value of the individuals that have the same burial orientation? Similarly to what we did to plot them.

```{r stats chg2}
## Based on average value across individuals with the same angle
dim_chg <- orientation[, mean(CHG, na.rm = T), by = c("BurialOrientationMean_sin", "BurialOrientationMean_cos")]
dim_chg <- dim_chg[!is.nan(V1),]
names(dim_chg)[3] <- "CHG"
# Get the angle from sin and cos (in radians)
dim_chg[, BurialOrientationMean := sapply(1:nrow(dim_chg), function(x) atan2(BurialOrientationMean_sin[x], BurialOrientationMean_cos[x]))]

# New variables for the linear regression model
y2 <- circular(dim_chg$BurialOrientationMean, units = "radians", type = "angles", rotation = "clock")
x2 <- cbind(dim_chg$CHG, rep(1, nrow(dim_chg)))
colnames(x2) <- c("CHG", "Intercept")
mu <- as.numeric(mean.circular(y2))
m <- mean(x2[,1])

# Model
reg.chg2 <- lm.circular(y = y2, x = x2, type = "c-l", init = c(0,1))
reg.chg2

# Plot
mu.df <- data.frame(mu = c(mu, reg.chg2$mu), type = c("Observed", "Estimated"))

ggplot() +
  geom_point(data = dim_chg,
             aes(BurialOrientationMean_sin, BurialOrientationMean_cos, fill = CHG),
             shape = 21,
             size = 5,
             ) +
  geom_point(data = mu.df, aes(sin(mu), cos(mu), shape = type), size = 8) +
  scale_fill_gradientn("CHG", colors = pals::parula(100)) +
  scale_shape_manual(expression(mu), values = c(7, 8)) +
  xlab(expression(sin(theta))) +
  ylab(expression(cos(theta))) +
  coord_equal() +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16))
```

### WHG

```{r stats whg, fig.dim = c(10,10)}
y <- circular(rad(orientation[!is.na(WHG)]$BurialOrientationMean), units = "radians", type = "angles", rotation = "clock")
x <- cbind(orientation[!is.na(WHG)]$WHG, rep(1,length(orientation[!is.na(WHG)]$WHG)))
colnames(x) <- c("WHG", "Intercept")
mu <- as.numeric(mean.circular(y))
reg.whg <- lm.circular(y = y, x = x, type = "c-l", init = c(0,1))
reg.whg
```

### EHG

```{r stats ehg}
y <- circular(rad(orientation[!is.na(EHG)]$BurialOrientationMean), units = "radians", type = "angles", rotation = "clock")
x <- cbind(orientation[!is.na(EHG)]$EHG, rep(1,length(orientation[!is.na(EHG)]$EHG)))
colnames(x) <- c("EHG", "Intercept")
mu <- as.numeric(mean.circular(y))
reg.ehg <- lm.circular(y = y, x = x, type = "c-l", init = c(0,1))
reg.ehg
```

### LVN

```{r stats lvn}
y <- circular(rad(orientation[!is.na(LVN)]$BurialOrientationMean), units = "radians", type = "angles", rotation = "clock")
x <- cbind(orientation[!is.na(LVN)]$LVN, rep(1,length(orientation[!is.na(LVN)]$LVN)))
colnames(x) <- c("LVN", "Intercept")
mu <- as.numeric(mean.circular(y))
reg.lvn <- lm.circular(y = y, x = x, type = "c-l", init = c(0,1))
reg.lvn
```

### Mobility

```{r stats mobility}
y <- circular(rad(orientation[!is.na(mobility)]$BurialOrientationMean), units = "radians", type = "angles", rotation = "clock")
x <- cbind(orientation[!is.na(mobility)]$mobility, rep(1,length(orientation[!is.na(mobility)]$mobility)))
colnames(x) <- c("Mobility", "Intercept")
mu <- as.numeric(mean.circular(y))
reg.mob <- lm.circular(y = y, x = x, type = "c-l", init = c(0,1))
reg.mob
```



