---
title: "Final models - left, right, back"
author: "Elisabetta Canteri"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

# Introduction
Here we are going to model the **left**, **right**, and **back** burial sides against human ancestry and genetic estimates of mobility. Both the left and the right burial sides are associated with the **crouched** body position, while the back burial side is associated with the **extended** body position.
Genetic inferred values of human ancestry and mobility were paired to the burial individuals by taking the average of up to 4 of the closest individuals within a radius of 350 km and 300 years apart.
Human mobility was recalculated using the methods described in Schmid & Schiffels (2023) https://www.pnas.org/doi/pdf/10.1073/pnas.2218375120 using 250 years as retrospective distance.


```{r, message=FALSE, warning=FALSE, include=FALSE, echo = F}
library(data.table)
library(ggplot2)
library(terra)
library(sf)
library(spatstat)
library(inlabru)
library(INLA)
library(fmesher)
library(pbapply)
library(mobest)
library(knitr)
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, 
                      message = FALSE, cache.lazy = FALSE)
setwd("/maps/projects/racimolab/people/msb290/Graves/Scripts/")
proj <- "+proj=laea +lon_0=0 +lat_0=49.06 +datum=WGS84 +units=km +no_defs"
proj2 <- "+proj=aea +lon_0=44.296875 +lat_1=43.7864128 +lat_2=69.9512657 +lat_0=56.8688392 +datum=WGS84 +units=m +no_defs"
```

# Data
```{r, message=FALSE, warning=FALSE, include = F, echo = F}
## Data
graves <- fread("../Data/Burial.csv", na.strings = "")
graves$DepositionType <- as.factor(graves$DepositionType)
graves$BodyPositioning <- as.factor(graves$BodyPositioning)
graves$BurialSide <- as.factor(graves$BurialSide)

## Coastlines and continents poly
conts <- rnaturalearth::ne_download(scale = 10, type = "coastline", "physical", returnclass = "sf")
land <- rnaturalearth::ne_download(scale = 10, type = "land", "physical", returnclass = "sf")
countries <- rnaturalearth::ne_download(scale = 50, type = "countries", "cultural", returnclass = "sf")
grat <- rnaturalearth::ne_download(scale = 10, type = "graticules_10", "physical", returnclass = "sf")
conts_prj <- st_transform(conts, crs = proj)
land_prj <- st_transform(land, proj)
countries_prj <- st_transform(countries, crs = proj)
grat_prj <- st_transform(grat, proj)
```

```{r data prep, message=FALSE, warning=FALSE}
# Create spatial points from data
pnts <- project(
  vect(
    graves[, .(Num, Longitude, Latitude, YearBP,
               BurialSide_left,
               BurialSide_right,
               BurialSide_back,
               mobility = mobility_MDS_250y_retrospective,
               EHG = ANCE2_v2,
               WHG = ANCE4_v2,
               LVN = ANCE6_v2,
               CHG = ANCE8_v2 
    )],
    geom = c("Longitude", "Latitude"),
    crs = crs(rast())
  ),
  proj
)

poly <- vect("../Spatial/IntersectionVector.shp")
pnts <- intersect(pnts, poly)

# Remove data without age and too old sample
pnts <- st_as_sf(pnts)
pnts <- pnts[!is.na(pnts$YearBP),]
pnts <- pnts[which(pnts$YearBP < 15000),]

# Subset the pnts to remove duplicated sites
coords <- st_coordinates(pnts)[, c("X", "Y")]
coords <- coords[!duplicated(coords),]
pnts_new <- vect(coords, crs = proj)
n <- dim(coords)[1]
```

# Model settings
### Spatial mesh
```{r smesh, message=FALSE, warning=FALSE}
# Define spatial range and parameters - diagonal (euclidean distance) of the bounding box around the points, in the unit of the coordinates.
bbox_coords <- st_bbox(pnts_new)
summary(dist(coords))
sp.range <- sqrt(sum(c(diff(bbox_coords[c(1,3)])^2, diff(bbox_coords[c(2,4)])^2)))

# Boundary
bound <- fmesher::fm_nonconvex_hull_inla(coords, convex = -0.2, crs = proj)

# Spatial mesh
mesh <-  fm_mesh_2d(loc.domain = coords,
                    boundary = bound,
                    max.edge = c(0.03, 0.1)*sp.range,
                    offset = 0.3*sp.range,
                    cutoff = 0.01*sp.range,
                    crs = proj)

# Plot
ggplot() +
  gg(mesh) +
  geom_sf(data = conts_prj, col = "green4") +
  geom_sf(data = st_as_sf(pnts_new), shape = 21, fill = 'blue', size = 1) +
  theme_bw() +
  coord_sf(xlim = range(mesh$loc[,1]), ylim = range(mesh$loc[,2]))

# Create a sampler polygon for the model
cover <- st_read("../Spatial/cover.shp")
st_crs(cover) <- proj2
cover <- st_transform(cover, proj)
```

### Temporal mesh
```{r tmesh, warning=FALSE, message=FALSE}
# Temporal mesh
tknots <- seq(0, 12000, 1000)
mesh.t <- fm_mesh_1d(tknots, boundary = "free")
k <- length(tknots)
```

### Matern
```{r matern, warning=F, message=F}
## Spatial point pattern
pp <- ppp(coords[,1], coords[,2], bbox_coords[c(1,3)], bbox_coords[c(2,4)], unitname = c("kilometers"))

# Cross Validated Bandwidth Selection
bw <- round(bw.diggle(pp))

# Spatial priors
range.s <- c(bw, 0.01)
sigma.s <- c(1, 0.01)

# Matern
matern <- inla.spde2.pcmatern(mesh,
  prior.sigma = sigma.s,
  prior.range = range.s
)

# Prior for temporal auto-regressive parameter
h.spec <- list(theta = list(prior = "pccor1", param = c(0, 0.9)))
```

We start with the burial side left.

# BURIAL SIDE: LEFT
```{r left, warning=F, message=F}
# Data
left <- pnts[, c("YearBP", "BurialSide_left", "mobility", "EHG", "WHG", "LVN", "CHG", "geometry")]
left <- left[!is.na(left$BurialSide_left),]
nrow(left)

# Jitter the duplicates
duplicates <- left[duplicated(left),]
duplicates <- st_jitter(duplicates, 0.05)

plot(conts_prj$geometry, xlim = ext(left)[1:2], ylim = ext(left)[3:4]) 
points(left, pch = 1, cex = 1); points(duplicates, pch = 1, col = 'red', cex = 1)

# Add duplicates to the left data
left <- left[!duplicated(left),]
left <- rbind(left, duplicates)
nrow(left) # same as before

# Divide into time bins
left$YearBin <- plyr:::round_any(left$YearBP, 250)
```

### Null model: spatiotemporal process
```{r left m, warning=F, message=F}
knitr::opts_chunk$set(cache = TRUE)
# Components
cmp <- BurialSide_left ~ Intercept(1) +
  spatemp(geometry,
      model = matern,
      group = YearBin,
      group_mapper = bru_mapper(mesh.t, indexed = TRUE),
      control.group = list(model = "ar1", hyper = h.spec))

# Model
set.seed(1234)
left.m <- bru(cmp,
          data = left,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(left.m)

# Predict
pix <- fm_pixels(mesh, mask = cover)
pred.t <- seq(2000, 10000, 1000)
df <- fm_cprod(pix, data.frame(YearBin = pred.t))

pred <- predict(left.m,
                df,
                ~ 1 / (1 + exp(-(Intercept + spatemp))))

# Plot data
plot_df <- left
plot_df$YearBin <- plyr:::round_any(plot_df$YearBin, 1000)

ggplot(data = cbind(as.data.frame(plot_df), st_coordinates(plot_df)),
       aes(X, Y, shape = factor(BurialSide_left, levels = c(1,0)),
           size = factor(BurialSide_left, levels = c(1,0)),
           color = factor(BurialSide_left, levels = c(1,0)),
           alpha = factor(BurialSide_left, levels = c(1,0)))) +
  geom_sf(data = cover, alpha = 0.1, inherit.aes = F) +
  geom_sf(data = conts_prj, inherit.aes = F) +
  facet_wrap(~-YearBin) +
  geom_point() +
  scale_shape_manual(values = c(8, 19)) +
  scale_size_manual(values = c(2, 1)) +
  scale_alpha_manual(values = c(1, 0.5)) +
  coord_sf(xlim = st_bbox(cover)[c(1,3)], ylim = st_bbox(cover)[c(2,4)]) +
  theme_classic() +
  ggtitle("Left burial side") +
  theme(legend.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = 'bold'))

# Plot model prediction
pred_df <- cbind(as.data.frame(pred), st_coordinates(pred))
g <- pblapply(c("mean", "sd", "median"), function(p){
  pl <- ggplot() +
    geom_tile(data = pred_df, aes_string("X", "Y", fill = p)) +
    geom_sf(data = cover, inherit.aes = F, fill = 'transparent', alpha = 0.1) +
    
    scale_fill_gradientn(colors = rev(pals::brewer.rdylbu(100)),
                         breaks = seq(0, 1, 0.2),
                         limits = c(0,1)) +
    facet_wrap(~-YearBin) +
    ggtitle(label = "Left burial side", subtitle = p) +
    theme_bw() +
    theme(axis.text = element_blank(),
          axis.title = element_blank(),
          axis.ticks = element_blank(),
          legend.title = element_blank())
  return(pl)
})
g[[1]]
g[[2]]
g[[3]]
```

### All covariates
```{r left m1, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(cache = TRUE)

## Define covariates
namescov <- c("mobility", "EHG", "WHG", "LVN", "CHG")
newnames <- paste0(namescov, "_scaled")
covs <- data.frame(left)[, namescov]
covs <- scale(covs)
colnames(covs) <- newnames
left <- cbind(left, covs)

# Components
cmp <- BurialSide_left ~ Intercept(1) +
    mobility_scaled +
    EHG_scaled +
    WHG_scaled +
    LVN_scaled +
    CHG_scaled +
  spt(geometry,
      model = matern,
      group = YearBin,
      group_mapper = bru_mapper(mesh.t, indexed = TRUE),
      control.group = list(model = "ar1", hyper = h.spec))

# Model
set.seed(1234)
left.m1 <- bru(cmp,
          data = left,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(left.m1)
pl <- lapply(1:length(newnames), function(v){
    plot(left.m1, newnames[v]) +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      xlab(newnames[v]) +
      theme_bw() +
      theme(axis.title.y = element_blank())
})
multiplot(plotlist = pl)
```

### Ancestry only
```{r left m2, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(cache = TRUE)

# Components
cmp <- BurialSide_left ~ Intercept(1) +
    EHG_scaled +
    WHG_scaled +
    LVN_scaled +
    CHG_scaled +
  spt(geometry,
      model = matern,
      group = YearBin,
      group_mapper = bru_mapper(mesh.t, indexed = TRUE),
      control.group = list(model = "ar1", hyper = h.spec))

# Model
set.seed(1234)
left.m2 <- bru(cmp,
          data = left,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(left.m2)
pl <- lapply(2:length(newnames), function(v){
    plot(left.m2, newnames[v]) +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("Left burial side") +
      theme_bw()
})
multiplot(plotlist = pl)

# Extract ancestry values
ance <- readRDS("../../MesoNeo/SelectedAncestries_100ystep.RDS")
time <- raster::getZ(ance[[1]])
layers <- which(raster::getZ(ance[[1]]) %in% sort(pred.t*-1))
ance <- pblapply(ance, rast)
ance <- pblapply(ance, project, proj)
ance_pred <- lapply(ance, function(a){
    pblapply(layers, function(r){
        extract(a[[r]], st_coordinates(pix))
    })
})
ance_pred <- lapply(1:length(ance_pred), function(a){
    pblapply(1:length(ance_pred[[a]]), function(d){
        df <- cbind(st_coordinates(pix), ance_pred[[a]][[d]], data.frame(YearBin = sort(pred.t, decreasing = T)[d]))
        colnames(df)[3] <- names(ance_pred[a])
        return(as.data.table(df))
    })
})
ance_pred <- pblapply(ance_pred, rbindlist)

data_pred <- cbind(ance_pred[[1]][, c(1,2,4)], ance_pred[[1]][, 3], ance_pred[[2]][, 3], ance_pred[[3]][,3], ance_pred[[4]][,3])
data_pred <- st_as_sf(vect(data_pred, geom=c("X", "Y"), crs = proj))
colnames(data_pred)[2:5] <- newnames[-1]

# Predict
pred2 <- predict(
    left.m2,
    data_pred,
    ~ 1 / (1 + exp(-(Intercept + EHG_scaled + WHG_scaled + LVN_scaled + CHG_scaled + spt)))
)
pred_df <- cbind(as.data.frame(pred2), st_coordinates(pred2))
g <- pblapply(c("mean", "sd"), function(p){
  pl <- ggplot() +
    geom_tile(data = pred_df, aes_string("X", "Y", fill = p)) +
    geom_sf(data = cover, inherit.aes = F, fill = 'transparent', alpha = 0.1) +
    
    scale_fill_gradientn(colors = rev(pals::brewer.rdylbu(100)),
                         breaks = seq(0, 1, 0.2),
                         limits = c(0,1)) +
    facet_wrap(~-YearBin) +
    ggtitle(label = "Left burial side", subtitle = p) +
    theme_bw() +
    theme(axis.text = element_blank(),
          axis.title = element_blank(),
          axis.ticks = element_blank(),
          legend.title = element_blank())
  return(pl)
})
g[[1]]
```

### Single ancestry: EHG
```{r left EHG, message=F, warning = F}
knitr::opts_chunk$set(cache = TRUE)
# Components
cmp <- BurialSide_left ~ Intercept(1) +
    EHG_scaled +
    spt(geometry,
    model = matern,
    group = YearBin,
    group_mapper = bru_mapper(mesh.t, indexed = TRUE),
    control.group = list(model = "ar1", hyper = h.spec))

# Model
left.EHG <- bru(cmp,
          data = left,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(left.EHG)
plot(left.EHG, "EHG_scaled") +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("Left burial side") +
      theme_bw()
```

### Single ancestry: WHG
```{r left WHG, message=F, warning = F}
knitr::opts_chunk$set(cache = TRUE)
# Components
cmp <- BurialSide_left ~ Intercept(1) +
    WHG_scaled +
    spt(geometry,
    model = matern,
    group = YearBin,
    group_mapper = bru_mapper(mesh.t, indexed = TRUE),
    control.group = list(model = "ar1", hyper = h.spec))

# Model
left.WHG <- bru(cmp,
          data = left,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(left.WHG)
plot(left.WHG, "WHG_scaled") +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("Left burial side") +
      theme_bw()
```

### Single ancestry: CHG
```{r left CHG, message=F, warning = F}
knitr::opts_chunk$set(cache = TRUE)
# Components
cmp <- BurialSide_left ~ Intercept(1) +
    CHG_scaled +
    spt(geometry,
    model = matern,
    group = YearBin,
    group_mapper = bru_mapper(mesh.t, indexed = TRUE),
    control.group = list(model = "ar1", hyper = h.spec))

# Model
left.CHG <- bru(cmp,
          data = left,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(left.CHG)
plot(left.CHG, "CHG_scaled") +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("Left burial side") +
      theme_bw()
```

### Single ancestry: LVN
```{r left LVN, message=F, warning = F}
knitr::opts_chunk$set(cache = TRUE)
# Components
cmp <- BurialSide_left ~ Intercept(1) +
    LVN_scaled +
    spt(geometry,
    model = matern,
    group = YearBin,
    group_mapper = bru_mapper(mesh.t, indexed = TRUE),
    control.group = list(model = "ar1", hyper = h.spec))

# Model
left.LVN <- bru(cmp,
          data = left,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(left.LVN)
plot(left.LVN, "LVN_scaled") +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("Left burial side") +
      theme_bw()
```

### Only mobility
```{r left mobility, warning=F, message=F}
knitr::opts_chunk$set(cache = TRUE)
# Components
cmp <- BurialSide_left ~ Intercept(1) +
  mobility_scaled +
  spt(geometry,
    model = matern,
    group = YearBin,
    group_mapper = bru_mapper(mesh.t, indexed = TRUE),
    control.group = list(model = "ar1", hyper = h.spec)
  )

# Model
left.m3 <- bru(cmp,
          data = left,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(left.m3)
plot(left.m3, "mobility_scaled") +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("Left burial side") +
      theme_bw()
```

### Only ancestry: no WHG
```{r left no WHG, message = F, warning = F}
knitr::opts_chunk$set(cache = TRUE)
# Components
cmp <- BurialSide_left ~ Intercept(1) +
    EHG_scaled +
    LVN_scaled +
    CHG_scaled +
  spt(geometry,
      model = matern,
      group = YearBin,
      group_mapper = bru_mapper(mesh.t, indexed = TRUE),
      control.group = list(model = "ar1", hyper = h.spec))

# Model
left.m4 <- bru(cmp,
          data = left,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(left.m4)
pl <- lapply(newnames[-c(1,3)], function(v){
    plot(left.m4, v) +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("Left burial side") +
      theme_bw()
})
multiplot(plotlist = pl)

# Predict
pred_ance <- predict(
    left.m4,
    data_pred,
    ~ 1 / (1 + exp(-(Intercept + EHG_scaled + LVN_scaled + CHG_scaled + spt)))
)
pred_df <- cbind(as.data.frame(pred_ance), st_coordinates(pred_ance))
g <- pblapply(c("mean", "sd"), function(p){
  pl <- ggplot() +
    geom_tile(data = pred_df, aes_string("X", "Y", fill = p)) +
    geom_sf(data = cover, inherit.aes = F, fill = 'transparent', alpha = 0.1) +
    scale_fill_gradientn(colors = rev(pals::brewer.rdylbu(100)),
                         breaks = seq(0, 1, 0.2),
                         limits = c(0,1)) +
    facet_wrap(~-YearBin) +
    ggtitle(label = "Left burial side", subtitle = p) +
    theme_bw() +
    theme(axis.text = element_blank(),
          axis.title = element_blank(),
          axis.ticks = element_blank(),
          legend.title = element_blank())
  return(pl)
})
g[[1]]
```

### Extract WAIC/DIC scores
```{r left waic, warning=F, message=F}
mlist <- list(left.m, left.m1, left.m2, left.EHG, left.WHG, left.CHG, left.LVN, left.m3, left.m4)
waic <- sapply(mlist, function(x) x$waic$waic)
dic <- sapply(mlist, function(x) x$dic$dic)

msel <- data.table(
  Model = factor(c("Spatiotemporal", "All", "Ancestry_only", "EHG", "WHG", "CHG", "LVN", "Mobility_only", "No_WHG"),
    levels = c("Spatiotemporal", "All", "Ancestry_only", "EHG", "WHG", "CHG", "LVN", "Mobility_only", "No_WHG")
  ),
  WAIC = waic, DIC = dic
)

ggplot(data = msel, aes(x = Model, y = WAIC, color = Model, group = 1)) +
  geom_point(size = 2) +
  geom_line() +
  scale_color_manual(values = pals::parula(9)) +
  theme_classic()

ggplot(data = msel, aes(x = Model, y = DIC, color = Model, group = 1)) +
  geom_point(size = 2) +
  geom_line() +
  scale_color_manual(values = pals::parula(9)) +
  theme_classic()
```

### Extract marginal log-likelihood scores
```{r mlik left, warning=F, message=F}
mlist <- list(left.m, left.m1, left.m2, left.EHG, left.WHG, left.CHG, left.LVN, left.m3, left.m4)
mlik <- sapply(mlist, function(x) x$mlik[2,1])

msel <- data.table(
  Model = factor(c("Spatiotemporal", "All", "Ancestry_only", "EHG", "WHG", "CHG", "LVN", "Mobility_only", "No_WHG"),
    levels = c("Spatiotemporal", "All", "Ancestry_only", "EHG", "WHG", "CHG", "LVN", "Mobility_only", "No_WHG")
  ),
  MLIK = mlik
)

ggplot(data = msel, aes(x = Model, y = MLIK, color = Model, group = 1)) +
  geom_point(size = 2) +
  geom_line() +
  scale_color_manual(values = pals::parula(9)) +
  theme_classic() +
  theme(axis.text = element_text(hjust = 1, angle = 45))

```

#### Ancestries without WHG: before "steppe"
```{r before steppe left, message = F, warning = F}
knitr::opts_chunk$set(cache = TRUE)
# Components
cmp <- BurialSide_left ~ Intercept(1) +
    EHG_scaled +
    LVN_scaled +
    CHG_scaled +
  spt(geometry,
      model = matern,
      group = YearBin,
      group_mapper = bru_mapper(mesh.t, indexed = TRUE),
      control.group = list(model = "ar1", hyper = h.spec))

# Model
left_m4.1 <- bru(cmp,
          data = left[which(left$YearBin > 5000),],
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(left_m4.1)
pl <- lapply(newnames[-c(1,3)], function(v){
    plot(left_m4.1, v) +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("Before steppe") +
      theme_bw()
})
multiplot(plotlist = pl)
```

#### Ancestries without WHG: after "steppe"
```{r after steppe left, message = F, warning = F}
knitr::opts_chunk$set(cache = TRUE)

# Create dataset
left_steppe <- left[which(left$YearBin <= 5000),]
left_steppe$Steppe_scaled <- left_steppe$EHG_scaled + left_steppe$CHG_scaled

# Components
cmp <- BurialSide_left ~ Intercept(1) +
    Steppe_scaled +
    LVN_scaled +
  spt(geometry,
      model = matern,
      group = YearBin,
      group_mapper = bru_mapper(mesh.t, indexed = TRUE),
      control.group = list(model = "ar1", hyper = h.spec))

# Model
left_m4.2 <- bru(cmp,
          data = left_steppe,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(left_m4.2)
pl <- lapply(c("Steppe_scaled", "LVN_scaled"), function(v){
    plot(left_m4.2, v) +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("After steppe") +
      theme_bw()
})
multiplot(plotlist = pl)
```

#### Ancestries without WHG: only "steppe" after 5k bp
```{r only steppe left, message = F, warning = F}
knitr::opts_chunk$set(cache = TRUE)
# Components
cmp <- BurialSide_left ~ Intercept(1) +
    Steppe_scaled +
  spt(geometry,
      model = matern,
      group = YearBin,
      group_mapper = bru_mapper(mesh.t, indexed = TRUE),
      control.group = list(model = "ar1", hyper = h.spec))

# Model
left_m4.3 <- bru(cmp,
          data = left_steppe,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(left_m4.3)
plot(left_m4.3, "Steppe_scaled") +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("Steppe only") +
      theme_bw()
```

### Extract WAIC/DIC scores for steppe models
```{r waic2 left, warning=F, message=F}
mlist <- list(left_m4.2, left_m4.3)
waic <- sapply(mlist, function(x) x$waic$waic)
dic <- sapply(mlist, function(x) x$dic$dic)

msel <- data.table(
  Model = factor(c("After_steppe", "Steppe_only"),
    levels = c("After_steppe", "Steppe_only")
  ),
  WAIC = waic, DIC = dic
)

ggplot(data = msel, aes(x = Model, y = WAIC, color = Model, group = 1)) +
  geom_point(size = 2) +
  geom_line() +
  scale_color_manual(values = pals::parula(2)) +
  theme_classic() +
  theme(axis.text = element_text(hjust = 1, angle = 45))

ggplot(data = msel, aes(x = Model, y = DIC, color = Model, group = 1)) +
  geom_point(size = 2) +
  geom_line() +
  scale_color_manual(values = pals::parula(12)) +
  theme_classic() +
  theme(axis.text = element_text(hjust = 1, angle = 45))
```

### Extract marginal log-likelihood scores for steppe models
```{r mlik2 left, warning=F, message=F}
mlist <- list(left_m4.2, left_m4.3)
mlik <- sapply(mlist, function(x) x$mlik[2,1])

msel <- data.table(
  Model = factor(c("After_steppe", "Steppe_only"),
    levels = c("After_steppe", "Steppe_only")
  ),
  MLIK = mlik
)

ggplot(data = msel, aes(x = Model, y = MLIK, color = Model, group = 1)) +
  geom_point(size = 2) +
  geom_line() +
  scale_color_manual(values = pals::parula(2)) +
  theme_classic() +
  theme(axis.text = element_text(hjust = 1, angle = 45))

```

# BURIAL SIDE: RIGHT
```{r right, warning=F, message=F}
# Data
right <- pnts[, c("YearBP", "BurialSide_right", "mobility", "EHG", "WHG", "LVN", "CHG", "geometry")]
right <- right[!is.na(right$BurialSide_right),]
nrow(right)

# Jitter the duplicates
duplicates <- right[duplicated(right),]
duplicates <- st_jitter(duplicates, 0.05)

plot(conts_prj$geometry, xlim = ext(right)[1:2], ylim = ext(right)[3:4]) 
points(right, pch = 1, cex = 1); points(duplicates, pch = 1, col = 'red', cex = 1)

# Add duplicates to the right data
right <- right[!duplicated(right),]
right <- rbind(right, duplicates)
nrow(right) # same as before

# Divide into time bins
right$YearBin <- plyr:::round_any(right$YearBP, 250)
```

### Null model: spatiotemporal process
```{r right m, warning=F, message=F}
knitr::opts_chunk$set(cache = TRUE)
# Components
cmp <- BurialSide_right ~ Intercept(1) +
  spatemp(geometry,
      model = matern,
      group = YearBin,
      group_mapper = bru_mapper(mesh.t, indexed = TRUE),
      control.group = list(model = "ar1", hyper = h.spec))

# Model
set.seed(1234)
right.m <- bru(cmp,
          data = right,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(right.m)

# Predict
pix <- fm_pixels(mesh, mask = cover)
pred.t <- seq(2000, 10000, 1000)
df <- fm_cprod(pix, data.frame(YearBin = pred.t))

pred <- predict(right.m,
                df,
                ~ 1 / (1 + exp(-(Intercept + spatemp))))

# Plot data
plot_df <- right
plot_df$YearBin <- plyr:::round_any(plot_df$YearBin, 1000)

ggplot(data = cbind(as.data.frame(plot_df), st_coordinates(plot_df)),
       aes(X, Y, shape = factor(BurialSide_right, levels = c(1,0)),
           size = factor(BurialSide_right, levels = c(1,0)),
           color = factor(BurialSide_right, levels = c(1,0)),
           alpha = factor(BurialSide_right, levels = c(1,0)))) +
  geom_sf(data = cover, alpha = 0.1, inherit.aes = F) +
  geom_sf(data = conts_prj, inherit.aes = F) +
  facet_wrap(~-YearBin) +
  geom_point() +
  scale_shape_manual(values = c(8, 19)) +
  scale_size_manual(values = c(2, 1)) +
  scale_alpha_manual(values = c(1, 0.5)) +
  coord_sf(xlim = st_bbox(cover)[c(1,3)], ylim = st_bbox(cover)[c(2,4)]) +
  theme_classic() +
  ggtitle("Right burial side") +
  theme(legend.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = 'bold'))

# Plot model prediction
pred_df <- cbind(as.data.frame(pred), st_coordinates(pred))
g <- pblapply(c("mean", "sd", "median"), function(p){
  pl <- ggplot() +
    geom_tile(data = pred_df, aes_string("X", "Y", fill = p)) +
    geom_sf(data = cover, inherit.aes = F, fill = 'transparent', alpha = 0.1) +
    
    scale_fill_gradientn(colors = rev(pals::brewer.rdylbu(100)),
                         breaks = seq(0, 1, 0.2),
                         limits = c(0,1)) +
    facet_wrap(~-YearBin) +
    ggtitle(label = "Right burial side", subtitle = p) +
    theme_bw() +
    theme(axis.text = element_blank(),
          axis.title = element_blank(),
          axis.ticks = element_blank(),
          legend.title = element_blank())
  return(pl)
})
g[[1]]
g[[2]]
g[[3]]
```

### All covariates
```{r right m1, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(cache = TRUE)

## Define covariates
namescov <- c("mobility", "EHG", "WHG", "LVN", "CHG")
newnames <- paste0(namescov, "_scaled")
covs <- data.frame(right)[, namescov]
covs <- scale(covs)
colnames(covs) <- newnames
right <- cbind(right, covs)

# Components
cmp <- BurialSide_right ~ Intercept(1) +
    mobility_scaled +
    EHG_scaled +
    WHG_scaled +
    LVN_scaled +
    CHG_scaled +
  spt(geometry,
      model = matern,
      group = YearBin,
      group_mapper = bru_mapper(mesh.t, indexed = TRUE),
      control.group = list(model = "ar1", hyper = h.spec))

# Model
set.seed(1234)
right.m1 <- bru(cmp,
          data = right,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(right.m1)
pl <- lapply(1:length(newnames), function(v){
    plot(right.m1, newnames[v]) +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      xlab(newnames[v]) +
      theme_bw() +
      theme(axis.title.y = element_blank())
})
multiplot(plotlist = pl)
```

### Ancestry only
```{r right m2, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(cache = TRUE)

# Components
cmp <- BurialSide_right ~ Intercept(1) +
    EHG_scaled +
    WHG_scaled +
    LVN_scaled +
    CHG_scaled +
  spt(geometry,
      model = matern,
      group = YearBin,
      group_mapper = bru_mapper(mesh.t, indexed = TRUE),
      control.group = list(model = "ar1", hyper = h.spec))

# Model
set.seed(1234)
right.m2 <- bru(cmp,
          data = right,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(right.m2)
pl <- lapply(2:length(newnames), function(v){
    plot(right.m2, newnames[v]) +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("Right burial side") +
      theme_bw()
})
multiplot(plotlist = pl)

# Extract ancestry values
ance <- readRDS("../../MesoNeo/SelectedAncestries_100ystep.RDS")
time <- raster::getZ(ance[[1]])
layers <- which(raster::getZ(ance[[1]]) %in% sort(pred.t*-1))
ance <- pblapply(ance, rast)
ance <- pblapply(ance, project, proj)
ance_pred <- lapply(ance, function(a){
    pblapply(layers, function(r){
        extract(a[[r]], st_coordinates(pix))
    })
})
ance_pred <- lapply(1:length(ance_pred), function(a){
    pblapply(1:length(ance_pred[[a]]), function(d){
        df <- cbind(st_coordinates(pix), ance_pred[[a]][[d]], data.frame(YearBin = sort(pred.t, decreasing = T)[d]))
        colnames(df)[3] <- names(ance_pred[a])
        return(as.data.table(df))
    })
})
ance_pred <- pblapply(ance_pred, rbindlist)

data_pred <- cbind(ance_pred[[1]][, c(1,2,4)], ance_pred[[1]][, 3], ance_pred[[2]][, 3], ance_pred[[3]][,3], ance_pred[[4]][,3])
data_pred <- st_as_sf(vect(data_pred, geom=c("X", "Y"), crs = proj))
colnames(data_pred)[2:5] <- newnames[-1]

# Predict
pred2 <- predict(
    right.m2,
    data_pred,
    ~ 1 / (1 + exp(-(Intercept + EHG_scaled + WHG_scaled + LVN_scaled + CHG_scaled + spt)))
)
pred_df <- cbind(as.data.frame(pred2), st_coordinates(pred2))
g <- pblapply(c("mean", "sd"), function(p){
  pl <- ggplot() +
    geom_tile(data = pred_df, aes_string("X", "Y", fill = p)) +
    geom_sf(data = cover, inherit.aes = F, fill = 'transparent', alpha = 0.1) +
    
    scale_fill_gradientn(colors = rev(pals::brewer.rdylbu(100)),
                         breaks = seq(0, 1, 0.2),
                         limits = c(0,1)) +
    facet_wrap(~-YearBin) +
    ggtitle(label = "Right burial side", subtitle = p) +
    theme_bw() +
    theme(axis.text = element_blank(),
          axis.title = element_blank(),
          axis.ticks = element_blank(),
          legend.title = element_blank())
  return(pl)
})
g[[1]]
```

### Single ancestry: EHG
```{r right EHG, message=F, warning = F}
knitr::opts_chunk$set(cache = TRUE)
# Components
cmp <- BurialSide_right ~ Intercept(1) +
    EHG_scaled +
    spt(geometry,
    model = matern,
    group = YearBin,
    group_mapper = bru_mapper(mesh.t, indexed = TRUE),
    control.group = list(model = "ar1", hyper = h.spec))

# Model
right.EHG <- bru(cmp,
          data = right,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(right.EHG)
plot(right.EHG, "EHG_scaled") +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("Right burial side") +
      theme_bw()
```

### Single ancestry: WHG
```{r right WHG, message=F, warning = F}
knitr::opts_chunk$set(cache = TRUE)
# Components
cmp <- BurialSide_right ~ Intercept(1) +
    WHG_scaled +
    spt(geometry,
    model = matern,
    group = YearBin,
    group_mapper = bru_mapper(mesh.t, indexed = TRUE),
    control.group = list(model = "ar1", hyper = h.spec))

# Model
right.WHG <- bru(cmp,
          data = right,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(right.WHG)
plot(right.WHG, "WHG_scaled") +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("Right burial side") +
      theme_bw()
```

### Single ancestry: CHG
```{r right CHG, message=F, warning = F}
knitr::opts_chunk$set(cache = TRUE)
# Components
cmp <- BurialSide_right ~ Intercept(1) +
    CHG_scaled +
    spt(geometry,
    model = matern,
    group = YearBin,
    group_mapper = bru_mapper(mesh.t, indexed = TRUE),
    control.group = list(model = "ar1", hyper = h.spec))

# Model
right.CHG <- bru(cmp,
          data = right,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(right.CHG)
plot(right.CHG, "CHG_scaled") +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("Right burial side") +
      theme_bw()
```

### Single ancestry: LVN
```{r right LVN, message=F, warning = F}
knitr::opts_chunk$set(cache = TRUE)
# Components
cmp <- BurialSide_right ~ Intercept(1) +
    LVN_scaled +
    spt(geometry,
    model = matern,
    group = YearBin,
    group_mapper = bru_mapper(mesh.t, indexed = TRUE),
    control.group = list(model = "ar1", hyper = h.spec))

# Model
right.LVN <- bru(cmp,
          data = right,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(right.LVN)
plot(right.LVN, "LVN_scaled") +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("Right burial side") +
      theme_bw()
```

### Only mobility
```{r right mobility, warning=F, message=F}
knitr::opts_chunk$set(cache = TRUE)
# Components
cmp <- BurialSide_right ~ Intercept(1) +
  mobility_scaled +
  spt(geometry,
    model = matern,
    group = YearBin,
    group_mapper = bru_mapper(mesh.t, indexed = TRUE),
    control.group = list(model = "ar1", hyper = h.spec)
  )

# Model
right.m3 <- bru(cmp,
          data = right,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(right.m3)
plot(right.m3, "mobility_scaled") +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("Right burial side") +
      theme_bw()
```

### Only ancestry: no WHG
```{r right no WHG, message = F, warning = F}
knitr::opts_chunk$set(cache = TRUE)
# Components
cmp <- BurialSide_right ~ Intercept(1) +
    EHG_scaled +
    LVN_scaled +
    CHG_scaled +
  spt(geometry,
      model = matern,
      group = YearBin,
      group_mapper = bru_mapper(mesh.t, indexed = TRUE),
      control.group = list(model = "ar1", hyper = h.spec))

# Model
right.m4 <- bru(cmp,
          data = right,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(right.m4)
pl <- lapply(newnames[-c(1,3)], function(v){
    plot(right.m4, v) +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("Right burial side") +
      theme_bw()
})
multiplot(plotlist = pl)

# Predict
pred_ance <- predict(
    right.m4,
    data_pred,
    ~ 1 / (1 + exp(-(Intercept + EHG_scaled + LVN_scaled + CHG_scaled + spt)))
)
pred_df <- cbind(as.data.frame(pred_ance), st_coordinates(pred_ance))
g <- pblapply(c("mean", "sd"), function(p){
  pl <- ggplot() +
    geom_tile(data = pred_df, aes_string("X", "Y", fill = p)) +
    geom_sf(data = cover, inherit.aes = F, fill = 'transparent', alpha = 0.1) +
    scale_fill_gradientn(colors = rev(pals::brewer.rdylbu(100)),
                         breaks = seq(0, 1, 0.2),
                         limits = c(0,1)) +
    facet_wrap(~-YearBin) +
    ggtitle(label = "Right burial side", subtitle = p) +
    theme_bw() +
    theme(axis.text = element_blank(),
          axis.title = element_blank(),
          axis.ticks = element_blank(),
          legend.title = element_blank())
  return(pl)
})
g[[1]]
```

### Extract WAIC/DIC scores
```{r right waic, warning=F, message=F}
mlist <- list(right.m, right.m1, right.m2, right.EHG, right.WHG, right.CHG, right.LVN, right.m3, right.m4)
waic <- sapply(mlist, function(x) x$waic$waic)
dic <- sapply(mlist, function(x) x$dic$dic)

msel <- data.table(
  Model = factor(c("Spatiotemporal", "All", "Ancestry_only", "EHG", "WHG", "CHG", "LVN", "Mobility_only", "No_WHG"),
    levels = c("Spatiotemporal", "All", "Ancestry_only", "EHG", "WHG", "CHG", "LVN", "Mobility_only", "No_WHG")
  ),
  WAIC = waic, DIC = dic
)

ggplot(data = msel, aes(x = Model, y = WAIC, color = Model, group = 1)) +
  geom_point(size = 2) +
  geom_line() +
  scale_color_manual(values = pals::parula(9)) +
  theme_classic()

ggplot(data = msel, aes(x = Model, y = DIC, color = Model, group = 1)) +
  geom_point(size = 2) +
  geom_line() +
  scale_color_manual(values = pals::parula(9)) +
  theme_classic()
```

### Extract marginal log-likelihood scores
```{r mlik right, warning=F, message=F}
mlist <- list(right.m, right.m1, right.m2, right.EHG, right.WHG, right.CHG, right.LVN, right.m3, right.m4)
mlik <- sapply(mlist, function(x) x$mlik[2,1])

msel <- data.table(
  Model = factor(c("Spatiotemporal", "All", "Ancestry_only", "EHG", "WHG", "CHG", "LVN", "Mobility_only", "No_WHG"),
    levels = c("Spatiotemporal", "All", "Ancestry_only", "EHG", "WHG", "CHG", "LVN", "Mobility_only", "No_WHG")
  ),
  MLIK = mlik
)

ggplot(data = msel, aes(x = Model, y = MLIK, color = Model, group = 1)) +
  geom_point(size = 2) +
  geom_line() +
  scale_color_manual(values = pals::parula(9)) +
  theme_classic() +
  theme(axis.text = element_text(hjust = 1, angle = 45))

```

#### Ancestries without WHG: before "steppe"
```{r before steppe right, message = F, warning = F}
knitr::opts_chunk$set(cache = TRUE)
# Components
cmp <- BurialSide_right ~ Intercept(1) +
    EHG_scaled +
    LVN_scaled +
    CHG_scaled +
  spt(geometry,
      model = matern,
      group = YearBin,
      group_mapper = bru_mapper(mesh.t, indexed = TRUE),
      control.group = list(model = "ar1", hyper = h.spec))

# Model
right_m4.1 <- bru(cmp,
          data = right[which(right$YearBin > 5000),],
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(right_m4.1)
pl <- lapply(newnames[-c(1,3)], function(v){
    plot(right_m4.1, v) +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("Before steppe") +
      theme_bw()
})
multiplot(plotlist = pl)
```

#### Ancestries without WHG: after "steppe"
```{r after steppe right, message = F, warning = F}
knitr::opts_chunk$set(cache = TRUE)

# Create dataset
right_steppe <- right[which(right$YearBin <= 5000),]
right_steppe$Steppe_scaled <- right_steppe$EHG_scaled + right_steppe$CHG_scaled

# Components
cmp <- BurialSide_right ~ Intercept(1) +
    Steppe_scaled +
    LVN_scaled +
  spt(geometry,
      model = matern,
      group = YearBin,
      group_mapper = bru_mapper(mesh.t, indexed = TRUE),
      control.group = list(model = "ar1", hyper = h.spec))

# Model
right_m4.2 <- bru(cmp,
          data = right_steppe,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(right_m4.2)
pl <- lapply(c("Steppe_scaled", "LVN_scaled"), function(v){
    plot(right_m4.2, v) +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("After steppe") +
      theme_bw()
})
multiplot(plotlist = pl)
```

#### Ancestries without WHG: only "steppe" after 5k bp
```{r only steppe right, message = F, warning = F}
knitr::opts_chunk$set(cache = TRUE)
# Components
cmp <- BurialSide_right ~ Intercept(1) +
    Steppe_scaled +
  spt(geometry,
      model = matern,
      group = YearBin,
      group_mapper = bru_mapper(mesh.t, indexed = TRUE),
      control.group = list(model = "ar1", hyper = h.spec))

# Model
right_m4.3 <- bru(cmp,
          data = right_steppe,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(right_m4.3)
plot(right_m4.3, "Steppe_scaled") +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("Steppe only") +
      theme_bw()
```

### Extract WAIC/DIC scores for steppe models
```{r waic2 right, warning=F, message=F}
mlist <- list(right_m4.2, right_m4.3)
waic <- sapply(mlist, function(x) x$waic$waic)
dic <- sapply(mlist, function(x) x$dic$dic)

msel <- data.table(
  Model = factor(c("After_steppe", "Steppe_only"),
    levels = c("After_steppe", "Steppe_only")
  ),
  WAIC = waic, DIC = dic
)

ggplot(data = msel, aes(x = Model, y = WAIC, color = Model, group = 1)) +
  geom_point(size = 2) +
  geom_line() +
  scale_color_manual(values = pals::parula(2)) +
  theme_classic() +
  theme(axis.text = element_text(hjust = 1, angle = 45))

ggplot(data = msel, aes(x = Model, y = DIC, color = Model, group = 1)) +
  geom_point(size = 2) +
  geom_line() +
  scale_color_manual(values = pals::parula(12)) +
  theme_classic() +
  theme(axis.text = element_text(hjust = 1, angle = 45))
```

### Extract marginal log-likelihood scores for steppe models
```{r mlik2 right, warning=F, message=F}
mlist <- list(right_m4.2, right_m4.3)
mlik <- sapply(mlist, function(x) x$mlik[2,1])

msel <- data.table(
  Model = factor(c("After_steppe", "Steppe_only"),
    levels = c("After_steppe", "Steppe_only")
  ),
  MLIK = mlik
)

ggplot(data = msel, aes(x = Model, y = MLIK, color = Model, group = 1)) +
  geom_point(size = 2) +
  geom_line() +
  scale_color_manual(values = pals::parula(2)) +
  theme_classic() +
  theme(axis.text = element_text(hjust = 1, angle = 45))

```

# BURIAL SIDE: BACK
```{r back, warning=F, message=F}
# Data
back <- pnts[, c("YearBP", "BurialSide_back", "mobility", "EHG", "WHG", "LVN", "CHG", "geometry")]
back <- back[!is.na(back$BurialSide_back),]
nrow(back)

# Jitter the duplicates
duplicates <- back[duplicated(back),]
duplicates <- st_jitter(duplicates, 0.05)

plot(conts_prj$geometry, xlim = ext(back)[1:2], ylim = ext(back)[3:4]) 
points(back, pch = 1, cex = 1); points(duplicates, pch = 1, col = 'red', cex = 1)

# Add duplicates to the back data
back <- back[!duplicated(back),]
back <- rbind(back, duplicates)
nrow(back) # same as before

# Divide into time bins
back$YearBin <- plyr:::round_any(back$YearBP, 250)
```

### Null model: spatiotemporal process
```{r back m, warning=F, message=F}
knitr::opts_chunk$set(cache = TRUE)
# Components
cmp <- BurialSide_back ~ Intercept(1) +
  spatemp(geometry,
      model = matern,
      group = YearBin,
      group_mapper = bru_mapper(mesh.t, indexed = TRUE),
      control.group = list(model = "ar1", hyper = h.spec))

# Model
set.seed(1234)
back.m <- bru(cmp,
          data = back,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(back.m)

# Predict
pix <- fm_pixels(mesh, mask = cover)
pred.t <- seq(2000, 10000, 1000)
df <- fm_cprod(pix, data.frame(YearBin = pred.t))

pred <- predict(back.m,
                df,
                ~ 1 / (1 + exp(-(Intercept + spatemp))))

# Plot data
plot_df <- back
plot_df$YearBin <- plyr:::round_any(plot_df$YearBin, 1000)

ggplot(data = cbind(as.data.frame(plot_df), st_coordinates(plot_df)),
       aes(X, Y, shape = factor(BurialSide_back, levels = c(1,0)),
           size = factor(BurialSide_back, levels = c(1,0)),
           color = factor(BurialSide_back, levels = c(1,0)),
           alpha = factor(BurialSide_back, levels = c(1,0)))) +
  geom_sf(data = cover, alpha = 0.1, inherit.aes = F) +
  geom_sf(data = conts_prj, inherit.aes = F) +
  facet_wrap(~-YearBin) +
  geom_point() +
  scale_shape_manual(values = c(8, 19)) +
  scale_size_manual(values = c(2, 1)) +
  scale_alpha_manual(values = c(1, 0.5)) +
  coord_sf(xlim = st_bbox(cover)[c(1,3)], ylim = st_bbox(cover)[c(2,4)]) +
  theme_classic() +
  ggtitle("Back burial side") +
  theme(legend.title = element_blank(),
        plot.title = element_text(hjust = 0.5, face = 'bold'))

# Plot model prediction
pred_df <- cbind(as.data.frame(pred), st_coordinates(pred))
g <- pblapply(c("mean", "sd", "median"), function(p){
  pl <- ggplot() +
    geom_tile(data = pred_df, aes_string("X", "Y", fill = p)) +
    geom_sf(data = cover, inherit.aes = F, fill = 'transparent', alpha = 0.1) +
    
    scale_fill_gradientn(colors = rev(pals::brewer.rdylbu(100)),
                         breaks = seq(0, 1, 0.2),
                         limits = c(0,1)) +
    facet_wrap(~-YearBin) +
    ggtitle(label = "Back burial side", subtitle = p) +
    theme_bw() +
    theme(axis.text = element_blank(),
          axis.title = element_blank(),
          axis.ticks = element_blank(),
          legend.title = element_blank())
  return(pl)
})
g[[1]]
g[[2]]
g[[3]]
```

### All covariates
```{r back m1, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(cache = TRUE)

## Define covariates
namescov <- c("mobility", "EHG", "WHG", "LVN", "CHG")
newnames <- paste0(namescov, "_scaled")
covs <- data.frame(back)[, namescov]
covs <- scale(covs)
colnames(covs) <- newnames
back <- cbind(back, covs)

# Components
cmp <- BurialSide_back ~ Intercept(1) +
    mobility_scaled +
    EHG_scaled +
    WHG_scaled +
    LVN_scaled +
    CHG_scaled +
  spt(geometry,
      model = matern,
      group = YearBin,
      group_mapper = bru_mapper(mesh.t, indexed = TRUE),
      control.group = list(model = "ar1", hyper = h.spec))

# Model
set.seed(1234)
back.m1 <- bru(cmp,
          data = back,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(back.m1)
pl <- lapply(1:length(newnames), function(v){
    plot(back.m1, newnames[v]) +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      xlab(newnames[v]) +
      theme_bw() +
      theme(axis.title.y = element_blank())
})
multiplot(plotlist = pl)
```

### Ancestry only
```{r back m2, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(cache = TRUE)

# Components
cmp <- BurialSide_back ~ Intercept(1) +
    EHG_scaled +
    WHG_scaled +
    LVN_scaled +
    CHG_scaled +
  spt(geometry,
      model = matern,
      group = YearBin,
      group_mapper = bru_mapper(mesh.t, indexed = TRUE),
      control.group = list(model = "ar1", hyper = h.spec))

# Model
set.seed(1234)
back.m2 <- bru(cmp,
          data = back,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(back.m2)
pl <- lapply(2:length(newnames), function(v){
    plot(back.m2, newnames[v]) +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("Back burial side") +
      theme_bw()
})
multiplot(plotlist = pl)

# Extract ancestry values
ance <- readRDS("../../MesoNeo/SelectedAncestries_100ystep.RDS")
time <- raster::getZ(ance[[1]])
layers <- which(raster::getZ(ance[[1]]) %in% sort(pred.t*-1))
ance <- pblapply(ance, rast)
ance <- pblapply(ance, project, proj)
ance_pred <- lapply(ance, function(a){
    pblapply(layers, function(r){
        extract(a[[r]], st_coordinates(pix))
    })
})
ance_pred <- lapply(1:length(ance_pred), function(a){
    pblapply(1:length(ance_pred[[a]]), function(d){
        df <- cbind(st_coordinates(pix), ance_pred[[a]][[d]], data.frame(YearBin = sort(pred.t, decreasing = T)[d]))
        colnames(df)[3] <- names(ance_pred[a])
        return(as.data.table(df))
    })
})
ance_pred <- pblapply(ance_pred, rbindlist)

data_pred <- cbind(ance_pred[[1]][, c(1,2,4)], ance_pred[[1]][, 3], ance_pred[[2]][, 3], ance_pred[[3]][,3], ance_pred[[4]][,3])
data_pred <- st_as_sf(vect(data_pred, geom=c("X", "Y"), crs = proj))
colnames(data_pred)[2:5] <- newnames[-1]

# Predict
pred2 <- predict(
    back.m2,
    data_pred,
    ~ 1 / (1 + exp(-(Intercept + EHG_scaled + WHG_scaled + LVN_scaled + CHG_scaled + spt)))
)
pred_df <- cbind(as.data.frame(pred2), st_coordinates(pred2))
g <- pblapply(c("mean", "sd"), function(p){
  pl <- ggplot() +
    geom_tile(data = pred_df, aes_string("X", "Y", fill = p)) +
    geom_sf(data = cover, inherit.aes = F, fill = 'transparent', alpha = 0.1) +
    
    scale_fill_gradientn(colors = rev(pals::brewer.rdylbu(100)),
                         breaks = seq(0, 1, 0.2),
                         limits = c(0,1)) +
    facet_wrap(~-YearBin) +
    ggtitle(label = "Back burial side", subtitle = p) +
    theme_bw() +
    theme(axis.text = element_blank(),
          axis.title = element_blank(),
          axis.ticks = element_blank(),
          legend.title = element_blank())
  return(pl)
})
g[[1]]
```

### Single ancestry: EHG
```{r back EHG, message=F, warning = F}
knitr::opts_chunk$set(cache = TRUE)
# Components
cmp <- BurialSide_back ~ Intercept(1) +
    EHG_scaled +
    spt(geometry,
    model = matern,
    group = YearBin,
    group_mapper = bru_mapper(mesh.t, indexed = TRUE),
    control.group = list(model = "ar1", hyper = h.spec))

# Model
back.EHG <- bru(cmp,
          data = back,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(back.EHG)
plot(back.EHG, "EHG_scaled") +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("Back burial side") +
      theme_bw()
```

### Single ancestry: WHG
```{r back WHG, message=F, warning = F}
knitr::opts_chunk$set(cache = TRUE)
# Components
cmp <- BurialSide_back ~ Intercept(1) +
    WHG_scaled +
    spt(geometry,
    model = matern,
    group = YearBin,
    group_mapper = bru_mapper(mesh.t, indexed = TRUE),
    control.group = list(model = "ar1", hyper = h.spec))

# Model
back.WHG <- bru(cmp,
          data = back,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(back.WHG)
plot(back.WHG, "WHG_scaled") +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("Back burial side") +
      theme_bw()
```

### Single ancestry: CHG
```{r back CHG, message=F, warning = F}
knitr::opts_chunk$set(cache = TRUE)
# Components
cmp <- BurialSide_back ~ Intercept(1) +
    CHG_scaled +
    spt(geometry,
    model = matern,
    group = YearBin,
    group_mapper = bru_mapper(mesh.t, indexed = TRUE),
    control.group = list(model = "ar1", hyper = h.spec))

# Model
back.CHG <- bru(cmp,
          data = back,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(back.CHG)
plot(back.CHG, "CHG_scaled") +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("Back burial side") +
      theme_bw()
```

### Single ancestry: LVN
```{r back LVN, message=F, warning = F}
knitr::opts_chunk$set(cache = TRUE)
# Components
cmp <- BurialSide_back ~ Intercept(1) +
    LVN_scaled +
    spt(geometry,
    model = matern,
    group = YearBin,
    group_mapper = bru_mapper(mesh.t, indexed = TRUE),
    control.group = list(model = "ar1", hyper = h.spec))

# Model
back.LVN <- bru(cmp,
          data = back,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(back.LVN)
plot(back.LVN, "LVN_scaled") +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("Back burial side") +
      theme_bw()
```

### Only mobility
```{r back mobility, warning=F, message=F}
knitr::opts_chunk$set(cache = TRUE)
# Components
cmp <- BurialSide_back ~ Intercept(1) +
  mobility_scaled +
  spt(geometry,
    model = matern,
    group = YearBin,
    group_mapper = bru_mapper(mesh.t, indexed = TRUE),
    control.group = list(model = "ar1", hyper = h.spec)
  )

# Model
back.m3 <- bru(cmp,
          data = back,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(back.m3)
plot(back.m3, "mobility_scaled") +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("Back burial side") +
      theme_bw()
```

### Only ancestry: no WHG
```{r back no WHG, message = F, warning = F}
knitr::opts_chunk$set(cache = TRUE)
# Components
cmp <- BurialSide_back ~ Intercept(1) +
    EHG_scaled +
    LVN_scaled +
    CHG_scaled +
  spt(geometry,
      model = matern,
      group = YearBin,
      group_mapper = bru_mapper(mesh.t, indexed = TRUE),
      control.group = list(model = "ar1", hyper = h.spec))

# Model
back.m4 <- bru(cmp,
          data = back,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(back.m4)
pl <- lapply(newnames[-c(1,3)], function(v){
    plot(back.m4, v) +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("Back burial side") +
      theme_bw()
})
multiplot(plotlist = pl)

# Predict
pred_ance <- predict(
    back.m4,
    data_pred,
    ~ 1 / (1 + exp(-(Intercept + EHG_scaled + LVN_scaled + CHG_scaled + spt)))
)
pred_df <- cbind(as.data.frame(pred_ance), st_coordinates(pred_ance))
g <- pblapply(c("mean", "sd"), function(p){
  pl <- ggplot() +
    geom_tile(data = pred_df, aes_string("X", "Y", fill = p)) +
    geom_sf(data = cover, inherit.aes = F, fill = 'transparent', alpha = 0.1) +
    scale_fill_gradientn(colors = rev(pals::brewer.rdylbu(100)),
                         breaks = seq(0, 1, 0.2),
                         limits = c(0,1)) +
    facet_wrap(~-YearBin) +
    ggtitle(label = "Back burial side", subtitle = p) +
    theme_bw() +
    theme(axis.text = element_blank(),
          axis.title = element_blank(),
          axis.ticks = element_blank(),
          legend.title = element_blank())
  return(pl)
})
g[[1]]
```

### Extract WAIC/DIC scores
```{r back waic, warning=F, message=F}
mlist <- list(back.m, back.m1, back.m2, back.EHG, back.WHG, back.CHG, back.LVN, back.m3, back.m4)
waic <- sapply(mlist, function(x) x$waic$waic)
dic <- sapply(mlist, function(x) x$dic$dic)

msel <- data.table(
  Model = factor(c("Spatiotemporal", "All", "Ancestry_only", "EHG", "WHG", "CHG", "LVN", "Mobility_only", "No_WHG"),
    levels = c("Spatiotemporal", "All", "Ancestry_only", "EHG", "WHG", "CHG", "LVN", "Mobility_only", "No_WHG")
  ),
  WAIC = waic, DIC = dic
)

ggplot(data = msel, aes(x = Model, y = WAIC, color = Model, group = 1)) +
  geom_point(size = 2) +
  geom_line() +
  scale_color_manual(values = pals::parula(9)) +
  theme_classic()

ggplot(data = msel, aes(x = Model, y = DIC, color = Model, group = 1)) +
  geom_point(size = 2) +
  geom_line() +
  scale_color_manual(values = pals::parula(9)) +
  theme_classic()
```

### Extract marginal log-likelihood scores
```{r mlik back, warning=F, message=F}
mlist <- list(back.m, back.m1, back.m2, back.EHG, back.WHG, back.CHG, back.LVN, back.m3, back.m4)
mlik <- sapply(mlist, function(x) x$mlik[2,1])

msel <- data.table(
  Model = factor(c("Spatiotemporal", "All", "Ancestry_only", "EHG", "WHG", "CHG", "LVN", "Mobility_only", "No_WHG"),
    levels = c("Spatiotemporal", "All", "Ancestry_only", "EHG", "WHG", "CHG", "LVN", "Mobility_only", "No_WHG")
  ),
  MLIK = mlik
)

ggplot(data = msel, aes(x = Model, y = MLIK, color = Model, group = 1)) +
  geom_point(size = 2) +
  geom_line() +
  scale_color_manual(values = pals::parula(9)) +
  theme_classic() +
  theme(axis.text = element_text(hjust = 1, angle = 45))

```

#### Ancestries without WHG: before "steppe"
```{r before steppe back, message = F, warning = F}
knitr::opts_chunk$set(cache = TRUE)
# Components
cmp <- BurialSide_back ~ Intercept(1) +
    EHG_scaled +
    LVN_scaled +
    CHG_scaled +
  spt(geometry,
      model = matern,
      group = YearBin,
      group_mapper = bru_mapper(mesh.t, indexed = TRUE),
      control.group = list(model = "ar1", hyper = h.spec))

# Model
back_m4.1 <- bru(cmp,
          data = back[which(back$YearBin > 5000),],
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(back_m4.1)
pl <- lapply(newnames[-c(1,3)], function(v){
    plot(back_m4.1, v) +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("Before steppe") +
      theme_bw()
})
multiplot(plotlist = pl)
```

#### Ancestries without WHG: after "steppe"
```{r after steppe back, message = F, warning = F}
knitr::opts_chunk$set(cache = TRUE)

# Create dataset
back_steppe <- back[which(back$YearBin <= 5000),]
back_steppe$Steppe_scaled <- back_steppe$EHG_scaled + back_steppe$CHG_scaled

# Components
cmp <- BurialSide_back ~ Intercept(1) +
    Steppe_scaled +
    LVN_scaled +
  spt(geometry,
      model = matern,
      group = YearBin,
      group_mapper = bru_mapper(mesh.t, indexed = TRUE),
      control.group = list(model = "ar1", hyper = h.spec))

# Model
back_m4.2 <- bru(cmp,
          data = back_steppe,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(back_m4.2)
pl <- lapply(c("Steppe_scaled", "LVN_scaled"), function(v){
    plot(back_m4.2, v) +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("After steppe") +
      theme_bw()
})
multiplot(plotlist = pl)
```

#### Ancestries without WHG: only "steppe" after 5k bp
```{r only steppe back, message = F, warning = F}
knitr::opts_chunk$set(cache = TRUE)
# Components
cmp <- BurialSide_back ~ Intercept(1) +
    Steppe_scaled +
  spt(geometry,
      model = matern,
      group = YearBin,
      group_mapper = bru_mapper(mesh.t, indexed = TRUE),
      control.group = list(model = "ar1", hyper = h.spec))

# Model
back_m4.3 <- bru(cmp,
          data = back_steppe,
          domain = list(geometry = mesh, YearBin = mesh.t),
          family = "binomial",
          options = list(safe = T)
)
summary(back_m4.3)
plot(back_m4.3, "Steppe_scaled") +
      geom_vline(aes(xintercept = 0), linetype = 2, color = 'red4') +
      ggtitle("Steppe only") +
      theme_bw()
```

### Extract WAIC/DIC scores for steppe models
```{r waic2 back, warning=F, message=F}
mlist <- list(back_m4.2, back_m4.3)
waic <- sapply(mlist, function(x) x$waic$waic)
dic <- sapply(mlist, function(x) x$dic$dic)

msel <- data.table(
  Model = factor(c("After_steppe", "Steppe_only"),
    levels = c("After_steppe", "Steppe_only")
  ),
  WAIC = waic, DIC = dic
)

ggplot(data = msel, aes(x = Model, y = WAIC, color = Model, group = 1)) +
  geom_point(size = 2) +
  geom_line() +
  scale_color_manual(values = pals::parula(2)) +
  theme_classic() +
  theme(axis.text = element_text(hjust = 1, angle = 45))

ggplot(data = msel, aes(x = Model, y = DIC, color = Model, group = 1)) +
  geom_point(size = 2) +
  geom_line() +
  scale_color_manual(values = pals::parula(12)) +
  theme_classic() +
  theme(axis.text = element_text(hjust = 1, angle = 45))
```

### Extract marginal log-likelihood scores for steppe models
```{r mlik2 back, warning=F, message=F}
mlist <- list(back_m4.2, back_m4.3)
mlik <- sapply(mlist, function(x) x$mlik[2,1])

msel <- data.table(
  Model = factor(c("After_steppe", "Steppe_only"),
    levels = c("After_steppe", "Steppe_only")
  ),
  MLIK = mlik
)

ggplot(data = msel, aes(x = Model, y = MLIK, color = Model, group = 1)) +
  geom_point(size = 2) +
  geom_line() +
  scale_color_manual(values = pals::parula(2)) +
  theme_classic() +
  theme(axis.text = element_text(hjust = 1, angle = 45))

```